### 2022-08-20 TeX/baseline/paper.tex
Began drafting baseline report

\documentclass[10pt,journal,compsoc]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{Conference Paper Title\\
    % {\footnotesize \textsuperscript{*}Note: Sub-titles are not captured in Xplore and
    % should not be used}
    % \thanks{Identify applicable funding agency here. If none, delete this.}
}
\author{}
% \author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
% \textit{name of organization (of Aff.)}\\
% City, Country \\
% email address or ORCID}
% \and
% \IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
% \textit{name of organization (of Aff.)}\\
% City, Country \\
% email address or ORCID}

\maketitle

\begin{abstract}
    This document is a model and instructions for \LaTeX.
    This and the IEEEtran.cls file define the components of your paper [title, text, heads, etc.]. *CRITICAL: Do Not Use Symbols, Special Characters, Footnotes,
    or Math in Paper Title or Abstract.
\end{abstract}

\section{Introduction}
A common practice to gauge patients' physiological condition and acute clinical stability is using basic homeostatic measures like vital signs, haematologic findings, and neurological examinations. Various early warning systems have been proposed that use these measurements to assess whether patients face an imminent risk of deterioration or death [Smith13].

The National Early Warning Score (NEWS), developed in conjunction with the Royal College of Physicians, has been widely implemented across the NHS and by healthcare providers outside the UK [RCP17, pp.13]. It is a points-based early warning system to identify clinical deterioration in acutely ill patients based on routinely recorded vital signs: pulse, breathing rate, blood pressure, blood oxygen level, level of consciousness, and temperature. The paper-based NEWS allocates points in a weighted manner based on these measurements. The sum of these points, which makes up the patient's aggregate NEW score, is associated with "triggers" - thresholds that correspond to recommended levels of clinical response and monitoring frequency.

While initially designed to monitor for deterioration in secondary-care, the NEWS has been widely tested and validated across healthcare settings. In its latest revision of the standard, the RCP recommended that the NEWS also be used in emergency departments to "aid the initial assessment of patiens, ongoing monitoring and patient triage decisions" [RCP17, pp.18].

Across studies, the NEWS has been found to perform well in discriminating ward patients at risk of death, critical care admission, or cardiac arrest. The present report introduces the novel SCI dataset, which contains records of adult patients with acute or medical presentation at the Salford Royal NHS Foundation Trust from 2014-2022. We then use this dataset to assess the performance of NEWS in predicting adverse patient outcomes, such as those mentioned above, in the dataset population.

\section{Methodology}
\subsection{Data Collection and Contents}
The SCI dataset is collated from EPR records collected in real time from admissions to acute beds between 1st April 2014 and 31st March 2022. Further, the dataset includes administrative data produced in retrospect for financial (tarrif) purposes.

Our dataset includes records of patients who received ambulatory emergency care (AEC) or same-day emergency care (SDEC) and were discharged before midnight on the day of admission [NHS18]. AEC, alongside the medical admissions unit (EAU)  are the two common entry points for adult patients with general medical emergencies. Exceptions present in our records include certain booked admissions and patients directly transferred and admitted to medical or critical care wards.

As a routine part of admission, the responsible staff member (nurse or support worker) records the patient's vital signs withing $\sim 30$ minutes of arrival. The vital signs that make up the NEWS are measured simultaneously, in a standardised manner, using Dinamap monitors. The readings are manually transcribed into EPR and onto the patient's paper chart. Specifically, this data includes:
\begin{itemize}
    \item Body temperature ($^{\circ}$C)
    \item Pulse (beats/min)
    \item Diastolic and systolic blood pressure (mmHg)
    \item Peripheral oxygen saturation (\%)
\end{itemize}
Further, we record the patient's level of consciousness (AVPU), presence of pain, nausea, or vomiting, whether the patient was receiving oxygen at the time of SpO2 measurement and, if applicable, the oxygen flow-rate and mode of delivery. Once these parameters are inputted into EPR, the NEWS score and component sub-scores are automatically computed for the patient. Blood test results (including VBG), where those were peformed, are automatically recorded in the laboratory information management system LIMS) and subsequently imported into EPR.

Other patient information recorded on admission includes: Patient identifier data such as their unique patient number, age, sex, and address. Admission method (e.g., emergency A\&E, emergency GP referral, etc), their presenting complaint, arrival time, and the main A\&E diagnosis (following initial assessment).

Information gathered in retrospect includes the date and time of discharge, total length of stay (LOS), the wards the patient was admitted to (in chronological order) and the LOS per ward. Diagnoses coded using the ICD-10-CM standard are complied by a clinical coding team based on the existing codes and clinical notes recorded in EPR. Procedures and services are also recorded in detail in EPR, and are similarly coded in retrospect using the OPCS-4 standard.

\subsection{Outcomes} The outcomes we track are death during hospitalisation, death within 30 days after discharge, unanticipated admission to critical care, or unanticipated readmission to the same hospital. To assess the NEWS, we check for the occurrence of death or critical care admission within 24 hours, 48 hours, or at any point after the patient's arrival. For readmissions, we consider thresholds of 48 hours, 7 days, or 30 days after discharge.

Where it occurrs, patient death during their hospitalisation is identified from EPR. Mortality after discharge (within 30 days) is primarily determined from the regional primary care databases (Salford Integrated Records \& Greater Manchester Care Record). Critical care admission is identified from patients being admitted into the hospital's critical care unit (CCU) or the high-dependency medical unit (HH1M).

\subsection{Data Preparation} We perform all data manipulation in Python using the Numpy and Pandas libraries. As they are the only parameters that are transcribed manually into EPR, we check the patients' vital signs for spurious values against hard ranges (e.g., $0-100\%$ for SpO2) as well as soft thresholds that we base on the range of physiologically possible values as determined by professional clinical opinion. For parameters that have a recorded NEWS sub-score but their raw value is missing or marked as spurious, we infer the correct value as the midpoint of the relevant NEWS range. Where a parameter has a valid value but the NEWS sub-score is missing, we compute it in accordance with the specificaton [RCP17, pp.]. We use the same specification to compute the final NEWS score for any patients where this is not recorded in the original data. Full details of how each parameter is pre-processed are available in appendix A.

\subsection{Data Analysis} We assess the ability of NEWS to discriminate between patients who did and did not suffer an adverse outcome using AUROC (area under the ROC curve). We compute this using the scikit-Learn library. The minimum possible AUROC value is $0.5$ and corresponds to a completely random relationship between the score in question and the tracked outcomes. Generally, a score of $0.7-0.8$ indicates reasonable discrimination and values over $0.8$ indicate good discrimination.

We investigate the discriminative performance of both the raw NEWS score and the pre-defined "trigger" thresholds for low, medium, and high clinical risk (NEWS of $0-4$, $5-6$, and $\geq 7$ respectively) [RCP17, pp.30].

We measure correlation between ordinal variables (NEWS and LOS) using Spearman's rank correlation. We compare the means of independent samples of normally distributed variables using Welch's unequal variances t-test, and we use the Mann-Whitney U test for non-normally distributed variables. We present characteristics of variables as $mean \pm std$.

\section{Results}
\subsection{Population Characteristics} Records of $170833$ admissions were collated, of which $124162$ records, corresponding to $63450$ unique patients, had a NEW score recorded on-admission or computed by us during pre-processing. NEW scores ranges from $0-18$ with a median score of $1.0$ (IQR=$2.0$). Key characteristics of the dataset population are given in Table \ref{tab:population}.

\begin{table}[!t]
    \renewcommand{\arraystretch}{1.3}
    \centering
    \caption{SCI dataset key characteristics}
    \label{tab:population}
    \begin{tabular}{l l l}
        \hline\hline
        Parameter            & Category              & $n\;(\%)$           \\
        \hline
        Sex                  & Female                & $65096\;(52.43\%)$  \\\\
        Admission Method     & Emergency A\&E        & $113518\;(91.43\%)$ \\
                             & Emergency GP Referral & $5955\;(4.8\%)$     \\\\
        Admission Ward       & EAU                   & $82541\;(66.48\%)$  \\
                             & AEC                   & $36546\;(15.27\%)$  \\
                             & Other Medical         & $2980\;(2.4\%)$     \\
                             & High-Dependency       & $2095\;(1.69\%)$    \\\\
        Mortality            & No                    & $14768\;(92.43\%)$  \\
                             & Died During Stay      & $5308\;(4.28\%)$    \\
                             & Died within 30 Days   & $4086\;(3.29\%)$    \\\\
        Critical Care        & Yes                   & $4371\;(3.52\%)$    \\\\
        Readmitted (30 days) & Yes                   & $15836\;(12.75)$    \\\\
        NEWS clinical risk   & Low (0-4)             & $118223\;(95.1\%)$  \\
                             & Medium (5-6)          & $3748\;(3.02\%)$    \\
                             & High (>= 7)           & $2333\;(1.88\%)$    \\
        \hline\hline
    \end{tabular}
\end{table}

\subsection{Critical Event Occurrence} The original study on the NEWS' discriminative ability considered death within 24 hours of a given set of observations. We define a "Critical Events" as the occurrence of either critical care admission or death during the hospital spell. Since the vital signs and NEWS in our dataset are measured on admission, we investigate the subsets of hospital spells with total LOS within certain thresholds, such as 24 or 48 hours.

A total of $8826$ with recorded NEWS experienced a critical event, comprising $4371$ critical care admissions and $5308$ deaths. This was significantly correlated with the NEWS for the patient being over the defined threshold for high clinical risk ($p<0.0001$). The AUROCs ($95\%$ CI) of NEWS for critical events in records with 24h LOS, 48h LOS, or any LOS were respectively 0.914 (0.896-0.932), 0.890 (0.877-0.903), and 0.769 (0.763-0.775). The ROC curves are given in figure \ref{fig:criticalevent_roc}. The method for computing the AUROC confidence intervals is given in Appendix \ref{appendix:auroc}.

\begin{figure*}[htbp]
    \centering
    \includegraphics[width=\textwidth]{img/critical_event_roc.png}
    \caption{ROC curves of NEWS for discriminating critical event occurrence in stays with LOS under 24h (left), under 48h (middle), or any LOS (right).}
    \label{fig:criticalevent_roc}
\end{figure*}

Critical event occurrence significantly correlated with the NEWS at the same LOS thresholds: chi-square with NEWS $\geq 7$: $p < 0.0001$ in all cases. In total, $1423/8826 \; (16\%)$ of critical event occurrences had NEWS $\geq 7$. The full chi-square results are given in Appendix \ref{appendix:stats}. The same appendix gives the results for mortality and critical care admission examined separately.

Patients who experienced critical events were older ($71.72 \pm 16.85$ years; Welch's unequal variances t-test: $p < 0.0001$) and had higher NEWS ($3.51 \pm 3.1$; Mann-Whitney U test: $p < 0.0001$) compared to patients who did not (mean age $64.5 \pm 20.62$ years; NEWS $1.04 \pm 1.4$). The mean total LOS was $16.32 \pm 24.25$ for patients who experienced critical events and $6.07 \pm 13.62$ for those who did not (Mann-Whitney U test: $p < 0.0001$). Table \ref{tab:criticalevent_stats} gives the same information across the tested LOS thresholds.

\begin{table}[!t]
    \renewcommand{\arraystretch}{1.3}
    \centering
    \caption{Distribution comparison dep. on critical event occurrence.}
    \label{tab:criticalevent_stats}
    \begin{tabular}{l l l l l l}
        \hline\hline
        Variable & LOS & Mean (Event)      & Mean ($\lnot$Event) & Test  & $p$   \\
        \hline
        Age      & 24h & $71.4 \pm 18.92$  & $57.62 \pm 20.92$   & Welch & $0.0$ \\
                 & 48h & $70.54 \pm 19.56$ & $58.71 \pm 21.09$   &       & $0.0$ \\
                 & Any & $71.72 \pm 16.85$ & $64.5 \pm 20.62$    &       & $0.0$ \\
        NEWS     & 24h & $6.82 \pm 4.38$   & $0.75 \pm 1.07$     & Mann  & $0.0$ \\
                 & 48h & $5.91 \pm 4.18$   & $0.78 \pm 1.1$      &       & $0.0$ \\
                 & Any & $3.51 \pm 3.1$    & $1.04 \pm 1.4$      &       & $0.0$ \\
        \hline\hline
    \end{tabular}
\end{table}

\onecolumn
\appendices

\section{Data Pre-processing}
\label{appendix:preprocessing}

\section{AUROC}
\label{appendix:auroc}
We assess the discriminative ability of the NEWS using an area under the receiver-operating characteristics (AUROC) curve. This measurement is commonly used to assess the accuracy of a criterion variable whose value will be used to make a binary decision. We construct the ROC curve by plotting the false-positive rate ($1$ minus the specificity) on the x-axis against the true-positive rate (sensitivity) on the y-axis for each decision threshold for prediction between $0-100\%$. In our methodology, we use scikit-Learn for automatic plotting and AUC computation.

For large samples, the distribution of the AUC is approximately normal. Hence, we use the standard normal distribution to claculate the confidence interval for the AUC as:

\begin{equation}
    AUC \pm z_{\alpha/2}SE(AUC)
\end{equation}

The formula for $SE(AUC)$ is [Henley82]:

\begin{equation}
    SE(AUC) = \sqrt{\frac{AUC(1-AUC) + (N_1-1)(Q_1-AUC^2) + (N_2-1)(Q_2-AUC^2)}{N_1N_2}}
\end{equation}
where,

\begin{equation}
    Q_1 = \frac{AUC}{2-AUC}, \;
    Q_2 = \frac{2AUC^2}{1+AUC}
\end{equation}

\end{document}

### 2022-08-21 TeX/baseline/paper.tex
.

@@
-\documentclass[10pt,journal,compsoc]{IEEEtran}
+\documentclass[10pt,journal, compsoc]{IEEEtran}
 \IEEEoverridecommandlockouts
@@
 \usepackage{xcolor}
+\usepackage{booktabs, array, bm}
 
@@
 \section{Results}
+Table \ref{tab:result_summary} summarises our findings on the discriminative ability of NEWS for various outcomes.
+\begin{table}[!t]
+    \renewcommand{\arraystretch}{1.3}
+    \centering
+    \caption{Summary of Findings}
+    \label{tab:result_summary}
+    \begin{tabular}{lllr}
+        \toprule
+        Outcome        & Time         & NEWS $\geq 7$ & ROAUC (95\% CI)              \\
+        \midrule
+        Critical Event & 24h          & $51.13\%$     & $\bm{0.914}$ ($0.896-0.932$) \\
+                       & 48h          & $41.24\%$     & $0.890$ ($0.877-0.903$)      \\
+                       & Any          & $16.12\%$     & $0.769$ ($0.763-0.775$)      \\
+        \hline
+        Mortality      & 24h          & $62.01\%$     & $\bm{0.952}$ ($0.936-0.967$) \\
+                       & 48h          & $51.15\%$     & $0.938$ ($0.926-0.95$)       \\
+                       & Any          & $18.67\%$     & $0.770$ ($0.762-0.777$)      \\
+        \hline
+        Critical Care  & 24h          & $26.92\%$     & $\bm{0.831}$ ($0.787-0.874$) \\
+                       & 48h          & $24.64\%$     & $0.8$ ($0.772-0.829$)        \\
+                       & Any          & $15.1\%$      & $0.767$ ($0.758-0.775$)      \\
+        \hline
+        Long LOS       & $\geq$24h    & $2.50\%$      & $0.602$ ($0.598-0.605$)      \\
+                       & $\geq$48h    & $2.72\%$      & $\bm{0.607}$ ($0.604-0.61$)  \\
+                       & $\geq$7 days & $3.25\%$      & $0.598$ ($0.594-0.602$)      \\
+        \hline
+        30-Day Mor/ty  & 24h          & $3.68\%$      & $\bm{0.655}$ ($0.628-0.682$) \\
+                       & 48h          & $3.78\%$      & $0.655$ ($0.634-0.677$)      \\
+                       & Any          & $4.28\%$      & $0.629$ ($0.62-0.638$)       \\
+        \hline
+        Readmission    & 2 days       & $1.14\%$      & $0.496$ ($0.466-0.526$)      \\
+                       & 7 days       & $1.04\%$      & $0.503$ ($0.487-0.52$)       \\
+                       & 30 days      & $1.40\%$      & $\bm{0.532}$ ($0.522-0.541$) \\
+        \bottomrule
+    \end{tabular}
+\end{table}
 \subsection{Population Characteristics} Records of $170833$ admissions were collated, of which $124162$ records, corresponding to $63450$ unique patients, had a NEW score recorded on-admission or computed by us during pre-processing.
@@
     \begin{tabular}{l l l}
-        \hline\hline
+        \toprule
         Parameter            & Category              & $n\;(\%)$           \\
-        \hline
+        \midrule
         Sex                  & Female                & $65096\;(52.43\%)$  \\\\
@@
         Readmitted (30 days) & Yes                   & $15836\;(12.75)$    \\\\
-        NEWS clinical risk   & Low (0-4)             & $118223\;(95.1\%)$  \\
-                             & Medium (5-6)          & $3748\;(3.02\%)$    \\
-                             & High (>= 7)           & $2333\;(1.88\%)$    \\
-        \hline\hline
+        NEWS clinical risk   & Low ($0-4$)           & $118223\;(95.1\%)$  \\
+                             & Medium ($5-6$)        & $3748\;(3.02\%)$    \\
+                             & High ($\geq 7$)       & $2333\;(1.88\%)$    \\
+        \bottomrule
     \end{tabular}
@@
 A total of $8826$ with recorded NEWS experienced a critical event, comprising $4371$ critical care admissions and $5308$ deaths.
-This was significantly correlated with the NEWS for the patient being over the defined threshold for high clinical risk ($p<0.0001$).
-The AUROCs ($95\%$ CI) of NEWS for critical events in records with 24h LOS, 48h LOS, or any LOS were respectively 0.914 (0.896-0.932), 0.890 (0.877-0.903), and 0.769 (0.763-0.775).
+The AUROCs ($95\%$ CI) of NEWS for critical events in records with 24h LOS, 48h LOS, or any LOS were respectively $0.914$ ($0.896-0.932$), $0.890$ ($0.877-0.903$), and $0.769$ ($0.763-0.775$).
 The ROC curves are given in figure \ref{fig:criticalevent_roc}.
@@
 Critical event occurrence significantly correlated with the NEWS at the same LOS thresholds: chi-square with NEWS $\geq 7$: $p < 0.0001$ in all cases.
-In total, $1423/8826 \; (16\%)$ of critical event occurrences had NEWS $\geq 7$.
-The full chi-square results are given in Appendix \ref{appendix:stats}.
-The same appendix gives the results for mortality and critical care admission examined separately.
+In total, $1423/8826$ ($16\%$) of critical event occurrences had NEWS $\geq 7$.
+Appendix \ref{appendix:criticalevent_breakdown} examines the occurrence of mortality and critical care admission separately.
 
@@
     \label{tab:criticalevent_stats}
-    \begin{tabular}{l l l l l l}
-        \hline\hline
+    \begin{tabular}{llllll}
+        \toprule
         Variable & LOS & Mean (Event)      & Mean ($\lnot$Event) & Test  & $p$   \\
-        \hline
+        \midrule
         Age      & 24h & $71.4 \pm 18.92$  & $57.62 \pm 20.92$   & Welch & $0.0$ \\
@@
                  & Any & $3.51 \pm 3.1$    & $1.04 \pm 1.4$      &       & $0.0$ \\
-        \hline\hline
+
+        \bottomrule
     \end{tabular}
 \end{table}
+
+\subsection{Post-Discharge Mortality}
+The proportion of 30-day mortality in the dataset records was $3.29\%$ ($n=4086$) and corresponds to $6.44\%$ of patients.
+The AUROCs ($95\%$ CI) of NEWS for 30-day mortality in records with 24h LOS, 48h LOS, or any LOS were respectively $0.655$ ($0.628-0.682$), $0.656$ ($0.634-0.677$), and $0.629$ ($0.62-0.638$).
+Post-discharge mortality significantly correlated with NEWS at the same LOS thresholds: chi-square with NEWS $\geq 7$: $p < 0.0001$ in all cases.
+In total, $175/4090$ ($4.28\%$) of occurrences had NEWS $\geq 7$.
+
+Patients who died within 30 days were older ($77.05 \pm 13.10$ years; Welch's unequal variances t-test: $p < 0.0001$) and had higher NEWS ($2.0 \pm 2.11$; Mann-Whitney U test: $p < 0.0001$) compared to patients who survived (mean age $64.6 \pm 20.54$ years; NEWS $1.19 \pm 1.68$).
+The mean total LOS was $13.49 \pm 18.99$ for patients who died after discharge and $6.57 \pm 14.65$ for survivors (Mann-Whitney U test: $p < 0.0001$).
+Table \ref{tab:criticalevent_stats} gives the same information across the tested LOS thresholds.
 
@@
 
+\section{Results for Mortality and Critical Care Admission}
+\label{appendix:criticalevent_breakdown}
+\subsection{Mortality}
+The proportion of in-hospital mortality in the dataset records was $4.28\%$ ($n=5328$) and corresponds to $8.386\%$ of patients.
+The AUROCs ($95\%$ CI) of NEWS for mortality in records with 24h LOS, 48h LOS, or any LOS were respectively $0.952$ ($0.936-0.967$), $0.938$ ($0.926-0.95$), and $0.770$ ($0.762-0.777$).
+Mortality significantly correlated with NEWS at the same LOS thresholds: chi-square with NEWS $\geq 7$: $p < 0.0001$ in all cases.
+In total, $955/5328$ ($17.9\%$) of occurrences had NEWS $\geq 7$.
+\begin{figure*}[htbp]
+    \centering
+    \includegraphics[width=\textwidth]{img/mortality_roc.png}
+    \caption{ROC curves of NEWS for discriminating in-hospital mortality with LOS under 24h (left), under 48h (middle), or any LOS (right).}
+    \label{fig:mortality_roc}
+\end{figure*}
+
+\subsection{Critical Care Admission}
+The proportion of critical care admission in the dataset records was $3.53\%$ ($n=4384$) and corresponds to $6.90\%$ of patients.
+The AUROCs ($95\%$ CI) of NEWS for critical care admission in records with 24h LOS, 48h LOS, or any LOS were respectively $0.831$ ($0.787-0.874$), $0.8$ ($0.772-0.829$), and $0.767$ ($0.758-0.775$).
+Admission significantly correlated with NEWS at the same LOS thresholds: chi-square with NEWS $\geq 7$: $p < 0.0001$ in all cases.
+In total, $662/4384$ ($15.1\%$) of occurrences had NEWS $\geq 7$.
+\begin{figure*}[htbp]
+    \centering
+    \includegraphics[width=\textwidth]{img/criticalcare_roc.png}
+    \caption{ROC curves of NEWS for discriminating critical care admission in stays with LOS under 24h (left), under 48h (middle), or any LOS (right).}
+    \label{fig:criticalcare_roc}
+\end{figure*}
+
 \end{document}

### 2022-08-21 TeX/baseline/paper.tex
.

@@
 % The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
-\usepackage{cite}
 \usepackage{amsmath,amssymb,amsfonts}
@@
 \usepackage{booktabs, array, bm}
+
+\usepackage[
+    backend=biber,
+    style=numeric,
+    url=false,
+    isbn=false,
+    doi=false
+  ]{biblatex}
+\addbibresource{references.bib}
 
@@
 A common practice to gauge patients' physiological condition and acute clinical stability is using basic homeostatic measures like vital signs, haematologic findings, and neurological examinations.
-Various early warning systems have been proposed that use these measurements to assess whether patients face an imminent risk of deterioration or death [Smith13].
+Various early warning systems have been proposed that use these measurements to assess whether patients face an imminent risk of deterioration or death \cite{Smith13}.
 
-The National Early Warning Score (NEWS), developed in conjunction with the Royal College of Physicians, has been widely implemented across the NHS and by healthcare providers outside the UK [RCP17, pp.13].
+The National Early Warning Score (NEWS), developed in conjunction with the Royal College of Physicians, has been widely implemented across the NHS and by healthcare providers outside the UK \cite[pp.~13]{RCP17}.
 It is a points-based early warning system to identify clinical deterioration in acutely ill patients based on routinely recorded vital signs: pulse, breathing rate, blood pressure, blood oxygen level, level of consciousness, and temperature.
@@
 While initially designed to monitor for deterioration in secondary-care, the NEWS has been widely tested and validated across healthcare settings.
-In its latest revision of the standard, the RCP recommended that the NEWS also be used in emergency departments to "aid the initial assessment of patiens, ongoing monitoring and patient triage decisions" [RCP17, pp.18].
+In its latest revision of the standard, the RCP recommended that the NEWS also be used in emergency departments to "aid the initial assessment of patiens, ongoing monitoring and patient triage decisions" \cite[pp.~18]{RCP17}.
 
@@
 
-Our dataset includes records of patients who received ambulatory emergency care (AEC) or same-day emergency care (SDEC) and were discharged before midnight on the day of admission [NHS18].
+Our dataset includes records of patients who received ambulatory emergency care (AEC) or same-day emergency care (SDEC) and were discharged before midnight on the day of admission \cite{NHS18}.
 AEC, alongside the medical admissions unit (EAU)  are the two common entry points for adult patients with general medical emergencies.
@@
 For parameters that have a recorded NEWS sub-score but their raw value is missing or marked as spurious, we infer the correct value as the midpoint of the relevant NEWS range.
-Where a parameter has a valid value but the NEWS sub-score is missing, we compute it in accordance with the specificaton [RCP17, pp.].
+Where a parameter has a valid value but the NEWS sub-score is missing, we compute it in accordance with the specificaton \cite[pp.~28-30]{RCP17}.
 We use the same specification to compute the final NEWS score for any patients where this is not recorded in the original data.
@@
 
-We investigate the discriminative performance of both the raw NEWS score and the pre-defined "trigger" thresholds for low, medium, and high clinical risk (NEWS of $0-4$, $5-6$, and $\geq 7$ respectively) [RCP17, pp.30].
+We investigate the discriminative performance of both the raw NEWS score and the pre-defined "trigger" thresholds for low, medium, and high clinical risk (NEWS of $0-4$, $5-6$, and $\geq 7$ respectively) \cite[pp.~30]{RCP17}.
 
@@
         \hline
-        Readmission    & 2 days       & $1.14\%$      & $0.496$ ($0.466-0.526$)      \\
-                       & 7 days       & $1.04\%$      & $0.503$ ($0.487-0.52$)       \\
-                       & 30 days      & $1.40\%$      & $\bm{0.532}$ ($0.522-0.541$) \\
+        Readmission    & 2 days       & $1.67\%$      & $0.495$ ($0.478-0.511$)      \\
+                       & 7 days       & $1.06\%$      & $0.503$ ($0.482-0.501$)      \\
+                       & 30 days      & $1.31\%$      & $0.523$ ($0.518-0.528$)      \\
         \bottomrule
@@
 
+\subsection{Length of Stay} The total LOS for the patient was significantly correlated with NEWS (Spearman Rank Correlation: $p < 0.0001$).
+Median length of stay more than doubled for NEWS $\geq 7$ compared to $<7$ ($5.90$ and $2.21$, respectively).
+
+We further investigate whether NEWS can discriminate stays where the LOS will be greater than a certain amount of time; for example, LOS over one day, meaning the patient needs to be monitored overnight.
+The AUROCs for LOS $\geq$24h, $\geq$48h, and $\geq$7 days were $0.602$ ($0.598-0.605$), $0.607$ ($0.604-0.61$), and $0.598$ ($0.594-0.602$) respectively.
+LOS at the measured thresholds significantly correlated with NEWS: chi-square with NEWS $\geq 7$: $p < 0.0001$ in all cases.
+
 \subsection{Post-Discharge Mortality}
@@
 Table \ref{tab:criticalevent_stats} gives the same information across the tested LOS thresholds.
+
+\subsection{Readmission} The proportion of 30-day readmissions in the dataset was $11.26\%$ (n=$14003$).
+The AUROCs ($95\%$ CI) of NEWS for readmission within 2 days, 7 days, and 30 days were respectively $0.495$ ($0.478-0.511$), $0.491$ ($0.482-0.501$), and $0.523$ ($0.518-0.528$), indicating the NEWS is no better at predicting readmission than random chance.
+NEWS was not significantly correlated with readmission at any time threshold: chi-square with NEWS $\geq 7$: $p > 0.23$ in all cases.
+
+\section{Discussion}
+We have investigated applying the NEWS to a large vital signs dataset from a single acute care trust, mainly consisting of emergency admissions and not considering elective procedures or surgical admissions.
+
+The original study by the NEWSDIG to evaluate the then-newly-proposed NEWS tracked death within 24 hours of readings and reported an AUROC score of $0.89$.
+A subsequent study in emergency departments in the Netherlands gave an AUROC score of $0.768$ for 30-day mortality based on the admission NEWS.
+
+This report evaluated the NEWS against the combined outcome of unplanned critical care admission and death.
+We found it appropriate to consider these outcomes as one, as we expect both to be preceded by deranged physiology and the clinical response to both events, in terms of urgency and skill, is very similar \cite[pp.~4]{Smith13}.
+We did not track the occurrence of cardiac arrest, as this was not a parameter recorded in our data.
+However, the foundational NEWS study by \cite{Smith13} concluded this outcome was of limited use as it is largely indistinguishable from mortality except for the fact that a cardiac arrest team has been called in \cite[pp.~4]{Smith13}.
+
+Our findings regarding the NEWS' discriminative performance are consistent with the previous studies.
+The NEWS performed best in detecting critical events (in-hospital mortality or admission to critical care).
+It showed potential in identifying patients at risk of extended hospital stays or post-discharge mortality.
+The NEWS was not effective in detecting which patients in our data would be readmitted to the same hospital in the future within any amount of time.
+
+We showed that parameters present in the data other than vital signs (age) differ in their distribution between patients who experienced adverse outcomes and those who did not.
+In their comparison of EWS, \cite{Smith13} found that existing EWS systems that considered age showed poor discriminative ability in critical care admission compared to death.
+The authors noted that the observed performance of an EWS on predicting mortality might not line up with the "true" desired property of an EWS: to effectively recognise patients whose condition is deteriorating but can be improved by timely medical intervention rather than those who will likely not survive \cite[pp.~4]{Smith13}.
+In our view, this makes a strong case for ensuring good discrimination of patients in terms of other adverse outcomes, such as critical care admission, post-discharge mortality, and lengthy hospital stays, to inform and assist clinical decision-making.
+
+In their specification, the NEWSDIG determined it was unnecessary to apply arbitrary weighting to the score based on the patient's age \cite[pp.~19]{RCP17}.
+Similarly, the NEWS does not consider comorbidities as it is intended to be generic and reflect the physiological perturbations any condition may cause.
+Given the differing health needs and expected physiological state of patients depending on these excluded factors, we leave the investigation of alternative, stratified early-warning systems as future work.
+
+\subsection{Limitations}
+Our observational dataset is limited to only a single acute care trust.
+Making definitive statements about the utility of the NEWS requires further validation in different patient populations and clinical settings.
+
+Patient data were collected primarily by staff under everyday conditions.
+While this matches the real-life use of the NEWS, operational pressures may hinder the timeliness of data entry and the dataset's reliability.
+Further, we did not track whether patients had a DNAR order, which can lead to patient death in cases where it otherwise may not have occurred.
+
+A limitation of the dataset was that it only records NEWS measurements taken on admission.
+To compare with prior studies, we considered adverse event occurrence in hospital stays that lasted under 24 or 48 hours as a proxy for testing NEWS values measured shortly before these events.
+This practice may induce a bias in our results compared to evaluating the NEWS on data with repeated sets of observations for each patient throughout their hospital stay.
+
+\printbibliography
 
@@
 \end{equation}
-
+\newpage
 \section{Results for Mortality and Critical Care Admission}
@@
 
+\subsection{ROC for Length of Stay, Post-Discharge Mortality, and Readmission}
+\label{appendix:rocplots}
+\begin{figure*}[htbp]
+    \centering
+    \includegraphics[width=\textwidth]{img/los_roc.png}
+    \caption{ROC curves of NEWS for discriminating LOS $\geq$ 24h (left), 48h (middle), or 7 days (right).}
+    \label{fig:los_roc}
+\end{figure*}
+
+\begin{figure*}[htbp]
+    \centering
+    \includegraphics[width=\textwidth]{img/30daymortality_roc.png}
+    \caption{ROC curves of NEWS for discriminating post-discharge mortality in stays with LOS under 24h (left), under 48h (middle), or any LOS (right).}
+    \label{fig:30day_roc}
+\end{figure*}
+
+\begin{figure*}[htbp]
+    \centering
+    \includegraphics[width=\textwidth]{img/readmission_roc.png}
+    \caption{ROC curves of NEWS for discriminating readmission within 2 days (left), 7 days (middle), or 30 days (right).}
+    \label{fig:readmission_roc}
+\end{figure*}
+
+\section{Discussion}
+
 \end{document}

### 2022-08-21 dataset.py

@@ -218,6 +218,7 @@ class SCIData(pd.DataFrame):
         r["Readmitted"] = (
             timespan_reverse > pd.Timedelta(days=-readmission_thresh)
         ).fillna(False)
+        r["ReadmittedTimespan"] = timespan_reverse*(-1)
 
         return SCIData(data=r)
 

### 2022-08-21 TeX/baseline/paper.tex
.

@@
 
-\title{Conference Paper Title\\
-    % {\footnotesize \textsuperscript{*}Note: Sub-titles are not captured in Xplore and
-    % should not be used}
+\title{Evaluation of NEWS on the Salford Dataset:\\
+{ACP Baseline Model Report}
 % \thanks{Identify applicable funding agency here. If none, delete this.}
@@
 
-\begin{abstract}
-    This document is a model and instructions for \LaTeX.
-    This and the IEEEtran.cls file define the components of your paper [title, text, heads, etc.]. *CRITICAL: Do Not Use Symbols, Special Characters, Footnotes,
-    or Math in Paper Title or Abstract.
-\end{abstract}
+% \begin{abstract}
+%     This document is a model and instructions for \LaTeX.
+%     This and the IEEEtran.cls file define the components of your paper [title, text, heads, etc.]. *CRITICAL: Do Not Use Symbols, Special Characters, Footnotes,
+%     or Math in Paper Title or Abstract.
+% \end{abstract}
 
@@
 Across studies, the NEWS has been found to perform well in discriminating ward patients at risk of death, critical care admission, or cardiac arrest.
-The present report introduces the novel SCI dataset, which contains records of adult patients with acute or medical presentation at the Salford Royal NHS Foundation Trust from 2014-2022.
+The present report introduces the novel Salford dataset, which contains records of adult patients with acute or medical presentation at the Salford Royal NHS Foundation Trust from 2014-2022.
 We then use this dataset to assess the performance of NEWS in predicting adverse patient outcomes, such as those mentioned above, in the dataset population.
+
+Further, we investigate the performance of NEWS on post-discharge mortality, lengthy hospital stay, and readmission within certain timeframes.
+Anticipating these outcomes is in keeping with the main purpose of an EWS: to effectively recognise patients whose condition is deteriorating but can be improved by timely medical intervention  \cite[pp.~4]{Smith13}.
 
@@
 \subsection{Data Collection and Contents}
-The SCI dataset is collated from EPR records collected in real time from admissions to acute beds between 1st April 2014 and 31st March 2022.
+The Salford dataset is collated from EPR records collected in real time from admissions to acute beds between 1st April 2014 and 31st March 2022.
 Further, the dataset includes administrative data produced in retrospect for financial (tarrif) purposes.
@@
     \centering
-    \caption{SCI dataset key characteristics}
+    \caption{Salford dataset key characteristics}
     \label{tab:population}
@@
     \centering
-    \includegraphics[width=\textwidth]{img/critical_event_roc.png}
+    \includegraphics[width=\textwidth]{img/critical_roc_multi.png}
     \caption{ROC curves of NEWS for discriminating critical event occurrence in stays with LOS under 24h (left), under 48h (middle), or any LOS (right).}
@@
 It showed potential in identifying patients at risk of extended hospital stays or post-discharge mortality.
+
 The NEWS was not effective in detecting which patients in our data would be readmitted to the same hospital in the future within any amount of time.
+Prior work has found that clinical data indicating the severity of the patient's condition, including constructed scores, is useful in predicting 30-day readmission \cite[pp.~5]{Mahmoudim20}.
+Dedicated systems for predicting 30-day readmission, such as the HOSPITAL and LACE score, use parameters that the NEWS does not, such as blood test results, procedures the patient underwent, comorbidities and LOS \cite{Donze13}.
 
@@
 Our observational dataset is limited to only a single acute care trust.
+Some measured outcomes and population parameters are inherently variable by hospital.
+For example, critical care admission may vary between hospitals because each hospital's criteria for admission are different.
 Making definitive statements about the utility of the NEWS requires further validation in different patient populations and clinical settings.
@@
 \appendices
-
+\setlength{\parskip}{\baselineskip}%
+\setlength{\parindent}{0pt}%
 \section{Data Pre-processing}
 \label{appendix:preprocessing}
+For the components of the NEWS, we set the ranges given in Table \ref{tab:news_valid_ranges} and we consider values outside these ranges to be invalid.
+Where a value is missing or falls outside these ranges, we replace it with a value inferred from the recorded NEWS sub-score.
+Specifically, we set it to be the midpoint of the relevant NEWS range.
+Where the sub-score has not been recorded but the raw value is present (and not invalid), we use the value to compute the NEWS sub-score ourselves.
+
+\begin{table}[h]
+    \renewcommand{\arraystretch}{1.3}
+    \centering
+    \caption{Valid ranges for NEWS components}
+    \label{tab:news_valid_ranges}
+    \begin{tabular}{lll}
+        \toprule
+        Variable         & Range    & Unit        \\
+        \midrule
+        SpO2             & $40-100$ & $\%$        \\
+        Systolic BP      & $40-300$ & mmHg        \\
+        Temperature      & $25-45$  & $^{\circ}C$ \\
+        Pulse            & $25-300$ & Beats/min   \\
+        Respiration Rate & $5-80$   & Breaths/min \\
+        \bottomrule
+    \end{tabular}
+\end{table}
+
+\textbf{O2 Saturation (SpO2): } The NEWS-2 specification gives two scales for this parameter \cite[pp.~44]{RCP17}:
+\begin{itemize}
+    \item $SpO2_1$: By default.
+    \item $SpO2_2$: For patients with a prescribed oxygen saturation requirement of $88-92\%$ (e.g., in patients with hypercapnic respiratory failure).
+\end{itemize}
+Since the choice of scale is determined by the responsible clinical staff on a case-by-case basis, we use the following criteria to infer which scale to use when re-computing the NEWS sub-score for SpO2:
+\begin{itemize}
+    \item Anyone receiving oxygen using NIV, as recorded directly or in their set of coded procedures ('E85.2').
+    \item Anyone with COPD (as determined by presence of 'J44.*' in their coded diagnoses) AND is receiving oxygen via Venturi 24 or 28.
+    \item Anyone with COPD (as above) and measured SpO2 $<88\%$.
+\end{itemize}
+Finally, if the patient is receiving supplemental oxygen, there is ambiguity as to whether a high O2 NEWS sub-score indicates very high or very low saturation.
+In that case, we mark the value as missing.
+
+\textbf{Respiration Rate: } In addition to the range given in Table \ref{tab:news_valid_ranges}, we assume triple-digit values to be erroneous entries of two-digit values (e.g., $250 \rightarrow 25.0$).
+
+\textbf{Oxygen Flow Rate: } This supplemental parameter is enterred in mixed units (Litres/min or FiO2). We translate all values to FiO2 where possible:
+\begin{itemize}
+    \item Values $1-15$ are inferred to be in Litres/min.
+    \item Decimal values are inferred to be FiO2, with the exception of $0.5$.
+    \item Values of $0.5$ and any remaining values are determine based on the device used to deliver the oxygen.
+Nasal cannula and simple mask correspond to Litres/min, while other devices correspond to FiO2.
+\end{itemize}
+We convert Litres/min to FiO2 using the formula $FiO2 = 0.2 + (Litres/min * 4)/100$.
+
+\newpage
 
@@
 
-The formula for $SE(AUC)$ is [Henley82]:
+The formula for $SE(AUC)$ is \cite{Hanley82}:
 
@@
 \newpage
-\section{Results for Mortality and Critical Care Admission}
+\section{Extended Results}
 \label{appendix:criticalevent_breakdown}
-\subsection{Mortality}
+\subsection{Mortality \& Critical Care Admission}
 The proportion of in-hospital mortality in the dataset records was $4.28\%$ ($n=5328$) and corresponds to $8.386\%$ of patients.
@@
 In total, $955/5328$ ($17.9\%$) of occurrences had NEWS $\geq 7$.
-\begin{figure*}[htbp]
-    \centering
-    \includegraphics[width=\textwidth]{img/mortality_roc.png}
-    \caption{ROC curves of NEWS for discriminating in-hospital mortality with LOS under 24h (left), under 48h (middle), or any LOS (right).}
-    \label{fig:mortality_roc}
-\end{figure*}
 
-\subsection{Critical Care Admission}
 The proportion of critical care admission in the dataset records was $3.53\%$ ($n=4384$) and corresponds to $6.90\%$ of patients.
@@
 In total, $662/4384$ ($15.1\%$) of occurrences had NEWS $\geq 7$.
-\begin{figure*}[htbp]
-    \centering
-    \includegraphics[width=\textwidth]{img/criticalcare_roc.png}
-    \caption{ROC curves of NEWS for discriminating critical care admission in stays with LOS under 24h (left), under 48h (middle), or any LOS (right).}
-    \label{fig:criticalcare_roc}
-\end{figure*}
 
@@
 
-\section{Discussion}
-
 \end{document}

### 2022-08-22 dataset.py
.

@@ -62,35 +62,100 @@ class SCIData(pd.DataFrame):
             .clean_news()
         )
 
-    def augment_ccs(
-        self,
-        infile="data/ccs.h5",
-        ccs=None,
-        sentinel=[259, "Residual codes; unclassified"],
-    ):
-        """ Joins SCI and the CCS reference table fuzzily.
+    # def augment_ccs_main_only(
+    #     self,
+    #     infile="data/ccs.h5",
+    #     ccs=None,
+    #     sentinel=[259, "Residual codes; unclassified"],
+    # ):
+    #     """ Joins SCI and the CCS reference table fuzzily.
+    #     Codes that can't be matched exactly are matched with their 3-codes
+    #     :param infile: The HDF5 file to load the CCS matching table from
+    #     :param ccs: Pre-loaded CCS match table dataframe
+    #     :param sentinel: What to fill un-matched values with. Default is the code for unclassified/residual codes
+    #     :returns: New SCIData instance with augmented with CCS codes and descriptions
+    #     """
+    #     if ccs is None:
+    #         ccs = pd.read_hdf(infile, "codes")
+
+    #     no_dot = self.MainICD10.str.replace(".", "")
+    #     # Perfect full-ICD10 matches
+    #     r = self.join(ccs, on=no_dot)
+
+    #     # Approximate matches based on first 4 characters (e.g. M4479) or first 3 (e.g. A125)
+    #     approx = (self.join(ccs, on=no_dot.str[:_]) for _ in (3, 4))
+    #     for df in approx:
+    #         mask = df.CCSGroup.notna() & r.CCSGroup.isna()
+    #         r.loc[mask, ccs.columns] = df.loc[mask, ccs.columns]
+
+    #     # Fill remaining values
+    #     remaining = r.CCSGroup.isna() & r.MainICD10.notna()
+    #     r.loc[remaining, ccs.columns] = sentinel
+
+    #     return SCIData(r)
+
+    def augment_ccs(self, infile="data/ccs.h5", ccs=None, sentinel=259, drop_old=True):
+        """ Joins SCI and the CCS reference table fuzzily. Applies to ALL coded diagnoses, not just MainICD10
         Codes that can't be matched exactly are matched with their 3-codes 
         :param infile: The HDF5 file to load the CCS matching table from
         :param ccs: Pre-loaded CCS match table dataframe
[…]

### 2022-09-13 cross_validation.py
.

@@ -4,6 +4,15 @@ from functools import partial
 from sklearn.model_selection import cross_validate
 
 
+def cross_validate_linear(models, X, y, scoring, cv=3):
+    keys, values = zip(*models.items())
+    return dict(
+        zip(
+            keys, map(partial(cross_validate, X=X, y=y, cv=cv, scoring=scoring, n_jobs=-1), values)
+        )
+    )
+
+
 def cross_validate_parallel(models, X, y, scoring, cv=3, threads=None):
     if not threads:
         threads = len(models)
@@ -19,10 +28,10 @@ def cross_validate_parallel(models, X, y, scoring, cv=3, threads=None):
 
 
 def _run_cv(kv, X, y, scoring, cv):
-    print(f'STARTING {kv[0]}', flush = True)
+    print(f"STARTING {kv[0]}", flush=True)
     try:
         r = cross_validate(kv[1], X=X, y=y, cv=cv, scoring=scoring)
-        print(f'FINISHED {kv[0]}', flush = True)
+        print(f"FINISHED {kv[0]}", flush=True)
         return [
             {"model": kv[0], "err": False, **{k: v[i] for k, v in r.items()}}
             for i in range(cv)

### 2022-09-13 dataset.py

@@ -36,8 +36,12 @@ class SCIData(pd.DataFrame):
         return cls(pd.concat([X, pd.Series(y, name="y")], axis=1))
 
     def save(self, filename="data/sci_processed.h5"):
-        self.to_hdf(filename, "table")
-        return self
+        r = self.copy()
+        mask = r.select_dtypes(include="category")
+        r[mask.columns] = mask.astype(str)
+
+        r.to_hdf(filename, "table")
+        return SCIData(r)
 
     def derive_all(self):
         """Runs all methods to derive handcrafted features from the raw dataset
@@ -238,19 +242,17 @@ class SCIData(pd.DataFrame):
         :param labels: String labels for the returned bands
         :return: New SCIData instance with the new features added
         """
-        bins = [pd.Timedelta(days=_) for _ in bins]
+        bins = [pd.Timedelta(days=_) for _ in [-1] + bins]
+        labels = labels + ["N/A"]
 
         r = self.copy()
         grouped = r.sort_values(["PatientNumber", "AdmissionDateTime"]).groupby(
             "PatientNumber"
         )
         r["ReadmissionTimespan"] = grouped.AdmissionDateTime.diff().dropna()
-        r["ReadmissionBand"] = (
-            pd.cut(r.ReadmissionTimespan, bins, labels=labels)
-            .astype(str)
-            .replace("nan", np.nan)
-            .fillna("N/A")
-        )
+        r["ReadmissionBand"] = pd.cut(
+            r.ReadmissionTimespan, bins, labels=labels, ordered=False
+        ).fillna("N/A")
         r["Readmission"] = r.ReadmissionTimespan < pd.Timedelta(days=readmission_thresh)
 
         timespan_reverse = grouped.AdmissionDateTime.diff(-1).dropna()
@@ -260,6 +262,17 @@ class SCIData(pd.DataFrame):
         r["ReadmittedTimespan"] = timespan_reverse * (-1)
 
         return SCIData(data=r)
+
+    def fix_readmissionband(
+        self,
+        labels=["24 Hrs", "48 Hrs", "1 Week", "2 Weeks", "1 Month", "2 Months", "N/A"],
+    ):
+        r = self.copy()
+        r.ReadmissionBand = r.ReadmissionBand.astype(
+            pd.CategoricalDtype(labels, ordered=False)
+        )
+
+        return SCIData(r)
 
     def derive_mortality(self, within=1, col_name="DiedWithinThreshold"):
[…]

### 2022-09-13 dataset.py
.

@@ -905,23 +905,31 @@ class SCIData(pd.DataFrame):
         return SCIData(r)
 
     def describe_categories(self):
+        categorical_cols = [col for col, type in get_column_types().items() if type=='c']
+
         categorical_cols_idx = [
             self.columns.get_loc(_)
-            for _ in self.select_dtypes(include="category").columns
+            for _ in categorical_cols
         ]
         categorical_dims = list(
-            self.select_dtypes(include="category")
-            .apply(lambda x: x.cat.categories.shape[0])
-            .values
+            self[categorical_cols].apply(lambda x: x.cat.categories.shape[0]).values
         )
 
         return categorical_cols_idx, categorical_dims
+
+    def get_column_types(self):
+        return {
+            **{_:'q' for _ in self.columns},
+            **{_:'i' for _ in self.columns if _.startswith('CCS_')},
+            **SCICols.xgb_types
+        }
 
     def xy(
         self,
         x=[],
         dtype=None,
         dropna=False,
+        fillna=False,
         ordinal_encoding=False,
         outcome="DiedDuringStay",
     ):
@@ -941,6 +949,10 @@ class SCIData(pd.DataFrame):
 
         X = X.apply(lambda x: x.replace({True: 1.0, False: 0.0}))
 
+        if fillna:
+            X.select_dtypes(include='number').fillna(-1, inplace=True)
+            X.select_dtypes(include='object').fillna('NAN', inplace=True)
+        
         X = SCIData(X).categorize()
 
         if ordinal_encoding:
@@ -1381,3 +1393,44 @@ class SCICols:
             + SCICols.news
             + SCICols.news_data
         )
+
+    xgb_types = {
+        'Female': 'i',
+        'Age': 'int',
+        'ElectiveAdmission': 'i',
+        'AdmissionMethodDescription': 'c',
+        'AdmissionSpecialty': 'c',
+        'AgeBand': 'c',
+        'AssessmentAreaAdmission': 'i',
+        'LastSpecialty': 'c',
+        'MainICD10': 'c',
+        'SecDiag1': 'c',
+        'SecDiag2': 'c',
+        'SecDiag3': 'c',
+        'SecDiag4': 'c',
+        'SecDiag5': 'c',
+        'SecDiag6': 'c',
+        'Haemoglobin': 'q',
[…]

### 2022-09-13 evaluation.py

@@ -1,0 +1,53 @@
+import pandas as pd
+from IPython.display import display
+import matplotlib.pyplot as plt
+
+from sklearn.metrics import (
+    accuracy_score,
+    precision_score,
+    recall_score,
+    roc_auc_score,
+    f1_score,
+    fbeta_score,
+    make_scorer,
+    precision_recall_curve,
+    ConfusionMatrixDisplay,
+)
+
+f2_score = make_scorer(fbeta_score, beta=2)
+
+METRICS = {
+    "Accuracy": "accuracy",
+    "Precision": "precision",
+    "Recall": "recall",
+    "AUC": "roc_auc",
+    "F1 Score": "f1",
+    "F2 Score": f2_score,
+}
+
+def evaluate_from_pred(y_true, y_pred, y_pred_proba):
+    display(
+        pd.DataFrame(
+            {
+                "Accuracy": accuracy_score(y_true, y_pred),
+                "Precision": precision_score(y_true, y_pred),
+                "Recall": recall_score(y_true, y_pred),
+                "AUC": roc_auc_score(y_true, y_pred_proba[:, 1]),
+                "F1 Score": f1_score(y_true, y_pred),
+                "F2 Score": fbeta_score(y_true, y_pred, beta=2),
+            },
+            index=["Model"],
+        )
+    )
+
+    fig, ax = plt.subplots(1, 1, figsize=(5, 5))
+    ax.grid(False)
+    ConfusionMatrixDisplay.from_predictions(y_true, y_pred, ax=ax)
+
+
+
+def evaluate(model, X, y):
+    y_pred = model.predict(X)
+    y_pred_proba = model.predict_proba(X)
+
+    evaluate_from_pred(y, y_pred, y_pred_proba)

### 2022-09-13 dataset.py
.

@@ -861,7 +861,7 @@ class SCIData(pd.DataFrame):
         )
 
     def mandate_diagnoses(self):
-        return self.mandate(SCICols.diagnoses_ccs_encoded + SCICols.diagnoses)
+        return self.mandate([_ for _ in self.columns if _.startswith('CCS_')] + SCICols.diagnoses)
 
     def mandate_news(self):
         return self.mandate(SCICols.news_data)
@@ -905,15 +905,13 @@ class SCIData(pd.DataFrame):
         return SCIData(r)
 
     def describe_categories(self):
-        categorical_cols = [col for col, type in get_column_types().items() if type=='c']
+        categorical_cols = [_ for _ in self.columns if self.get_column_types()[_] == 'c']
 
         categorical_cols_idx = [
             self.columns.get_loc(_)
             for _ in categorical_cols
         ]
-        categorical_dims = list(
-            self[categorical_cols].apply(lambda x: x.cat.categories.shape[0]).values
-        )
+        categorical_dims = [self[_].unique().shape[0] for _ in categorical_cols]
 
         return categorical_cols_idx, categorical_dims
 

### 2022-09-13 evaluation.py

@@ -25,14 +25,14 @@ METRICS = {
     "F2 Score": f2_score,
 }
 
-def evaluate_from_pred(y_true, y_pred, y_pred_proba):
+def evaluate_from_pred(y_true, y_pred, y_pred_proba, plot_title=None):
     display(
         pd.DataFrame(
             {
                 "Accuracy": accuracy_score(y_true, y_pred),
                 "Precision": precision_score(y_true, y_pred),
                 "Recall": recall_score(y_true, y_pred),
-                "AUC": roc_auc_score(y_true, y_pred_proba[:, 1]),
+                "AUC": roc_auc_score(y_true, y_pred_proba),
                 "F1 Score": f1_score(y_true, y_pred),
                 "F2 Score": fbeta_score(y_true, y_pred, beta=2),
             },
@@ -42,12 +42,16 @@ def evaluate_from_pred(y_true, y_pred, y_pred_proba):
 
     fig, ax = plt.subplots(1, 1, figsize=(5, 5))
     ax.grid(False)
+    ax.set_title(plot_title)
     ConfusionMatrixDisplay.from_predictions(y_true, y_pred, ax=ax)
 
 
 
 def evaluate(model, X, y):
     y_pred = model.predict(X)
-    y_pred_proba = model.predict_proba(X)
+    try:
+        y_pred_proba = model.predict_proba(X)[:, 1]
+    except AttributeError:
+        y_pred_proba = model.score_samples(X)
 
     evaluate_from_pred(y, y_pred, y_pred_proba)

### 2022-09-14 evaluation.py
Uploaded ML algorithm tests

@@ -12,6 +12,7 @@ from sklearn.metrics import (
     make_scorer,
     precision_recall_curve,
     ConfusionMatrixDisplay,
+    RocCurveDisplay,
 )
 
 f2_score = make_scorer(fbeta_score, beta=2)
@@ -24,6 +25,7 @@ METRICS = {
     "F1 Score": "f1",
     "F2 Score": f2_score,
 }
+
 
 def evaluate_from_pred(y_true, y_pred, y_pred_proba, plot_title=None):
     display(
@@ -40,18 +42,20 @@ def evaluate_from_pred(y_true, y_pred, y_pred_proba, plot_title=None):
         )
     )
 
-    fig, ax = plt.subplots(1, 1, figsize=(5, 5))
-    ax.grid(False)
-    ax.set_title(plot_title)
-    ConfusionMatrixDisplay.from_predictions(y_true, y_pred, ax=ax)
+    fig, ax = plt.subplots(1, 2, figsize=(11, 5))
+
+    ax[1].grid(False)
+    ConfusionMatrixDisplay.from_predictions(y_true, y_pred, ax=ax[1])
+    RocCurveDisplay.from_predictions(y_true, y_pred_proba, ax=ax[0])
+
+    plt.suptitle(plot_title)
 
 
-
-def evaluate(model, X, y):
+def evaluate(model, X, y, plot_title=None):
     y_pred = model.predict(X)
     try:
         y_pred_proba = model.predict_proba(X)[:, 1]
     except AttributeError:
         y_pred_proba = model.score_samples(X)
 
-    evaluate_from_pred(y, y_pred, y_pred_proba)
+    evaluate_from_pred(y, y_pred, y_pred_proba, plot_title)

### 2022-09-14 old_code/ml_script1.py

@@ -1,0 +1,173 @@
+# %%
+import warnings, math, itertools, json, logging
+from hashlib import md5
+
+import numpy as np
+import pandas as pd
+
+from IPython.display import display
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+
+# %%
+pd.set_option("display.max_columns", None)
+sns.set_theme(style="darkgrid", palette="colorblind")
+sns.set(rc={"figure.figsize": (11.5, 8.5), "figure.dpi": 100})
+
+logging.basicConfig(
+    level=logging.INFO, format="%(asctime)s [%(levelname)s] %(message)s"
+)
+
+from dataset import SCIData, SCICols
+
+# %%
+
+# sci.clean_all().filter_vague_diagnoses().derive_readmission().omit_redundant().omit_vbg().omit_ae().save()
+X, y = (
+    SCIData.load("data/sci_processed.h5")
+    .augment_hsmr(onehot=True)
+    .impute_blood()
+    .raw_news()
+    .mandate_news()
+    .mandate_diagnoses()
+    .xy(outcome="DiedDuringStay")
+)
+
+# %%
+from sklearn.compose import make_column_selector, make_column_transformer
+from sklearn.preprocessing import OneHotEncoder, MinMaxScaler, FunctionTransformer
+from sklearn.pipeline import Pipeline
+
+numeric_pipeline = Pipeline(steps=[("scale", MinMaxScaler()),])
+
+categorical_pipeline = Pipeline(
+    steps=[("one-hot", OneHotEncoder(handle_unknown="ignore", sparse=False)),]
+)
+
+ct = make_column_transformer(
+    (numeric_pipeline, make_column_selector(dtype_include=np.number)),
+    (categorical_pipeline, make_column_selector(dtype_include=object)),
+)
+
+full_processor = Pipeline(steps=[("columns", ct)])
+
+
+# %%
+from sklearn.model_selection import train_test_split
+
+X_train, X_test, y_train, y_test = train_test_split(
+    full_processor.fit_transform(X), y, test_size=0.25
+)
+
+# %%
+from sklearn.metrics import (
+    accuracy_score,
+    precision_score,
+    recall_score,
+    roc_auc_score,
+    f1_score,
+    fbeta_score,
+    make_scorer,
+)
+
+f2_score = make_scorer(fbeta_score, beta=2)
+
+METRICS = {
+    "Accuracy": "accuracy",
[…]

### 2022-09-14 evaluation.py
.

@@ -13,6 +13,7 @@ from sklearn.metrics import (
     precision_recall_curve,
     ConfusionMatrixDisplay,
     RocCurveDisplay,
+    PrecisionRecallDisplay,
 )
 
 f2_score = make_scorer(fbeta_score, beta=2)
@@ -26,27 +27,45 @@ METRICS = {
     "F2 Score": f2_score,
 }
 
+ISOLATION_METRICS = {
+    "Accuracy": "accuracy",
+    "Precision": make_scorer(precision_score, pos_label=-1),
+    "Recall": make_scorer(recall_score, pos_label=-1),
+    "AUC": "roc_auc",
+    "F1 Score": make_scorer(fbeta_score, pos_label=-1, beta=1),
+    "F2 Score": make_scorer(fbeta_score, pos_label=-1, beta=2),
+}
 
-def evaluate_from_pred(y_true, y_pred, y_pred_proba, plot_title=None):
+
+def evaluate_from_pred(y_true, y_pred, y_pred_proba, plot_title=None, pos_label=1):
     display(
         pd.DataFrame(
             {
                 "Accuracy": accuracy_score(y_true, y_pred),
-                "Precision": precision_score(y_true, y_pred),
-                "Recall": recall_score(y_true, y_pred),
+                "Precision": precision_score(y_true, y_pred, pos_label=pos_label),
+                "Recall": recall_score(y_true, y_pred, pos_label=pos_label),
                 "AUC": roc_auc_score(y_true, y_pred_proba),
-                "F1 Score": f1_score(y_true, y_pred),
-                "F2 Score": fbeta_score(y_true, y_pred, beta=2),
+                "F1 Score": f1_score(y_true, y_pred, pos_label=pos_label),
+                "F2 Score": fbeta_score(y_true, y_pred, beta=2, pos_label=pos_label),
             },
             index=["Model"],
         )
     )
 
-    fig, ax = plt.subplots(1, 2, figsize=(11, 5))
+    fig, ax = plt.subplots(1, 3, figsize=(16, 5))
 
-    ax[1].grid(False)
-    ConfusionMatrixDisplay.from_predictions(y_true, y_pred, ax=ax[1])
-    RocCurveDisplay.from_predictions(y_true, y_pred_proba, ax=ax[0])
+    ax[2].grid(False)
+    RocCurveDisplay.from_predictions(
+        y_true, y_pred_proba, ax=ax[0],  # pos_label=pos_label
+    )
[…]

### 2022-09-14 ml_script_3.py

@@ -1,0 +1,71 @@
+from xgboost import XGBClassifier
+from imblearn.pipeline import Pipeline as ImbPipeline
+from imblearn.under_sampling import RandomUnderSampler
+from sklearn.model_selection import train_test_split
+from sklearn.model_selection import RandomizedSearchCV
+import pickle
+from evaluation import f2_score
+
+from dataset import SCIData, SCICols
+
+# SCIData.load('data/sci.h5').clean_all().filter_vague_diagnoses().derive_readmission().omit_vbg()
+sci = (
+    SCIData.load("data/sci_processed_2.h5")
+    .fix_readmissionband()
+    .derive_critical_event(within=2)
+)
+
+
+X, y = (
+    sci.omit_redundant()
+    .drop(["ReadmissionBand", "AgeBand"], axis=1)
+    .omit_ae()
+    .raw_news()
+    .mandate_news()
+    .mandate_blood()
+    .augment_hsmr()
+    .encode_ccs_onehot()
+    .xy(
+        outcome="CriticalEvent", dropna=False, ordinal_encoding=True
+    )  # Use ordinal encoding because of bug in XGB that prevents use of SHAP when pd.categorical is involved
+)
+
+X_train, X_test, y_train, y_test = train_test_split(
+    X, y, stratify=y, test_size=0.33, random_state=42
+)
+
+
+model = ImbPipeline(
+    steps=[
+        ("undersampling", RandomUnderSampler(sampling_strategy=0.1)),
+        (
+            "XGB",
+            XGBClassifier(
+                tree_method="approx",
+                enable_categorical=True,
+                scale_pos_weight=(y_train.shape[0] / (2 * y_train.sum())),
+            ),
+        ),
+    ]
+)
+
+param_grid = {
+    "XGB__max_depth": [3, 5, 6, 10, 15, 20],
+    "XGB__learning_rate": [0.01, 0.1, 0.2, 0.3],
+    "XGB__subsample": np.arange(0.5, 1.0, 0.1),
+    "XGB__colsample_bytree": np.arange(0.4, 1.0, 0.1),
+    "XGB__colsample_bylevel": np.arange(0.4, 1.0, 0.1),
+    "XGB__n_estimators": [100, 500, 1000],
+}
+
+clf = RandomizedSearchCV(
+    model, param_distributions=param_grid, scoring=f2_score, n_iter=25, verbose=1
+)
+
+clf.fit(X_train, y_train)
+
[…]

### 2022-09-15 evaluation.py
Updated baseline modelling.

@@ -1,4 +1,6 @@
 import pandas as pd
+import numpy as np
+
 from IPython.display import display
 import matplotlib.pyplot as plt
 
@@ -62,7 +64,7 @@ def evaluate_from_pred(y_true, y_pred, y_pred_proba, plot_title=None, pos_label=
         y_true, y_pred_proba, ax=ax[1], pos_label=pos_label
     )
 
-    if (-1) in y_true.values:
+    if (-1) in np.array(y_true):
         get = {1: True, -1: False}.get
         y_true, y_pred = (list(map(get, y_true)), list(map(get, y_pred)))
     ConfusionMatrixDisplay.from_predictions(y_true, y_pred, ax=ax[2])
@@ -76,8 +78,8 @@ def evaluate(model, X, y, plot_title=None):
         y_pred_proba = model.predict_proba(X)[:, 1]
     except AttributeError:
         try:
-            y_pred_proba = model.score_samples(X)
-        except AttributeError:
             y_pred_proba = model.decision_function(X)
+        except AttributeError:
+            y_pred_proba = model.score_samples(X)
 
     evaluate_from_pred(y, y_pred, y_pred_proba, plot_title)

### 2022-09-15 utils/evaluation.py
Producing percentages in confusion matrix

@@ -67,7 +67,7 @@ def evaluate_from_pred(y_true, y_pred, y_pred_proba, plot_title=None, pos_label=
     if (-1) in np.array(y_true):
         get = {1: True, -1: False}.get
         y_true, y_pred = (list(map(get, y_true)), list(map(get, y_pred)))
-    ConfusionMatrixDisplay.from_predictions(y_true, y_pred, ax=ax[2])
+    ConfusionMatrixDisplay.from_predictions(y_true, y_pred, ax=ax[2], normalize='true', values_format='.2%')
 
     plt.suptitle(plot_title)
 

### 2022-09-15 TeX/baseline/paper.tex
Updated baseline report

@@
 
-\title{Evaluation of NEWS on the Salford Dataset:\\
-{ACP Baseline Model Report}
+\title{ACP Baseline Model Report:\\
+{Evaluation of NEWS on the SRFT Dataset}
 % \thanks{Identify applicable funding agency here. If none, delete this.}
@@
 \section{Introduction}
-A common practice to gauge patients' physiological condition and acute clinical stability is using basic homeostatic measures like vital signs, haematologic findings, and neurological examinations.
-Various early warning systems have been proposed that use these measurements to assess whether patients face an imminent risk of deterioration or death \cite{Smith13}.
+In acute secondary care hospitals, standard practice is to gauge patients' physiological condition and acute clinical stability in a structured way using basic homeostatic measures.
+These include cardiorespiratory parameters, conscious level, fever and pain scales.
+Various early warning systems have been implemented that use these measurements to assess whether patients face an imminent risk of deterioration or death \cite{Smith13}.
 
-The National Early Warning Score (NEWS), developed in conjunction with the Royal College of Physicians, has been widely implemented across the NHS and by healthcare providers outside the UK \cite[pp.~13]{RCP17}.
-It is a points-based early warning system to identify clinical deterioration in acutely ill patients based on routinely recorded vital signs: pulse, breathing rate, blood pressure, blood oxygen level, level of consciousness, and temperature.
-The paper-based NEWS allocates points in a weighted manner based on these measurements.
-The sum of these points, which makes up the patient's aggregate NEW score, is associated with "triggers" - thresholds that correspond to recommended levels of clinical response and monitoring frequency.
+The National Early Warning Score (NEWS), developed in conjunction with the Royal College of Physicians, has been widely implemented across the National Health Service (NHS) and by healthcare providers outside the United Kingdom \cite[pp.~13]{RCP17}.
+It is a points-based early warning system to identify clinical deterioration in acutely ill patients based on routinely recorded vital signs.
+These include heart rate, respiratory rate, blood pressure, arterial oxygen saturation, level of consciousness, and temperature.
+The paper-recorded NEWS allocates points in a weighted manner based on these measurements.
+The sum of these points, which makes up the patient's aggregate NEWS, is associated with "triggers" - thresholds corresponing to recommended levels of clinical response and monitoring frequency.
 
-While initially designed to monitor for deterioration in secondary-care, the NEWS has been widely tested and validated across healthcare settings.
-In its latest revision of the standard, the RCP recommended that the NEWS also be used in emergency departments to "aid the initial assessment of patiens, ongoing monitoring and patient triage decisions" \cite[pp.~18]{RCP17}.
+While initially designed to monitor for deterioration in secondary care, the NEWS has been widely tested and validated across healthcare settings.
+In its latest revision of the standard, the RCP recommended that the NEWS also be used in emergency departments to "aid the initial assessment of patients, ongoing monitoring and patient triage decisions" \cite[pp.~18]{RCP17}.
+Across studies, the NEWS has been found to perform well in discriminating patients at risk of death, critical care admission, or cardiac arrest.
 
-Across studies, the NEWS has been found to perform well in discriminating ward patients at risk of death, critical care admission, or cardiac arrest.
-The present report introduces the novel Salford dataset, which contains records of adult patients with acute or medical presentation at the Salford Royal NHS Foundation Trust from 2014-2022.
-We then use this dataset to assess the performance of NEWS in predicting adverse patient outcomes, such as those mentioned above, in the dataset population.
+The present report assesses the performance of NEWS in predicting the above adverse patient outcomes using the electronic patient record (EPR) dataset of adult patients with acute medical presentation to the Salford Royal NHS Foundation Trust from 2014-2022.
+The aim was to specifically determine the performance of NEWS in the context of the first recorded set of parameters at the point of admission.
 
-Further, we investigate the performance of NEWS on post-discharge mortality, lengthy hospital stay, and readmission within certain timeframes.
-Anticipating these outcomes is in keeping with the main purpose of an EWS: to effectively recognise patients whose condition is deteriorating but can be improved by timely medical intervention  \cite[pp.~4]{Smith13}.
+Further, we investigate the performance of NEWS on additional potential functions.
+These are post-discharge mortality, length of hospital stay, and readmission within certain timeframes.
+Anticipating these outcomes is both in keeping with the main purpose of an EWS (to effectively recognise patients whose condition is deteriorating but can be improved by timely medical intervention \cite[pp.~4]{Smith13}) whilst also providing a tool to aid in very early clinical decision making at an organisational and patient level with regards to discharge planning and bed occupancy from the time of arrival.
+All of these functions would indicate value of using NEWS to facilitate triaging patients to ambulatory care, acute medical unit (AMU), in-patient wards or areas where continuous monitoring is available.
 
@@
 \subsection{Data Collection and Contents}
-The Salford dataset is collated from EPR records collected in real time from admissions to acute beds between 1st April 2014 and 31st March 2022.
-Further, the dataset includes administrative data produced in retrospect for financial (tarrif) purposes.
+Salford Royal Hospital is a digitally mature, "paper light" NHS secondary care hospital with $>100,000$ emergency department attendances and approximately $40,000$ unplanned admissions each year.
+The Salford EPR captures clinical episode data in real time from arrival at the Emergency Department to discharge.
+Selected data are exported pseudonymously to an internal data warehouse which is used to drive local quality improvement and service development projects.
+These data are supplemented by downstream administrative and outcome data including ICD-10, OPCS4, and HRG codes alongside re-admission and community mortality events.
 
-Our dataset includes records of patients who received ambulatory emergency care (AEC) or same-day emergency care (SDEC) and were discharged before midnight on the day of admission \cite{NHS18}.
-AEC, alongside the medical admissions unit (EAU)  are the two common entry points for adult patients with general medical emergencies.
-Exceptions present in our records include certain booked admissions and patients directly transferred and admitted to medical or critical care wards.
+We analysed such records for the period 1st April 2014 to 31st March 2022, and selected episodes in patients aged $\geq 18$ years who were admitted to the Acute Medical Unit, at Salford known as the Emergency Admissions Unit (EAU).
+Our dataset includes records of patients who received ambulatory emergency care (AEC) or same-day emergency care (SDEC).
+Paediatric patients, planned admissions, day case reviews and acute medical patients admitted directly from ED to critical care were excluded.
 
-As a routine part of admission, the responsible staff member (nurse or support worker) records the patient's vital signs withing $\sim 30$ minutes of arrival.
+As a routine part of EAU admission, the responsible staff member (nurse or support worker) records the patients' vital signs within a target of 30 minutes of arrival.
 The vital signs that make up the NEWS are measured simultaneously, in a standardised manner, using Dinamap monitors.
-The readings are manually transcribed into EPR and onto the patient's paper chart.
-Specifically, this data includes:
+The readings are manually transcribed into EPR.
+Specifically, these data include:
+
 \begin{itemize}
@@
 \end{itemize}
-Further, we record the patient's level of consciousness (AVPU), presence of pain, nausea, or vomiting, whether the patient was receiving oxygen at the time of SpO2 measurement and, if applicable, the oxygen flow-rate and mode of delivery.
+Other parameters are measured using manual observation and direct questions.
+These are the patient's level of consciousness (AVPU), presence of pain, nausea, or vomiting, whether the patient was receiving oxygen at the time of SpO2 measurement and, if applicable, the oxygen flow-rate and mode of delivery.
 Once these parameters are inputted into EPR, the NEWS score and component sub-scores are automatically computed for the patient.
-Blood test results (including VBG), where those were peformed, are automatically recorded in the laboratory information management system LIMS) and subsequently imported into EPR.
 
-Other patient information recorded on admission includes: Patient identifier data such as their unique patient number, age, sex, and address.
-Admission method (e.g., emergency A\&E, emergency GP referral, etc), their presenting complaint, arrival time, and the main A\&E diagnosis (following initial assessment).
+Independent of this, blood test results (including venous blood gas [VBG]), where those were performed, are automatically recorded in the laboratory information management system (LIMS) and imported into EPR in real time.
+
+Other patient information available on admission includes: Patient identifier data such as their unique patient number, age, sex, and address.
+Admission method (e.g., emergency ED, emergency GP referral, etc), their presenting complaint, arrival time, ED location, and the main ED diagnosis (following initial assessment).
+For patients with prior hospital visits, major co-morbidities and previous admission events are available from the point of admission.
 
@@
 Procedures and services are also recorded in detail in EPR, and are similarly coded in retrospect using the OPCS-4 standard.
+Real time EPR entries relating to each of these are captured in individual structured documents.
+For the purposes of this study, the post-discharge coded entries were used as surrogates for real-time, peri-admission clinical event and intervention recording.
 
-\subsection{Outcomes} The outcomes we track are death during hospitalisation, death within 30 days after discharge, unanticipated admission to critical care, or unanticipated readmission to the same hospital.
-To assess the NEWS, we check for the occurrence of death or critical care admission within 24 hours, 48 hours, or at any point after the patient's arrival.
-For readmissions, we consider thresholds of 48 hours, 7 days, or 30 days after discharge.
+\subsection{Outcomes} The outcomes tracked were death during hospitalisation, death within 30 days after discharge, admission to critical care from the ward, or unplanned readmission to the same hospital.
+To assess the NEWS, we tracked death or critical care admission within 24 hours, 48 hours, or at any point after admission.
+For readmissions, we considered 48 hours, 7 days, or 30 days after discharge.
 
-Where it occurrs, patient death during their hospitalisation is identified from EPR.
-Mortality after discharge (within 30 days) is primarily determined from the regional primary care databases (Salford Integrated Records \& Greater Manchester Care Record).
+Where it occurrs, patient death during hospitalisation is identified from EPR.
+Mortality after discharge (within 30 days) is determined from the regional primary care databases (Salford Integrated Records \& Greater Manchester Care Record).
 Critical care admission is identified from patients being admitted into the hospital's critical care unit (CCU) or the high-dependency medical unit (HH1M).
 
-\subsection{Data Preparation} We perform all data manipulation in Python using the Numpy and Pandas libraries.
+\subsection{Data Preparation} We performed all data manipulation in Python using the Numpy and Pandas libraries.
 As they are the only parameters that are transcribed manually into EPR, we check the patients' vital signs for spurious values against hard ranges (e.g., $0-100\%$ for SpO2) as well as soft thresholds that we base on the range of physiologically possible values as determined by professional clinical opinion.
 For parameters that have a recorded NEWS sub-score but their raw value is missing or marked as spurious, we infer the correct value as the midpoint of the relevant NEWS range.
-Where a parameter has a valid value but the NEWS sub-score is missing, we compute it in accordance with the specificaton \cite[pp.~28-30]{RCP17}.
+Where a parameter has a valid value but the NEWS sub-score is missing, we compute it in accordance with the NEWS specificaton.
 We use the same specification to compute the final NEWS score for any patients where this is not recorded in the original data.
-Full details of how each parameter is pre-processed are available in appendix A.
+Full details of how each parameter is pre-processed are available in appendix \ref{appendix:preprocessing}.
 
-\subsection{Data Analysis} We assess the ability of NEWS to discriminate between patients who did and did not suffer an adverse outcome using AUROC (area under the ROC curve).
-We compute this using the scikit-Learn library.
+\subsection{Data Analysis}
+The original study on the NEWS' discriminative ability considered death within 24 hours of a given set of observations.
+We defined a "Critical Event" as the occurrence of either critical care admission or death during the hospital spell.
+Since the vital signs and NEWS in our dataset are measured on admission, we investigate occurrence within a certain timeframe after admission, such as within $24$ or $48$ hours.
+
+We assessed the ability of NEWS to discriminate between patients who did and did not suffer an adverse outcome using AUROC (area under the ROC curve).
 The minimum possible AUROC value is $0.5$ and corresponds to a completely random relationship between the score in question and the tracked outcomes.
@@
 
-We investigate the discriminative performance of both the raw NEWS score and the pre-defined "trigger" thresholds for low, medium, and high clinical risk (NEWS of $0-4$, $5-6$, and $\geq 7$ respectively) \cite[pp.~30]{RCP17}.
+We investigated the discriminative performance of both the raw NEWS score and the pre-defined "trigger" thresholds for low, medium, and high clinical risk (NEWS of $0-4$, $5-6$, and $\geq 7$ respectively) \cite[pp.~30]{RCP17}.
 
-We measure correlation between ordinal variables (NEWS and LOS) using Spearman's rank correlation.
-We compare the means of independent samples of normally distributed variables using Welch's unequal variances t-test, and we use the Mann-Whitney U test for non-normally distributed variables.
-We present characteristics of variables as $mean \pm std$.
+We measured correlation between ordinal variables (NEWS and LOS) using Spearman's rank correlation.
+We compared the means of independent samples of normally distributed variables using Welch's unequal variances t-test, and we used the Mann-Whitney U test for non-normally distributed variables.
+We present results as $mean \pm std$.
 
-\section{Results}
-Table \ref{tab:result_summary} summarises our findings on the discriminative ability of NEWS for various outcomes.
-\begin{table}[!t]
-    \renewcommand{\arraystretch}{1.3}
+\begin{table}[tbp]
+    \renewcommand{\arraystretch}{1.2}
     \centering
-    \caption{Summary of Findings}
-    \label{tab:result_summary}
-    \begin{tabular}{lllr}
-        \toprule
-        Outcome        & Time         & NEWS $\geq 7$ & ROAUC (95\% CI)              \\
-        \midrule
-        Critical Event & 24h          & $51.13\%$     & $\bm{0.914}$ ($0.896-0.932$) \\
-                       & 48h          & $41.24\%$     & $0.890$ ($0.877-0.903$)      \\
-                       & Any          & $16.12\%$     & $0.769$ ($0.763-0.775$)      \\
-        \hline
-        Mortality      & 24h          & $62.01\%$     & $\bm{0.952}$ ($0.936-0.967$) \\
-                       & 48h          & $51.15\%$     & $0.938$ ($0.926-0.95$)       \\
-                       & Any          & $18.67\%$     & $0.770$ ($0.762-0.777$)      \\
-        \hline
-        Critical Care  & 24h          & $26.92\%$     & $\bm{0.831}$ ($0.787-0.874$) \\
-                       & 48h          & $24.64\%$     & $0.8$ ($0.772-0.829$)        \\
-                       & Any          & $15.1\%$      & $0.767$ ($0.758-0.775$)      \\
-        \hline
-        Long LOS       & $\geq$24h    & $2.50\%$      & $0.602$ ($0.598-0.605$)      \\
-                       & $\geq$48h    & $2.72\%$      & $\bm{0.607}$ ($0.604-0.61$)  \\
-                       & $\geq$7 days & $3.25\%$      & $0.598$ ($0.594-0.602$)      \\
-        \hline
-        30-Day Mor/ty  & 24h          & $3.68\%$      & $\bm{0.655}$ ($0.628-0.682$) \\
-                       & 48h          & $3.78\%$      & $0.655$ ($0.634-0.677$)      \\
-                       & Any          & $4.28\%$      & $0.629$ ($0.62-0.638$)       \\
-        \hline
-        Readmission    & 2 days       & $1.67\%$      & $0.495$ ($0.478-0.511$)      \\
-                       & 7 days       & $1.06\%$      & $0.503$ ($0.482-0.501$)      \\
-                       & 30 days      & $1.31\%$      & $0.523$ ($0.518-0.528$)      \\
-        \bottomrule
-    \end{tabular}
-\end{table}
-\subsection{Population Characteristics} Records of $170833$ admissions were collated, of which $124162$ records, corresponding to $63450$ unique patients, had a NEW score recorded on-admission or computed by us during pre-processing.
-NEW scores ranges from $0-18$ with a median score of $1.0$ (IQR=$2.0$).
-Key characteristics of the dataset population are given in Table \ref{tab:population}.
-
-\begin{table}[!t]
-    \renewcommand{\arraystretch}{1.3}
-    \centering
-    \caption{Salford dataset key characteristics}
+    \caption{Key dataset characteristics}
     \label{tab:population}
@@
 
-\subsection{Critical Event Occurrence} The original study on the NEWS' discriminative ability considered death within 24 hours of a given set of observations.
-We define a "Critical Events" as the occurrence of either critical care admission or death during the hospital spell.
-Since the vital signs and NEWS in our dataset are measured on admission, we investigate the subsets of hospital spells with total LOS within certain thresholds, such as 24 or 48 hours.
+\section{Results}
+Table \ref{tab:result_summary} summarises our findings on the discriminative ability of NEWS for various outcomes.
+\begin{table}[!t]
+    \renewcommand{\arraystretch}{1.3}
+    \centering
+    \caption{Summary of Findings}
+    \label{tab:result_summary}
+    \begin{tabular}{lllr}
+        \toprule
+        Outcome        & Time         & NEWS $\geq 7$ & ROAUC (95\% CI)              \\
+        \midrule
+        Critical Event & 24h          & $33.0\%$      & $\bm{0.817}$ ($0.800-0.833$) \\
+                       & 48h          & $27.23\%$     & $0.807$ ($0.796-0.819$)      \\
+                       & Any          & $16.12\%$     & $0.769$ ($0.763-0.775$)      \\
+        \hline
+        Mortality      & 24h          & $62.01\%$     & $\bm{0.930}$ ($0.911-0.948$) \\
+                       & 48h          & $51.15\%$     & $0.912$ ($0.899-0.926$)      \\
+                       & Any          & $18.67\%$     & $0.770$ ($0.762-0.777$)      \\
+        \hline
+        Critical Care  & 24h          & $18.26\%$     & $\bm{0.759}$ ($0.736-0.781$) \\
+                       & 48h          & $16.19\%$     & $0.756$ ($0.741-0.771$)      \\
+                       & Any          & $15.1\%$      & $0.767$ ($0.758-0.775$)      \\
+        \hline
+        Long LOS       & $\geq$24h    & $2.50\%$      & $0.602$ ($0.598-0.605$)      \\
+                       & $\geq$48h    & $2.72\%$      & $\bm{0.607}$ ($0.604-0.61$)  \\
+                       & $\geq$7 days & $3.25\%$      & $0.598$ ($0.594-0.602$)      \\
+        \hline
+        30-Day Mor/ty  & 30 days      & $4.28\%$      & $0.629$ ($0.620-0.638$)      \\
+        \hline
+        Readmission    & 2 days       & $1.67\%$      & $0.495$ ($0.478-0.511$)      \\
+                       & 7 days       & $1.06\%$      & $0.503$ ($0.482-0.501$)      \\
+                       & 30 days      & $1.31\%$      & $0.523$ ($0.518-0.528$)      \\
+        \bottomrule
+    \end{tabular}
+\end{table}
+\subsection{Population Characteristics}
 
-A total of $8826$ with recorded NEWS experienced a critical event, comprising $4371$ critical care admissions and $5308$ deaths.
-The AUROCs ($95\%$ CI) of NEWS for critical events in records with 24h LOS, 48h LOS, or any LOS were respectively $0.914$ ($0.896-0.932$), $0.890$ ($0.877-0.903$), and $0.769$ ($0.763-0.775$).
+There were $170,833$ unique admissions during the study period, of which $124,162$ records, corresponding to $63,450$ unique patients, had a NEWS score recorded on admission or computed by us during pre-processing.
+NEWS scores ranged from $0-18$ with a median score of $1.0$ (IQR=$2.0$).
+Key population characteristics are given in Table \ref{tab:population}.
+
+\begin{figure*}[tbp]
+    \centering
+    \includegraphics[width=\textwidth]{img/critical_roc_multi.png}
+    \caption{ROC curves of NEWS for discriminating critical event occurrence within 24h of admission (left), 48h (middle), or at any time (right).}
+    \label{fig:criticalevent_roc}
+\end{figure*}
+
+\subsection{Critical Event Occurrence} A total of $8,826$ patients with recorded NEWS experienced a critical event, comprising $4371$ critical care admissions and $5,308$ deaths.
+The AUROCs ($95\%$ CI) of NEWS for critical events in records within 24h of admission, within 48h, or at any point were respectively $0.914$ ($0.896-0.932$), $0.890$ ($0.877-0.903$), and $0.769$ ($0.763-0.775$).
 The ROC curves are given in figure \ref{fig:criticalevent_roc}.
@@
 
-\begin{figure*}[htbp]
-    \centering
-    \includegraphics[width=\textwidth]{img/critical_roc_multi.png}
-    \caption{ROC curves of NEWS for discriminating critical event occurrence in stays with LOS under 24h (left), under 48h (middle), or any LOS (right).}
-    \label{fig:criticalevent_roc}
-\end{figure*}
-
 Critical event occurrence significantly correlated with the NEWS at the same LOS thresholds: chi-square with NEWS $\geq 7$: $p < 0.0001$ in all cases.
-In total, $1423/8826$ ($16\%$) of critical event occurrences had NEWS $\geq 7$.
-Appendix \ref{appendix:criticalevent_breakdown} examines the occurrence of mortality and critical care admission separately.
+In total, $1,423/8,826$ ($16\%$) of critical event occurrences had NEWS $\geq 7$.
+Appendix \ref{appendix:criticalevent_breakdown} examines mortality and critical care admission.
 
@@
 The mean total LOS was $16.32 \pm 24.25$ for patients who experienced critical events and $6.07 \pm 13.62$ for those who did not (Mann-Whitney U test: $p < 0.0001$).
-Table \ref{tab:criticalevent_stats} gives the same information across the tested LOS thresholds.
+Table \ref{tab:criticalevent_stats} gives the same information across the tested thresholds.
+We fail to reject at $\alpha=0.05$ when testing the mean age under critical event occurrence within 24 hours (Welch's unequal variances t-test: $p = 0.076$).
 
-\begin{table}[!t]
+\begin{table}[tbp]
     \renewcommand{\arraystretch}{1.3}
@@
         \toprule
-        Variable & LOS & Mean (Event)      & Mean ($\lnot$Event) & Test  & $p$   \\
+        Variable & Time & Mean (Event)      & Mean ($\lnot$Event) & Test  & $p$      \\
         \midrule
-        Age      & 24h & $71.4 \pm 18.92$  & $57.62 \pm 20.92$   & Welch & $0.0$ \\
-                 & 48h & $70.54 \pm 19.56$ & $58.71 \pm 21.09$   &       & $0.0$ \\
+        Age      & 24h  & $66.12 \pm 19.0$  & $65.0 \pm 20.47$    & Welch & $0.076$  \\
+                 & 48h  & $66.5 \pm 18.81$  & $64.98 \pm 20.48$   &       & $0.0003$ \\
                  & Any  & $71.72 \pm 16.85$ & $64.5 \pm 20.62$    &       & $0.0$    \\
-        NEWS     & 24h & $6.82 \pm 4.38$   & $0.75 \pm 1.07$     & Mann  & $0.0$ \\
-                 & 48h & $5.91 \pm 4.18$   & $0.78 \pm 1.1$      &       & $0.0$ \\
+        NEWS     & 24h  & $5.11 \pm 4.08$   & $1.19 \pm 1.64$     & Mann  & $0.0$    \\
+                 & 48h  & $4.61 \pm 3.76$   & $1.16 \pm 1.59$     &       & $0.0$    \\
                  & Any  & $3.51 \pm 3.1$    & $1.04 \pm 1.4$      &       & $0.0$    \\
@@
 
-\subsection{Length of Stay} The total LOS for the patient was significantly correlated with NEWS (Spearman Rank Correlation: $p < 0.0001$).
+\subsection{Length of Stay} The total LOS for patients was significantly correlated with NEWS (Spearman Rank Correlation: $p < 0.0001$).
 Median length of stay more than doubled for NEWS $\geq 7$ compared to $<7$ ($5.90$ and $2.21$, respectively).
 
-We further investigate whether NEWS can discriminate stays where the LOS will be greater than a certain amount of time; for example, LOS over one day, meaning the patient needs to be monitored overnight.
+We further investigated whether NEWS can discriminate stays where the LOS will be greater than a certain amount of time; for example, LOS over one day, meaning the patient needs to be monitored overnight.
 The AUROCs for LOS $\geq$24h, $\geq$48h, and $\geq$7 days were $0.602$ ($0.598-0.605$), $0.607$ ($0.604-0.61$), and $0.598$ ($0.594-0.602$) respectively.
@@
 \subsection{Post-Discharge Mortality}
-The proportion of 30-day mortality in the dataset records was $3.29\%$ ($n=4086$) and corresponds to $6.44\%$ of patients.
-The AUROCs ($95\%$ CI) of NEWS for 30-day mortality in records with 24h LOS, 48h LOS, or any LOS were respectively $0.655$ ($0.628-0.682$), $0.656$ ($0.634-0.677$), and $0.629$ ($0.62-0.638$).
-Post-discharge mortality significantly correlated with NEWS at the same LOS thresholds: chi-square with NEWS $\geq 7$: $p < 0.0001$ in all cases.
+The proportion of 30-day mortality in the dataset records was $3.29\%$ ($n=4,086$) and corresponds to $6.44\%$ of patients.
+The AUROC ($95\%$ CI) of NEWS for 30-day mortality was $0.629$ ($0.62-0.638$).
+Post-discharge mortality significantly correlated with NEWS: chi-square with NEWS $\geq 7$: $p < 0.0001$.
 In total, $175/4090$ ($4.28\%$) of occurrences had NEWS $\geq 7$.
@@
 The mean total LOS was $13.49 \pm 18.99$ for patients who died after discharge and $6.57 \pm 14.65$ for survivors (Mann-Whitney U test: $p < 0.0001$).
-Table \ref{tab:criticalevent_stats} gives the same information across the tested LOS thresholds.
 
-\subsection{Readmission} The proportion of 30-day readmissions in the dataset was $11.26\%$ (n=$14003$).
+\subsection{Readmission} The proportion of 30-day readmissions in the dataset was $11.26\%$ (n=$14,003$).
 The AUROCs ($95\%$ CI) of NEWS for readmission within 2 days, 7 days, and 30 days were respectively $0.495$ ($0.478-0.511$), $0.491$ ($0.482-0.501$), and $0.523$ ($0.518-0.528$), indicating the NEWS is no better at predicting readmission than random chance.
@@
 \section{Discussion}
-We have investigated applying the NEWS to a large vital signs dataset from a single acute care trust, mainly consisting of emergency admissions and not considering elective procedures or surgical admissions.
+We have investigated applying the NEWS to a large EPR dataset from a single acute care trust, consisting of admissions to the hospital's Acute Medical Unit.
 
-The original study by the NEWSDIG to evaluate the then-newly-proposed NEWS tracked death within 24 hours of readings and reported an AUROC score of $0.89$.
-A subsequent study in emergency departments in the Netherlands gave an AUROC score of $0.768$ for 30-day mortality based on the admission NEWS.
+The original study by the NEWSDIG to evaluate the then-newly-proposed NEWS tracked death within 24 hours of readings and reported an AUROC score of $0.89$ \cite{Smith13}.
+A subsequent study in emergency departments in the Netherlands gave an AUROC score of $0.768$ for 30-day mortality based on the admission NEWS \cite{Alam15}.
+Our findings regarding the NEWS' discriminative performance are consistent with the previous studies.
+The NEWS performed best in detecting critical events (in-hospital mortality or admission to critical care).
+It showed potential in identifying patients at risk of extended hospital stays or post-discharge mortality.
 
@@
 
-Our findings regarding the NEWS' discriminative performance are consistent with the previous studies.
-The NEWS performed best in detecting critical events (in-hospital mortality or admission to critical care).
-It showed potential in identifying patients at risk of extended hospital stays or post-discharge mortality.
-
-The NEWS was not effective in detecting which patients in our data would be readmitted to the same hospital in the future within any amount of time.
+The NEWS was not effective in detecting which patients in our data would be readmitted to the same hospital in the future.
 Prior work has found that clinical data indicating the severity of the patient's condition, including constructed scores, is useful in predicting 30-day readmission \cite[pp.~5]{Mahmoudim20}.
@@
 In their comparison of EWS, \cite{Smith13} found that existing EWS systems that considered age showed poor discriminative ability in critical care admission compared to death.
-The authors noted that the observed performance of an EWS on predicting mortality might not line up with the "true" desired property of an EWS: to effectively recognise patients whose condition is deteriorating but can be improved by timely medical intervention rather than those who will likely not survive \cite[pp.~4]{Smith13}.
-In our view, this makes a strong case for ensuring good discrimination of patients in terms of other adverse outcomes, such as critical care admission, post-discharge mortality, and lengthy hospital stays, to inform and assist clinical decision-making.
+The authors noted that the observed performance of an EWS on predicting mortality might not line up with the main purpose of an EWS (to recognise patients whose condition is deteriorating but can be improved by timely medical intervention).
+In our view, this objective makes a strong case for ensuring good discrimination of adverse outcomes besides in-hospital mortality, such as critical care admission, post-discharge mortality, and lengthy hospital stays.
 
@@
 Some measured outcomes and population parameters are inherently variable by hospital.
-For example, critical care admission may vary between hospitals because each hospital's criteria for admission are different.
+For example, critical care admission may vary between hospitals due to different admission criteria.
 Making definitive statements about the utility of the NEWS requires further validation in different patient populations and clinical settings.
 
-Patient data were collected primarily by staff under everyday conditions.
+Patient data were collected by staff under everyday conditions.
 While this matches the real-life use of the NEWS, operational pressures may hinder the timeliness of data entry and the dataset's reliability.
 Further, we did not track whether patients had a DNAR order, which can lead to patient death in cases where it otherwise may not have occurred.
-
-A limitation of the dataset was that it only records NEWS measurements taken on admission.
-To compare with prior studies, we considered adverse event occurrence in hospital stays that lasted under 24 or 48 hours as a proxy for testing NEWS values measured shortly before these events.
-This practice may induce a bias in our results compared to evaluating the NEWS on data with repeated sets of observations for each patient throughout their hospital stay.
+Finally, a limitation of the dataset was that it only records NEWS measurements taken on admission.
+Thus, to compare with prior studies, we tracked adverse event occurrence within 24 or 48 hours of admission.
+A more complete evaluation should include repeated sets of observations for each patient throughout their hospital stay.
 
@@
 \subsection{Mortality \& Critical Care Admission}
-The proportion of in-hospital mortality in the dataset records was $4.28\%$ ($n=5328$) and corresponds to $8.386\%$ of patients.
-The AUROCs ($95\%$ CI) of NEWS for mortality in records with 24h LOS, 48h LOS, or any LOS were respectively $0.952$ ($0.936-0.967$), $0.938$ ($0.926-0.95$), and $0.770$ ($0.762-0.777$).
+The proportion of in-hospital mortality in the dataset records was $4.28\%$ ($n=5,328$) and corresponds to $8.386\%$ of patients.
+The AUROCs ($95\%$ CI) of NEWS for mortality within 24h of admission, within 48h, or at any point were respectively $0.93$ ($0.911-0.948$), $0.912$ ($0.899-0.926$), and $0.770$ ($0.762-0.777$).
 Mortality significantly correlated with NEWS at the same LOS thresholds: chi-square with NEWS $\geq 7$: $p < 0.0001$ in all cases.
-In total, $955/5328$ ($17.9\%$) of occurrences had NEWS $\geq 7$.
+In total, $955/5,328$ ($17.9\%$) of occurrences had NEWS $\geq 7$.
 
-The proportion of critical care admission in the dataset records was $3.53\%$ ($n=4384$) and corresponds to $6.90\%$ of patients.
-The AUROCs ($95\%$ CI) of NEWS for critical care admission in records with 24h LOS, 48h LOS, or any LOS were respectively $0.831$ ($0.787-0.874$), $0.8$ ($0.772-0.829$), and $0.767$ ($0.758-0.775$).
+The proportion of critical care admission in the dataset records was $3.53\%$ ($n=4,384$) and corresponds to $6.90\%$ of patients.
+The AUROCs ($95\%$ CI) of NEWS for critical care admission within 24h of admission, within 48h, or at any point were respectively $0.759$ ($0.736-0.781$), $0.756$ ($0.741-0.71$), and $0.767$ ($0.758-0.775$).
 Admission significantly correlated with NEWS at the same LOS thresholds: chi-square with NEWS $\geq 7$: $p < 0.0001$ in all cases.
-In total, $662/4384$ ($15.1\%$) of occurrences had NEWS $\geq 7$.
+In total, $662/4,384$ ($15.1\%$) of occurrences had NEWS $\geq 7$.
 
@@
     \centering
-    \includegraphics[width=\textwidth]{img/30daymortality_roc.png}
-    \caption{ROC curves of NEWS for discriminating post-discharge mortality in stays with LOS under 24h (left), under 48h (middle), or any LOS (right).}
+    \includegraphics[width=0.3\textwidth]{img/30daymortality_roc.png}
+    \caption{ROC curve of NEWS for discriminating post-discharge mortality.}
     \label{fig:30day_roc}

### 2022-09-15 utils/evaluation.py

@@ -39,7 +39,7 @@ ISOLATION_METRICS = {
 }
 
 
-def evaluate_from_pred(y_true, y_pred, y_pred_proba, plot_title=None, pos_label=1):
+def evaluate_from_pred(y_true, y_pred, y_pred_proba, plot_title=None, pos_label=1, save=None):
     display(
         pd.DataFrame(
             {
@@ -71,8 +71,11 @@ def evaluate_from_pred(y_true, y_pred, y_pred_proba, plot_title=None, pos_label=
 
     plt.suptitle(plot_title)
 
+    if save:
+        plt.savefig(save, bbox_inches="tight")
 
-def evaluate(model, X, y, plot_title=None):
+
+def evaluate(model, X, y, plot_title=None, save=None):
     y_pred = model.predict(X)
     try:
         y_pred_proba = model.predict_proba(X)[:, 1]
@@ -82,4 +85,4 @@ def evaluate(model, X, y, plot_title=None):
         except AttributeError:
             y_pred_proba = model.score_samples(X)
 
-    evaluate_from_pred(y, y_pred, y_pred_proba, plot_title)
+    evaluate_from_pred(y, y_pred, y_pred_proba, plot_title=plot_title, save=save)

### 2022-09-16 dataset.py
.

@@ -830,12 +830,12 @@ class SCIData(pd.DataFrame):
 
         return SCIData(r)
 
-    def encode_ccs_onehot(self, drop_old=True):
+    def encode_ccs_onehot(self, drop_old=True, prefix="HSMR"):
         """Given DataFrame augmented with CCS groupings (CCS, SHMI, or HSMR), one-hot encodes them"""
 
-        r = self.encode_onehot(SCICols.diagnoses, "CCS", drop_old)
+        r = self.encode_onehot(SCICols.diagnoses, prefix, drop_old)
         r = r.rename(
-            columns={_: str(_) for _ in r.columns if str(_).startswith("CCS_")}
+            columns={_: str(_[:-2]) for _ in r.columns if str(_).startswith(prefix)}
         )
 
         return SCIData(r)
@@ -860,8 +860,10 @@ class SCIData(pd.DataFrame):
             self.dropna(how="any", subset=set(cols).intersection(self.columns),)
         )
 
-    def mandate_diagnoses(self):
-        return self.mandate([_ for _ in self.columns if _.startswith('CCS_')] + SCICols.diagnoses)
+    def mandate_diagnoses(self, prefix="HSMR"):
+        return self.mandate(
+            [_ for _ in self.columns if _.startswith(prefix)] + SCICols.diagnoses
+        )
 
     def mandate_news(self):
         return self.mandate(SCICols.news_data)
@@ -905,21 +907,24 @@ class SCIData(pd.DataFrame):
         return SCIData(r)
 
     def describe_categories(self):
-        categorical_cols = [_ for _ in self.columns if self.get_column_types()[_] == 'c']
-
-        categorical_cols_idx = [
-            self.columns.get_loc(_)
-            for _ in categorical_cols
+        categorical_cols = [
+            _ for _ in self.columns if self.get_column_types()[_] == "c"
         ]
+
+        categorical_cols_idx = [self.columns.get_loc(_) for _ in categorical_cols]
         categorical_dims = [self[_].unique().shape[0] for _ in categorical_cols]
 
         return categorical_cols_idx, categorical_dims
 
     def get_column_types(self):
         return {
-            **{_:'q' for _ in self.columns},
-            **{_:'i' for _ in self.columns if _.startswith('CCS_')},
-            **SCICols.xgb_types
[…]

### 2022-09-16 utils/evaluation.py

@@ -1,9 +1,12 @@
+from collections import defaultdict
 import pandas as pd
 import numpy as np
 
 from IPython.display import display
 import matplotlib.pyplot as plt
 
+from sklearn.base import clone as clone_estimator
+from sklearn.model_selection import train_test_split, cross_validate
 from sklearn.metrics import (
     accuracy_score,
     precision_score,
@@ -18,6 +21,12 @@ from sklearn.metrics import (
     PrecisionRecallDisplay,
 )
 
+
+from imblearn.pipeline import Pipeline as ImbPipeline
+from imblearn.under_sampling import RandomUnderSampler, TomekLinks
+from imblearn.over_sampling import SMOTE, SMOTENC
+
+
 f2_score = make_scorer(fbeta_score, beta=2)
 
 METRICS = {
@@ -29,17 +38,160 @@ METRICS = {
     "F2 Score": f2_score,
 }
 
-ISOLATION_METRICS = {
-    "Accuracy": "accuracy",
-    "Precision": make_scorer(precision_score, pos_label=-1),
-    "Recall": make_scorer(recall_score, pos_label=-1),
-    "AUC": "roc_auc",
-    "F1 Score": make_scorer(fbeta_score, pos_label=-1, beta=1),
-    "F2 Score": make_scorer(fbeta_score, pos_label=-1, beta=2),
-}
+
+def spotCheckDatasets(
+    models,
+    datasets,
+    cv=3,
+    set_params={},
+    order=["dataset", "model"],
+    fit_params={},
+    scoring=METRICS,
+):
+    """ Run stratified k-fold cross-validation on the given models over the given dataset variants
+        :param models: Single estimator or dict of the format {'Model Name': model}
+        :param datasets: Single tuple (X, y) or dict of the format {'Dataset name': (X, y)}
+        :param set_params: Hook to set model parameters based on each dataset. Callable of type function(X, y) -> Dict
+    """
+
+    models = {"Model": models} if type(models) != dict else models
+    datasets = {"Data": datasets} if type(datasets) != dict else datasets
+
+    param_callables = defaultdict(lambda: lambda X, y: dict())
+    param_callables.update(set_params)
+
+    fit_param_dict = defaultdict(lambda: dict())
+    fit_param_dict.update(fit_params)
+
+    tests = [
+        (
+            datakey,
+            (X, y),
+            modelkey,
+            clone_estimator(model).set_params(**(param_callables[modelkey](X, y))),
+            fit_param_dict[modelkey],
+        )
[…]

### 2022-09-16 dataset.py
.

@@ -113,7 +113,7 @@ class SCIData(pd.DataFrame):
         if onehot:
             r = r.encode_ccs_onehot()
 
-        return SCIData(r)
+        return r
 
     def _regroup_ccs(self, df, col, onehot=False):
         """Joins SCI and a given grouping table for CCS. Matches ICD-10 diagnoses with CCS if this has not already been done
@@ -131,7 +131,7 @@ class SCIData(pd.DataFrame):
         if onehot:
             r = r.encode_ccs_onehot()
 
-        return SCIData(r)
+        return r
 
     def augment_shmi(self, infile="data/ccs.h5", shmi=None, onehot=False):
         """Joins SCI and the SHMI matching table.

### 2022-09-16 utils/isolation_forest_wrapper.py

@@ -1,0 +1,37 @@
+import numpy as np
+from sklearn.ensemble import IsolationForest
+
+
+class IsolationForestWrapper(IsolationForest):
+    """ Wraps the scikit-learn Isolation Forest model to adapt it to our test harness. This is because the original gives non-standard outputs when predicting """
+
+    def __init__(
+        self,
+        *,
+        n_estimators=100,
+        max_samples="auto",
+        contamination="auto",
+        max_features=1.0,
+        bootstrap=False,
+        n_jobs=None,
+        random_state=None,
+        verbose=0,
+        warm_start=False,
+    ):
+        super().__init__(
+            bootstrap=bootstrap,
+            n_estimators=n_estimators,
+            max_samples=max_samples,
+            max_features=max_features,
+            warm_start=warm_start,
+            n_jobs=n_jobs,
+            random_state=random_state,
+            verbose=verbose,
+            contamination=contamination,
+        )
+
+    def predict(self, X):
+        return np.fromiter(map({-1: 0, 1: 1}.get, super().predict(X)), dtype=int)
+
+    def decision_function(self, X):
+        return -super().decision_function(X)

### 2022-09-16 dataset.py
.

@@ -113,7 +113,7 @@ class SCIData(pd.DataFrame):
         if onehot:
             r = r.encode_ccs_onehot()
 
-        return r
+        return SCIData(r)
 
     def _regroup_ccs(self, df, col, onehot=False):
         """Joins SCI and a given grouping table for CCS. Matches ICD-10 diagnoses with CCS if this has not already been done
@@ -131,7 +131,7 @@ class SCIData(pd.DataFrame):
         if onehot:
             r = r.encode_ccs_onehot()
 
-        return r
+        return SCIData(r)
 
     def augment_shmi(self, infile="data/ccs.h5", shmi=None, onehot=False):
         """Joins SCI and the SHMI matching table.
@@ -961,7 +961,7 @@ class SCIData(pd.DataFrame):
         if ordinal_encoding:
             X = SCIData(X).ordinal_encode_categories()
 
-        return X.copy(), y
+        return SCIData(X), y
 
     def drop(self, cols, **kwargs):
         return SCIData(super(SCIData, self).drop(cols, **kwargs))

### 2022-09-16 utils/evaluation.py

@@ -26,8 +26,19 @@ from imblearn.pipeline import Pipeline as ImbPipeline
 from imblearn.under_sampling import RandomUnderSampler, TomekLinks
 from imblearn.over_sampling import SMOTE, SMOTENC
 
+from pytorch_tabnet.metrics import Metric as TabNetMetric
+
 
 f2_score = make_scorer(fbeta_score, beta=2)
+
+class F2TabNet(TabNetMetric):
+    def __init__(self):
+        self._name = "F2"
+        self._maximize = True
+
+    def __call__(self, y_true, y_score):
+        y_pred = np.argmax(y_score, axis=1)
+        return fbeta_score(y_true, y_pred, beta=2)
 
 METRICS = {
     "Accuracy": "accuracy",

### 2022-09-17 utils/evaluation.py
.

@@ -20,7 +20,7 @@ from sklearn.metrics import (
     RocCurveDisplay,
     PrecisionRecallDisplay,
 )
-
+from sklearn.dummy import DummyClassifier
 
 from imblearn.pipeline import Pipeline as ImbPipeline
 from imblearn.under_sampling import RandomUnderSampler, TomekLinks
@@ -31,6 +31,7 @@ from pytorch_tabnet.metrics import Metric as TabNetMetric
 
 f2_score = make_scorer(fbeta_score, beta=2)
 
+
 class F2TabNet(TabNetMetric):
     def __init__(self):
         self._name = "F2"
@@ -39,6 +40,7 @@ class F2TabNet(TabNetMetric):
     def __call__(self, y_true, y_score):
         y_pred = np.argmax(y_score, axis=1)
         return fbeta_score(y_true, y_pred, beta=2)
+
 
 METRICS = {
     "Accuracy": "accuracy",
@@ -203,34 +205,34 @@ def with_sampling_strategies(clf, clf_name="Classifier", categorical_cols_idx=[]
 def evaluate_from_pred(
     y_true, y_pred, y_pred_proba, plot_title=None, pos_label=1, save=None
 ):
-    display(
-        pd.DataFrame(
-            {
-                "Accuracy": accuracy_score(y_true, y_pred),
-                "Precision": precision_score(y_true, y_pred, pos_label=pos_label),
-                "Recall": recall_score(y_true, y_pred, pos_label=pos_label),
-                "AUC": roc_auc_score(y_true, y_pred_proba),
-                "F1 Score": f1_score(y_true, y_pred, pos_label=pos_label),
-                "F2 Score": fbeta_score(y_true, y_pred, beta=2, pos_label=pos_label),
-            },
-            index=["Model"],
-        )
+    metric_df = pd.DataFrame(
+        {
+            "Accuracy": accuracy_score(y_true, y_pred),
+            "Precision": precision_score(y_true, y_pred, pos_label=pos_label),
+            "Recall": recall_score(y_true, y_pred, pos_label=pos_label),
+            "AUC": roc_auc_score(y_true, y_pred_proba),
+            "F1 Score": f1_score(y_true, y_pred, pos_label=pos_label),
+            "F2 Score": fbeta_score(y_true, y_pred, beta=2, pos_label=pos_label),
+        },
+        index=["Model"],
     )
+
[…]

### 2022-09-18 utils/evaluation.py
.

@@ -4,6 +4,8 @@ import numpy as np
 
 from IPython.display import display
 import matplotlib.pyplot as plt
+import seaborn as sns
+
 
 from sklearn.base import clone as clone_estimator
 from sklearn.model_selection import train_test_split, cross_validate
@@ -203,7 +205,13 @@ def with_sampling_strategies(clf, clf_name="Classifier", categorical_cols_idx=[]
 
 
 def evaluate_from_pred(
-    y_true, y_pred, y_pred_proba, plot_title=None, pos_label=1, save=None
+    y_true,
+    y_pred,
+    y_pred_proba,
+    plot_title=None,
+    pos_label=1,
+    save=None,
+    style="darkgrid",
 ):
     metric_df = pd.DataFrame(
         {
@@ -219,6 +227,7 @@ def evaluate_from_pred(
 
     display(metric_df)
 
+    sns.set_style(style)
     fig, ax = plt.subplots(1, 3, figsize=(16, 5))
 
     ax[2].grid(False)
@@ -244,7 +253,7 @@ def evaluate_from_pred(
     return metric_df, roc_fig, pr_fig, cm_fig
 
 
-def evaluate(model, X, y, plot_title=None, save=None):
+def evaluate(model, X, y, plot_title=None, save=None, style="darkgrid"):
     y_pred = model.predict(X)
     try:
         y_pred_proba = model.predict_proba(X)[:, 1]
@@ -254,7 +263,9 @@ def evaluate(model, X, y, plot_title=None, save=None):
         except AttributeError:
             y_pred_proba = model.score_samples(X)
 
-    return evaluate_from_pred(y, y_pred, y_pred_proba, plot_title=plot_title, save=save)
+    return evaluate_from_pred(
+        y, y_pred, y_pred_proba, plot_title=plot_title, save=save, style=style
+    )
 
 
 def ideal_pr_curve(y, **kwargs):

### 2022-09-18 dataset.py
.

@@ -966,6 +966,12 @@ class SCIData(pd.DataFrame):
     def drop(self, cols, **kwargs):
         return SCIData(super(SCIData, self).drop(cols, **kwargs))
 
+    def derive_long_los(self, over=1, col='LongLOS'):
+        r = self.copy()
+        r[col] = r.TotalLOS >= over
+
+        return SCIData(r)
+
 
 def justify(df, invalid_val=np.nan, axis=1, side="left"):
     """

### 2022-09-18 utils/evaluation.py

@@ -54,6 +54,46 @@ METRICS = {
 }
 
 
+def joint_plot(
+    subplots,
+    filename=None,
+    ax=None,
+    title=None,
+    style="white",
+    legend_location="lower right",
+    baseline_key="Baseline (NEWS)",
+    plot_baseline=True,
+):
+    sns.set_style(style)
+    plt.rc("axes", titlesize=16)
+
+    if ax is None:
+        fig, ax = plt.subplots(1, 1, figsize=(8, 8))
+
+    for key, plot in subplots.items():
+        if key != baseline_key:
+            plot.plot(ax=ax, name=key)
+
+    ax.set_title(title)
+    ax.legend(loc=legend_location)
+
+    if filename:
+        plt.savefig(
+            f"{Notebook.IMAGE_DIR}/{filename}_no_baseline.png", bbox_inches="tight"
+        )
+
+    if baseline_key in subplots.keys() and plot_baseline:
+        subplots[baseline_key].plot(
+            ax=ax, linestyle="--", color="dimgray", name=baseline_key
+        )
+    ax.legend(loc=legend_location)
+
+    if filename:
+        plt.savefig(f"{Notebook.IMAGE_DIR}/{filename}.png", bbox_inches="tight")
+
+    plt.rc("axes", titlesize=12)  # Revert to default
+
+
 def spotCheckDatasets(
     models,
     datasets,

### 2022-09-18 utils/evaluation.py
.

@@ -1,6 +1,8 @@
 from collections import defaultdict
 import pandas as pd
 import numpy as np
+from scipy import stats as st
+import math
 
 from IPython.display import display
 import matplotlib.pyplot as plt
@@ -48,10 +50,33 @@ METRICS = {
     "Accuracy": "accuracy",
     "Precision": "precision",
     "Recall": "recall",
-    "AUC": "roc_auc",
     "F1 Score": "f1",
     "F2 Score": f2_score,
+    "AUC": "roc_auc",
 }
+
+
+def roc_auc_ci(y_true, y_score):
+    """ Computes AUROC with 95% confidence intervals
+    Uses the formula from 
+    :param y_true: True labels or binary label indicators
+    :param y_score: Target scores
+    """
+
+    # See https://stackoverflow.com/a/20864883/7662085
+    # zscore inside of which 95% of data lies
+    za2 = st.norm.ppf(0.975)
+
+    # From: https://gist.github.com/doraneko94/e24643136cfb8baf03ef8a314ab9615c
+    AUC = roc_auc_score(y_true, y_score)
+    N1, N2 = sum(y_true == True), sum(y_true != True)
+    Q1, Q2 = AUC / (2 - AUC), 2 * AUC ** 2 / (1 + AUC)
+    SE_AUC = math.sqrt(
+        (AUC * (1 - AUC) + (N1 - 1) * (Q1 - AUC ** 2) + (N2 - 1) * (Q2 - AUC ** 2))
+        / (N1 * N2)
+    )
+    lower, upper = max(AUC - za2 * SE_AUC, 0), min(AUC + za2 * SE_AUC, 1)
+    return lower, AUC, upper
 
 
 def joint_plot(
@@ -78,9 +103,7 @@ def joint_plot(
     ax.legend(loc=legend_location)
 
     if filename:
-        plt.savefig(
-            f"{Notebook.IMAGE_DIR}/{filename}_no_baseline.png", bbox_inches="tight"
-        )
+        plt.savefig(f"{filename}_no_baseline.png", bbox_inches="tight")
 
     if baseline_key in subplots.keys() and plot_baseline:
         subplots[baseline_key].plot(
@@ -89,7 +112,7 @@ def joint_plot(
     ax.legend(loc=legend_location)
 
     if filename:
-        plt.savefig(f"{Notebook.IMAGE_DIR}/{filename}.png", bbox_inches="tight")
+        plt.savefig(f"{filename}.png", bbox_inches="tight")
 
[…]

### 2022-09-18 old_code/lgbm_sweep.py
.

@@ -1,0 +1,87 @@
+import warnings
+
+warnings.simplefilter(action="ignore", category=FutureWarning)
+from lightgbm import LGBMClassifier
+from imblearn.pipeline import Pipeline as ImbPipeline
+from imblearn.under_sampling import RandomUnderSampler
+from imblearn.over_sampling import SMOTE, SMOTENC
+from sklearn.model_selection import train_test_split
+from sklearn.model_selection import RandomizedSearchCV
+import pickle
+import numpy as np
+import pandas as pd
+from evaluation import f2_score
+import random
+
+from dataset import SCIData, SCICols
+
+# SCIData.load('data/sci.h5').clean_all().filter_vague_diagnoses().derive_readmission().omit_vbg()
+sci = (
+    SCIData.load("data/sci_processed_2.h5")
+    .fix_readmissionband()
+    .derive_critical_event(within=2)
+)
+
+X, y = (
+    sci.omit_redundant()
+    .drop(["ReadmissionBand", "AgeBand"], axis=1)
+    .omit_ae()
+    .raw_news()
+    .mandate_news()
+    .mandate_blood()
+    .augment_hsmr()
+    .encode_ccs_onehot()
+    .xy(
+        outcome="CriticalEvent", dropna=True
+    )  # Use ordinal encoding because of bug in XGB that prevents use of SHAP when pd.categorical is involved
+)
+categorical_cols_idx = X.describe_categories()[0]
+X = X.ordinal_encode_categories()
+
+X_train, X_test, y_train, y_test = train_test_split(
+    X, y, stratify=y, test_size=0.33, random_state=42
+)
+
+print(X_train.shape)
+
+
+model = ImbPipeline(
+    steps=[
+        ("smote", SMOTENC(categorical_features=categorical_cols_idx)),
+        ("LGBM", LGBMClassifier(metric=["auc"], n_jobs=5)),
+    ]
+)
+
+param_grid = {
+    "LGBM__learning_rate": [0.01, 0.025, 0.05, 0.085, 0.1, 0.15, 0.2, 0.25, 0.3],
+    "LGBM__boosting_type": ["gbdt", "dart", "goss"],
+    # "LGBM__sub_feature": np.arange(0, 1, 0.05),
+    "LGBM__num_leaves": np.arange(20, 300, 20),
+    "LGBM__min_child_samples": np.arange(10, 100, 10),
[…]

### 2022-09-18 old_code/ml_script_4.py

@@ -1,0 +1,331 @@
+import numpy as np
+import pandas as pd
+
+from evaluation import f2_score, METRICS
+
+from collections import defaultdict
+from sklearn.model_selection import cross_validate
+from sklearn.base import clone as clone_estimator
+from multiprocessing import Pool
+
+from evaluation import METRICS
+
+
+def cvf(datakey, xy, modelkey, model, fitparams):
+    return pd.DataFrame.from_dict(
+        cross_validate(
+            model,
+            X=xy[0],
+            y=xy[1],
+            cv=5,
+            n_jobs=1,
+            fit_params=fitparams,
+            scoring=METRICS,
+        )
+    ).assign(dataset=datakey, model=modelkey)
+
+
+def spotCheckDatasets(
+    models,
+    datasets,
+    cv=5,
+    set_params={},
+    order=["dataset", "model"],
+    fit_params={},
+    scoring=METRICS,
+):
+    """ Run stratified k-fold cross-validation on the given models over the given dataset variants
+        :param models: Single estimator or dict of the format {'Model Name': model}
+        :param datasets: Single tuple (X, y) or dict of the format {'Dataset name': (X, y)}
+        :param set_params: Hook to set model parameters based on each dataset. Callable of type function(X, y) -> Dict
+    """
+
+    models = {"Model": models} if type(models) != dict else models
+    datasets = {"Data": datasets} if type(datasets) != dict else datasets
+
+    param_callables = defaultdict(lambda: lambda X, y: dict())
+    param_callables.update(set_params)
+
+    fit_param_dict = defaultdict(lambda: dict())
+    fit_param_dict.update(fit_params)
+
+    tests = [
+        (
+            datakey,
+            (X, y),
+            modelkey,
+            clone_estimator(model).set_params(**(param_callables[modelkey](X, y))),
+            fit_param_dict[modelkey],
+        )
+        for datakey, (X, y) in datasets.items()
+        for modelkey, model in models.items()
+    ]
+
+    print("Spinning up 5 processes")
+    with Pool(24) as p:
+        results = p.starmap(cvf, tests)
+
+    return pd.concat(results).groupby(["dataset", "model"]).mean()
+
+
[…]

### 2022-09-18 old_code/tabnet_sweep.py

@@ -1,0 +1,97 @@
+import warnings
+
+warnings.simplefilter(action="ignore", category=FutureWarning)
+import torch
+from pytorch_tabnet.tab_model import TabNetClassifier
+from imblearn.pipeline import Pipeline as ImbPipeline
+from imblearn.under_sampling import RandomUnderSampler
+from sklearn.model_selection import train_test_split
+from sklearn.model_selection import RandomizedSearchCV
+import pickle
+import numpy as np
+import pandas as pd
+from evaluation import f2_score
+import random
+
+from dataset import SCIData, SCICols
+
+# SCIData.load('data/sci.h5').clean_all().filter_vague_diagnoses().derive_readmission().omit_vbg()
+sci = (
+    SCIData.load("data/sci_processed_2.h5")
+    .fix_readmissionband()
+    .derive_critical_event(within=2)
+)
+
+X, y = (
+    sci.omit_redundant()
+    .drop(["ReadmissionBand", "AgeBand"], axis=1)
+    .omit_ae()
+    .raw_news()
+    .mandate_news()
+    .mandate_blood()
+    .augment_hsmr()
+    .encode_ccs_onehot()
+    .xy(outcome="CriticalEvent", ordinal_encoding=True, dropna=True)
+)
+
+categorical_cols_idx, categorical_cols_dims = X.describe_categories()
+
+X_train, X_test, y_train, y_test = train_test_split(
+    X.to_numpy(), y.to_numpy(), stratify=y, test_size=0.25, random_state=42
+)
+X_train, X_valid, y_train, y_valid = train_test_split(
+    X_train, y_train, stratify=y_train, test_size=0.33, random_state=42
+)
+
+tabnet_params = dict(
+    cat_idxs=categorical_cols_idx,
+    cat_dims=categorical_cols_dims,
+    cat_emb_dim=1,
+    optimizer_fn=torch.optim.Adam,
+    # optimizer_params=dict(lr=2e-2),
+    #   scheduler_params=dict(step_size=50, gamma=0.9),
+    scheduler_fn=torch.optim.lr_scheduler.StepLR,
+    mask_type="entmax",
+    verbose=0,
+)
+
+tabnet_model = TabNetClassifier(**tabnet_params)
+
+
+param_grid = {
+    #'batch_size': [300],
+    "n_a": np.arange(4, 100, 4),
[…]

### 2022-09-19 utils/evaluation.py
.

@@ -10,7 +10,11 @@ import seaborn as sns
 
 
 from sklearn.base import clone as clone_estimator
-from sklearn.model_selection import train_test_split, cross_validate
+from sklearn.model_selection import (
+    train_test_split,
+    cross_validate,
+    StratifiedShuffleSplit,
+)
 from sklearn.metrics import (
     accuracy_score,
     precision_score,
@@ -77,6 +81,21 @@ def roc_auc_ci(y_true, y_score):
     )
     lower, upper = max(AUC - za2 * SE_AUC, 0), min(AUC + za2 * SE_AUC, 1)
     return lower, AUC, upper
+
+
+def roc_auc_ci_bootstrap(y_true, y_score):
+    """ Computes AUROC with 95% confidence intervals by boostrapping """
+    res = st.bootstrap(
+        data=(y_true.to_numpy(), y_score),
+        statistic=roc_auc_score,
+        confidence_level=0.95,
+        method="percentile",
+        vectorized=False,
+        paired=True,
+        random_state=42,
+    )
+
+    return res.confidence_interval.low, res.confidence_interval.high
 
 
 def joint_plot(
@@ -276,7 +295,7 @@ def evaluate_from_pred(
     save=None,
     style="darkgrid",
 ):
-    lower, auc, upper = roc_auc_ci(y_true, y_pred_proba)
+    lower, upper = roc_auc_ci_bootstrap(y_true, y_pred_proba)
     metric_df = pd.DataFrame(
         {
             "Accuracy": accuracy_score(y_true, y_pred),
@@ -285,7 +304,7 @@ def evaluate_from_pred(
             "F1 Score": f1_score(y_true, y_pred, pos_label=pos_label),
             "F2 Score": fbeta_score(y_true, y_pred, beta=2, pos_label=pos_label),
             "AUC": roc_auc_score(y_true, y_pred_proba),
-            "AUC_CI": f"{auc:.3f} ({lower:.3f}-{upper:.3f})",
+            "AUC_CI": f"{lower:.3f}-{upper:.3f}",
         },
         index=["Model"],
     )

### 2022-09-20 dataset.py
Cleanup

@@ -742,7 +742,7 @@ class SCIData(pd.DataFrame):
             "c_Breathing_device": "A - Air",
             "c_Oxygen_flow_rate": 0.0,
             "c_Pain": False,
-            "c_Alert": False,
+            "c_Alert": True,
             "c_Nausea": False,
             "c_Vomiting_since_last_round": False,
             "c_Lying_down": False,
@@ -966,7 +966,7 @@ class SCIData(pd.DataFrame):
     def drop(self, cols, **kwargs):
         return SCIData(super(SCIData, self).drop(cols, **kwargs))
 
-    def derive_long_los(self, over=1, col='LongLOS'):
+    def derive_long_los(self, over=1, col="LongLOS"):
         r = self.copy()
         r[col] = r.TotalLOS >= over
 
@@ -1209,7 +1209,7 @@ class SCICols:
     ]
 
     admission = [
-        "ElectiveAdmission",
+        "AdmittedAfterAEC",
         "AdmissionMethodDescription",
         "AssessmentAreaAdmission",
         "AssessmentAreaDischarge",
@@ -1406,7 +1406,7 @@ class SCICols:
     xgb_types = {
         "Female": "i",
         "Age": "int",
-        "ElectiveAdmission": "i",
+        "AdmittedAfterAEC": "i",
         "AdmissionMethodDescription": "c",
         "AdmissionSpecialty": "c",
         "AgeBand": "c",

### 2022-09-21 dataset.py
Added parameter sweeps

@@ -294,7 +294,11 @@ class SCIData(pd.DataFrame):
         :return: New SCIData instance with the new feature added
         """
         r = self.copy()
-        r[col_name] = r.DiedDuringStay & (r.TotalLOS <= within)
+
+        if within is not None:
+            r[col_name] = r.DiedDuringStay & (r.TotalLOS <= within)
+        else:
+            r[col_name] = r.DiedDuringStay
 
         return SCIData(r)
 
@@ -326,7 +330,8 @@ class SCIData(pd.DataFrame):
         los_on_critical_admission.index = los_on_critical_admission.index.droplevel(1)
 
         # r["CriticalCare"] = m.any(axis=1)
-        r[col_name] = los_on_critical_admission <= within
+        
+        r[col_name] = los_on_critical_admission <= (within or 999)
         r[col_name].fillna(False, inplace=True)
 
         return SCIData(r)
@@ -840,7 +845,7 @@ class SCIData(pd.DataFrame):
 
         return SCIData(r)
 
-    def derive_critical_event(self, within=None, col_name="CriticalEvent"):
+    def derive_critical_event(self, within=None, col_name="CriticalEvent", return_subcols=False):
         """Determines the patients' critical event outcome.
         :param within: Time since admission to consider a critical event. E.g., 1.0 means it occurred within 24 hours, otherwise lived past 24 hours
         :return: New SCIData instance with the new feature added
@@ -852,6 +857,9 @@ class SCIData(pd.DataFrame):
 
         r = self.copy()
         r[col_name] = col
+        if return_subcols:
+            r['DiedWithinThreshold'] = temp.DiedWithinThreshold
+            r['CriticalCare'] = temp.CriticalCare
 
         return SCIData(r)
 

### 2022-09-21 hyperparameter_tuning.py

@@ -1,0 +1,350 @@
+import warnings
+import numpy as np
+import torch, optuna
+from pytorch_tabnet.tab_model import TabNetClassifier
+from sklearn.model_selection import StratifiedKFold, cross_validate
+
+
+from xgboost import XGBClassifier
+from imblearn.pipeline import Pipeline as ImbPipeline
+from imblearn.under_sampling import RandomUnderSampler
+
+import optuna.integration.lightgbm as lgb
+from lightgbm import early_stopping
+from lightgbm import log_evaluation
+
+from sklearn.ensemble import RandomForestClassifier
+from sklearn.linear_model import LogisticRegression
+from utils.isolation_forest_wrapper import IsolationForestWrapper, isolationforestsplit
+
+optuna.logging.set_verbosity(optuna.logging.WARNING)
+
+
+class TabnetObjective:
+    def __init__(self, X_train, y_train, categorical_cols_idx, categorical_cols_dims):
+        (
+            self.X_train,
+            self.y_train,
+            self.categorical_cols_idx,
+            self.categorical_cols_dims,
+        ) = (
+            X_train.to_numpy(),
+            y_train.to_numpy(),
+            categorical_cols_idx,
+            categorical_cols_dims,
+        )
+
+    def __call__(self, trial):
+        mask_type = trial.suggest_categorical("mask_type", ["entmax", "sparsemax"])
+        n_da = trial.suggest_int("n_da", 56, 64, step=4)
+        n_steps = trial.suggest_int("n_steps", 1, 3, step=1)
+        gamma = trial.suggest_float("gamma", 1.0, 1.4, step=0.2)
+        n_shared = trial.suggest_int("n_shared", 1, 3)
+        lambda_sparse = trial.suggest_float("lambda_sparse", 1e-6, 1e-3, log=True)
+
+        tabnet_params = dict(
+            cat_idxs=self.categorical_cols_idx,
+            cat_dims=self.categorical_cols_dims,
+            n_d=n_da,
+            n_a=n_da,
+            n_steps=n_steps,
+            gamma=gamma,
+            lambda_sparse=lambda_sparse,
+            optimizer_fn=torch.optim.Adam,
+            optimizer_params=dict(lr=2e-2, weight_decay=1e-5),
+            mask_type=mask_type,
+            n_shared=n_shared,
+            scheduler_params=dict(
+                mode="min",
+                patience=trial.suggest_int(
+                    "patienceScheduler", low=3, high=10
[…]

### 2022-09-21 utils/evaluation.py

@@ -83,13 +83,14 @@ def roc_auc_ci(y_true, y_score):
     return lower, AUC, upper
 
 
-def roc_auc_ci_bootstrap(y_true, y_score):
+def roc_auc_ci_bootstrap(y_true, y_score, n_resamples=9999):
     """ Computes AUROC with 95% confidence intervals by boostrapping """
     res = st.bootstrap(
         data=(y_true.to_numpy(), y_score),
         statistic=roc_auc_score,
         confidence_level=0.95,
         method="percentile",
+        n_resamples=n_resamples,
         vectorized=False,
         paired=True,
         random_state=42,
@@ -294,8 +295,9 @@ def evaluate_from_pred(
     pos_label=1,
     save=None,
     style="darkgrid",
+    n_resamples=9999,
 ):
-    lower, upper = roc_auc_ci_bootstrap(y_true, y_pred_proba)
+    lower, upper = roc_auc_ci_bootstrap(y_true, y_pred_proba, n_resamples)
     metric_df = pd.DataFrame(
         {
             "Accuracy": accuracy_score(y_true, y_pred),
@@ -337,7 +339,9 @@ def evaluate_from_pred(
     return metric_df, roc_fig, pr_fig, cm_fig
 
 
-def evaluate(model, X, y, plot_title=None, save=None, style="darkgrid"):
+def evaluate(
+    model, X, y, plot_title=None, save=None, style="darkgrid", n_resamples=9999
+):
     y_pred = model.predict(X)
     try:
         y_pred_proba = model.predict_proba(X)[:, 1]
@@ -348,7 +352,13 @@ def evaluate(model, X, y, plot_title=None, save=None, style="darkgrid"):
             y_pred_proba = model.score_samples(X)
 
     return evaluate_from_pred(
-        y, y_pred, y_pred_proba, plot_title=plot_title, save=save, style=style
+        y,
+        y_pred,
+        y_pred_proba,
+        plot_title=plot_title,
+        save=save,
+        style=style,
+        n_resamples=n_resamples,
     )
 
 

### 2022-09-21 utils/isolation_forest_wrapper.py

@@ -1,5 +1,20 @@
 import numpy as np
+import pandas as pd
 from sklearn.ensemble import IsolationForest
+
+
+def isolationforestsplit(X_train, y_train, kf):
+    for train_index, test_index in kf.split(X_train, y_train):
+        if type(y_train) == pd.Series:
+            y_train = y_train.to_numpy()
+        trainy, validy = (
+            y_train[train_index],
+            y_train[test_index],
+        )
+
+        test_index = np.concatenate((train_index[trainy], test_index))
+        train_index = train_index[~trainy]
+        yield train_index, test_index
 
 
 class IsolationForestWrapper(IsolationForest):

### 2022-09-21 hyperparameter_tuning.py
.

@@ -90,10 +90,12 @@ class TabnetObjective:
         return avg
 
 
-def tune_tabnet(X_train, y_train, categorical_cols_idx, categorical_cols_dims):
+def tune_tabnet(
+    X_train, y_train, categorical_cols_idx, categorical_cols_dims, timeout=60 * 60
+):
     obj = TabnetObjective(X_train, y_train, categorical_cols_idx, categorical_cols_dims)
     study = optuna.create_study(direction="maximize", study_name="TabNet optimization")
-    study.optimize(obj, n_trials=1, n_jobs=-1, timeout=60 * 60)
+    study.optimize(obj, n_trials=1, n_jobs=-1, timeout=timeout)
     tabnet_params = dict(
         cat_idxs=categorical_cols_idx,
         cat_dims=categorical_cols_dims,
@@ -185,7 +187,7 @@ class XgboostObjective:
         model = ImbPipeline(
             steps=[("IMB", RandomUnderSampler()), ("XGB", XGBClassifier()),]
         ).set_params(**param)
-        cv = cross_validate(model, X_train, y_train, cv=5, scoring="roc_auc",)
+        cv = cross_validate(model, self.X_train, self.y_train, cv=5, scoring="roc_auc",)
         return cv["test_score"].mean()
 
 
@@ -292,7 +294,7 @@ class IsolationForestObjective:
                 self.y_train,
                 StratifiedKFold(n_splits=5, random_state=42, shuffle=True),
             ),
-            scoring='roc_auc'
+            scoring="roc_auc",
         )
         return cv["test_score"].mean()
 
@@ -312,30 +314,37 @@ def tune_isolationforest(X_train, y_train, n_trials=100, timeout=60 * 60, n_jobs
 
 class LogisticObjective:
     def __init__(self, X_train, y_train):
-        self.X_train, self.y_train = \
-            X_train, y_train
+        self.X_train, self.y_train = X_train, y_train
 
     def __call__(self, trial):
         param = {}
-        param['LR__penalty'] = trial.suggest_categorical('LR__penalty', ['l2', 'l1'])
-        if param['LR__penalty'] == 'l1':
-            param['LR__solver'] = 'saga'
+        param["LR__penalty"] = trial.suggest_categorical("LR__penalty", ["l2", "l1"])
+        if param["LR__penalty"] == "l1":
+            param["LR__solver"] = "saga"
         else:
-            param['LR__solver'] = 'lbfgs'
[…]

### 2022-09-22 hyperparameter_tuning.py
.

@@ -17,8 +17,7 @@ from sklearn.ensemble import RandomForestClassifier
 from sklearn.linear_model import LogisticRegression
 from utils.isolation_forest_wrapper import IsolationForestWrapper, isolationforestsplit
 
-optuna.logging.set_verbosity(optuna.logging.WARNING)
-
+optuna.logging.set_verbosity(optuna.logging.DEBUG)
 
 class TabnetObjective:
     def __init__(self, X_train, y_train, categorical_cols_idx, categorical_cols_dims):
@@ -91,11 +90,16 @@ class TabnetObjective:
 
 
 def tune_tabnet(
-    X_train, y_train, categorical_cols_idx, categorical_cols_dims, timeout=60 * 60
+    X_train,
+    y_train,
+    categorical_cols_idx,
+    categorical_cols_dims,
+    timeout=60 * 60,
+    n_jobs=-1,
 ):
     obj = TabnetObjective(X_train, y_train, categorical_cols_idx, categorical_cols_dims)
     study = optuna.create_study(direction="maximize", study_name="TabNet optimization")
-    study.optimize(obj, n_trials=1, n_jobs=-1, timeout=timeout)
+    study.optimize(obj, n_trials=1, n_jobs=n_jobs, timeout=timeout)
     tabnet_params = dict(
         cat_idxs=categorical_cols_idx,
         cat_dims=categorical_cols_dims,
@@ -131,6 +135,7 @@ class XgboostObjective:
     def __call__(self, trial):
         param = {
             "XGB__verbosity": 0,
+            "XGB__n_jobs": 1,
             "XGB__objective": "binary:logistic",
             "XGB__enable_categorical": True,
             # use exact for small dataset.
@@ -210,6 +215,7 @@ def tune_lgbm(X_train, y_train, timeout=60 * 60):
         "boosting_type": "gbdt",
         "is_unbalance": True,
         "verbose": -1,
+        "n_jobs": 1
     }
 
     tuner = lgb.LightGBMTunerCV(

### 2022-09-22 hyperparameter_tuning.py
.

@@ -19,6 +19,7 @@ from utils.isolation_forest_wrapper import IsolationForestWrapper, isolationfore
 
 optuna.logging.set_verbosity(optuna.logging.DEBUG)
 
+
 class TabnetObjective:
     def __init__(self, X_train, y_train, categorical_cols_idx, categorical_cols_dims):
         (
@@ -66,7 +67,7 @@ class TabnetObjective:
         )  # early stopping
         kf = StratifiedKFold(n_splits=5, random_state=42, shuffle=True)
         CV_score_array = []
-        for train_index, test_index in kf.split(X_train, y_train):
+        for train_index, test_index in kf.split(self.X_train, self.y_train):
             X_train_tn, X_valid_tn = (
                 self.X_train[train_index],
                 self.X_train[test_index],
@@ -136,57 +137,60 @@ class XgboostObjective:
         param = {
             "XGB__verbosity": 0,
             "XGB__n_jobs": 1,
+            "XGB__missing": -1,
             "XGB__objective": "binary:logistic",
             "XGB__enable_categorical": True,
             # use exact for small dataset.
             "XGB__tree_method": trial.suggest_categorical(
-                "tree_method", ["approx", "hist"]
+                "XGB__tree_method", ["approx", "hist"]
             ),
             # defines booster, gblinear for linear functions.
             "XGB__booster": trial.suggest_categorical(
-                "booster", ["gbtree", "gblinear", "dart"]
+                "XGB__booster", ["gbtree", "gblinear", "dart"]
             ),
             # L2 regularization weight.
-            "XGB__lambda": trial.suggest_float("lambda", 1e-8, 1.0, log=True),
+            "XGB__lambda": trial.suggest_float("XGB__lambda", 1e-8, 1.0, log=True),
             # L1 regularization weight.
-            "XGB__alpha": trial.suggest_float("alpha", 1e-8, 1.0, log=True),
+            "XGB__alpha": trial.suggest_float("XGB__alpha", 1e-8, 1.0, log=True),
             # sampling ratio for training data.
-            "XGB__subsample": trial.suggest_float("subsample", 0.2, 1.0),
+            "XGB__subsample": trial.suggest_float("XGB__subsample", 0.2, 1.0),
             # sampling according to each tree.
-            "XGB__colsample_bytree": trial.suggest_float("colsample_bytree", 0.2, 1.0),
[…]

### 2022-09-22 utils/data_profiling.py

@@ -6,6 +6,21 @@ import matplotlib.dates as mdates
 import matplotlib.ticker as ticker
 import seaborn as sns
 
+def drop_exclusive_cols(X1, X2):
+    exclusive_cols = set(X1.columns) ^ set(X2.columns)
+    X1.drop(exclusive_cols, axis=1, errors="ignore", inplace=True)
+    X2.drop(exclusive_cols, axis=1, errors="ignore", inplace=True)
+
+
+def ensure_categorical_overlap(X1, X2, cols_to_check):
+    for col in cols_to_check:
+        col = X1.columns[col]
+        X1_uniq, X2_uniq = X1[col].unique(), X2[col].unique()
+        to_remove = list(set(X1_uniq) ^ set(X2_uniq))
+        repl = list(set(X1_uniq) & set(X2_uniq))[0]
+        X1.loc[X1[col].isin(to_remove), col] = repl
+        X2.loc[X2[col].isin(to_remove), col] = repl
+
 
 def missing_data(
     df, subset=[], title="Proportion of missing values per column", ax=None

### 2022-09-22 hyperparameter_tuning.py
.

@@ -82,7 +82,7 @@ class TabnetObjective:
                 y_train=y_train_tn,
                 eval_set=[(X_valid_tn, y_valid_tn)],
                 patience=trial.suggest_int("patience", low=15, high=30),
-                max_epochs=trial.suggest_int("epochs", 1, 100),
+                max_epochs=50,
                 eval_metric=["auc"],
             )
             CV_score_array.append(clf.best_cost)
@@ -126,7 +126,7 @@ def tune_tabnet(
     print("BEST PARAMETERS")
     print(study.best_params)
 
-    return tabnet_params, study.best_params["epochs"], study.best_params["patience"]
+    return tabnet_params, 100, study.best_params["patience"]
 
 
 class XgboostObjective:
@@ -140,14 +140,15 @@ class XgboostObjective:
             "XGB__missing": -1,
             "XGB__objective": "binary:logistic",
             "XGB__enable_categorical": True,
+            "XGB__booster": "gbtree",
             # use exact for small dataset.
             "XGB__tree_method": trial.suggest_categorical(
                 "XGB__tree_method", ["approx", "hist"]
             ),
             # defines booster, gblinear for linear functions.
-            "XGB__booster": trial.suggest_categorical(
-                "XGB__booster", ["gbtree", "gblinear", "dart"]
-            ),
+            # "XGB__booster": trial.suggest_categorical(
+            #     "XGB__booster", ["gbtree", "gblinear", "dart"]  # Cant use this, due to SHAP limitations
+            # ),
             # L2 regularization weight.
             "XGB__lambda": trial.suggest_float("XGB__lambda", 1e-8, 1.0, log=True),
             # L1 regularization weight.

### 2022-09-22 utils/evaluation.py

@@ -23,6 +23,7 @@ from sklearn.metrics import (
     f1_score,
     fbeta_score,
     make_scorer,
+    confusion_matrix,
     precision_recall_curve,
     ConfusionMatrixDisplay,
     RocCurveDisplay,
@@ -312,6 +313,8 @@ def evaluate_from_pred(
     )
 
     display(metric_df)
+
+    display(confusion_matrix(y_true, y_pred))
 
     sns.set_style(style)
     fig, ax = plt.subplots(1, 3, figsize=(16, 5))

### 2022-09-23 dataset.py
.

@@ -311,7 +311,9 @@ class SCIData(pd.DataFrame):
         :return: New SCIData instance with the new features added
         """
         r = self.copy()
-        m = r[SCICols.wards].isin(critical_wards)
+        m = r[SCICols.wards].isin(
+            critical_wards
+        )  # Can also consider OPCS: ['E85.1', 'X50.3', 'X50.4'] for on-ward critical care but there are very few
         column_where_critical_appeared = m.idxmax(axis=1).where(m.any(1)).dropna()
         los_on_critical_admission = (
             r[SCICols.ward_los]
@@ -330,7 +332,7 @@ class SCIData(pd.DataFrame):
         los_on_critical_admission.index = los_on_critical_admission.index.droplevel(1)
 
         # r["CriticalCare"] = m.any(axis=1)
-        
+
         r[col_name] = los_on_critical_admission <= (within or 999)
         r[col_name].fillna(False, inplace=True)
 
@@ -845,7 +847,9 @@ class SCIData(pd.DataFrame):
 
         return SCIData(r)
 
-    def derive_critical_event(self, within=None, col_name="CriticalEvent", return_subcols=False):
+    def derive_critical_event(
+        self, within=None, col_name="CriticalEvent", return_subcols=False
+    ):
         """Determines the patients' critical event outcome.
         :param within: Time since admission to consider a critical event. E.g., 1.0 means it occurred within 24 hours, otherwise lived past 24 hours
         :return: New SCIData instance with the new feature added
@@ -858,8 +862,8 @@ class SCIData(pd.DataFrame):
         r = self.copy()
         r[col_name] = col
         if return_subcols:
-            r['DiedWithinThreshold'] = temp.DiedWithinThreshold
-            r['CriticalCare'] = temp.CriticalCare
+            r["DiedWithinThreshold"] = temp.DiedWithinThreshold
+            r["CriticalCare"] = temp.CriticalCare
 
         return SCIData(r)
 

### 2022-09-23 hyperparameter_tuning.py

@@ -1,11 +1,9 @@
 import warnings
 import numpy as np
 import torch, optuna
-from pytorch_tabnet.tab_model import TabNetClassifier
+
 from sklearn.model_selection import StratifiedKFold, cross_validate
 
-
-from xgboost import XGBClassifier
 from imblearn.pipeline import Pipeline as ImbPipeline
 from imblearn.under_sampling import RandomUnderSampler
 
@@ -13,11 +11,14 @@ import optuna.integration.lightgbm as lgb
 from lightgbm import early_stopping
 from lightgbm import log_evaluation
 
+from xgboost import XGBClassifier
+from lightgbm import LGBMClassifier
 from sklearn.ensemble import RandomForestClassifier
 from sklearn.linear_model import LogisticRegression
 from utils.isolation_forest_wrapper import IsolationForestWrapper, isolationforestsplit
+from pytorch_tabnet.tab_model import TabNetClassifier
 
-optuna.logging.set_verbosity(optuna.logging.DEBUG)
+optuna.logging.set_verbosity(optuna.logging.WARNING)
 
 
 class TabnetObjective:
@@ -84,6 +85,8 @@ class TabnetObjective:
                 patience=trial.suggest_int("patience", low=15, high=30),
                 max_epochs=50,
                 eval_metric=["auc"],
+                weights=1,
+                drop_last=False,
             )
             CV_score_array.append(clf.best_cost)
         avg = np.mean(CV_score_array)
@@ -212,32 +215,105 @@ def tune_xgboost(X_train, y_train, n_trials=100, timeout=60 * 60, n_jobs=-1):
     return study.best_params
 
 
-def tune_lgbm(X_train, y_train, timeout=60 * 60, n_jobs=-1):
-    dtrain = lgb.Dataset(X_train, label=y_train)
-    params = {
-        "objective": "binary",
-        "metrics": ["binary_logloss", "auc"],
-        "boosting_type": "gbdt",
-        "is_unbalance": True,
-        "verbose": -1,
-        "n_jobs": 1,
-    }
+class LgbmObjective:
+    def __init__(self, X_train, y_train, categorical_cols_idx):
+        self.X_train, self.y_train, self.categorical_cols_idx = (
+            X_train,
+            y_train,
+            categorical_cols_idx,
+        )
 
-    tuner = lgb.LightGBMTunerCV(
-        params,
-        dtrain,
-        folds=StratifiedKFold(n_splits=3),
-        callbacks=[
[…]

### 2022-09-23 hyperparameter_tuning.py
Added cuda support because tabnet was way too slow

@@ -36,25 +36,15 @@ class TabnetObjective:
         )
 
     def __call__(self, trial):
-        mask_type = trial.suggest_categorical("mask_type", ["entmax", "sparsemax"])
         n_da = trial.suggest_int("n_da", 56, 64, step=4)
-        n_steps = trial.suggest_int("n_steps", 1, 3, step=1)
-        gamma = trial.suggest_float("gamma", 1.0, 1.4, step=0.2)
-        n_shared = trial.suggest_int("n_shared", 1, 3)
-        lambda_sparse = trial.suggest_float("lambda_sparse", 1e-6, 1e-3, log=True)
 
         tabnet_params = dict(
             cat_idxs=self.categorical_cols_idx,
             cat_dims=self.categorical_cols_dims,
-            n_d=n_da,
-            n_a=n_da,
-            n_steps=n_steps,
-            gamma=gamma,
-            lambda_sparse=lambda_sparse,
             optimizer_fn=torch.optim.Adam,
-            optimizer_params=dict(lr=2e-2, weight_decay=1e-5),
-            mask_type=mask_type,
-            n_shared=n_shared,
+            scheduler_fn=torch.optim.lr_scheduler.ReduceLROnPlateau,
+            verbose=0,
+            device_name="cuda" if torch.cuda.is_available() else "cpu",
             scheduler_params=dict(
                 mode="min",
                 patience=trial.suggest_int(
@@ -63,25 +53,26 @@ class TabnetObjective:
                 min_lr=1e-5,
                 factor=0.5,
             ),
-            scheduler_fn=torch.optim.lr_scheduler.ReduceLROnPlateau,
-            verbose=0,
-        )  # early stopping
-        kf = StratifiedKFold(n_splits=5, random_state=42, shuffle=True)
+
+            n_d=n_da,
+            n_a=n_da,
+            n_steps=trial.suggest_int("n_steps", 1, 3, step=1),
+            gamma=trial.suggest_float("gamma", 1.0, 1.4, step=0.2),
+            lambda_sparse=trial.suggest_float("lambda_sparse", 1e-6, 1e-3, log=True),
+            optimizer_params=dict(lr=2e-2, weight_decay=1e-5),
+            mask_type=trial.suggest_categorical("mask_type", ["entmax", "sparsemax"]),
+            n_shared=trial.suggest_int("n_shared", 1, 3),
+        ) 
+
         CV_score_array = []
[…]

### 2022-09-26 old_code/ml_script1.py
Uploaded latest results

@@ -1,173 +1,0 @@
-# %%
-import warnings, math, itertools, json, logging
-from hashlib import md5
-
-import numpy as np
-import pandas as pd
-
-from IPython.display import display
-import matplotlib.pyplot as plt
-import seaborn as sns
-
-
-# %%
-pd.set_option("display.max_columns", None)
-sns.set_theme(style="darkgrid", palette="colorblind")
-sns.set(rc={"figure.figsize": (11.5, 8.5), "figure.dpi": 100})
-
-logging.basicConfig(
-    level=logging.INFO, format="%(asctime)s [%(levelname)s] %(message)s"
-)
-
-from dataset import SCIData, SCICols
-
-# %%
-
-# sci.clean_all().filter_vague_diagnoses().derive_readmission().omit_redundant().omit_vbg().omit_ae().save()
-X, y = (
-    SCIData.load("data/sci_processed.h5")
-    .augment_hsmr(onehot=True)
-    .impute_blood()
-    .raw_news()
-    .mandate_news()
-    .mandate_diagnoses()
-    .xy(outcome="DiedDuringStay")
-)
-
-# %%
-from sklearn.compose import make_column_selector, make_column_transformer
-from sklearn.preprocessing import OneHotEncoder, MinMaxScaler, FunctionTransformer
-from sklearn.pipeline import Pipeline
-
-numeric_pipeline = Pipeline(steps=[("scale", MinMaxScaler()),])
-
-categorical_pipeline = Pipeline(
-    steps=[("one-hot", OneHotEncoder(handle_unknown="ignore", sparse=False)),]
-)
-
-ct = make_column_transformer(
-    (numeric_pipeline, make_column_selector(dtype_include=np.number)),
-    (categorical_pipeline, make_column_selector(dtype_include=object)),
-)
-
-full_processor = Pipeline(steps=[("columns", ct)])
-
-
-# %%
-from sklearn.model_selection import train_test_split
-
-X_train, X_test, y_train, y_test = train_test_split(
-    full_processor.fit_transform(X), y, test_size=0.25
-)
-
-# %%
-from sklearn.metrics import (
-    accuracy_score,
-    precision_score,
-    recall_score,
-    roc_auc_score,
-    f1_score,
-    fbeta_score,
-    make_scorer,
-)
-
-f2_score = make_scorer(fbeta_score, beta=2)
-
-METRICS = {
-    "Accuracy": "accuracy",
[…]

### 2022-09-26 old_code/ml_script_3.py

@@ -1,71 +1,0 @@
-from xgboost import XGBClassifier
-from imblearn.pipeline import Pipeline as ImbPipeline
-from imblearn.under_sampling import RandomUnderSampler
-from sklearn.model_selection import train_test_split
-from sklearn.model_selection import RandomizedSearchCV
-import pickle
-from evaluation import f2_score
-
-from dataset import SCIData, SCICols
-
-# SCIData.load('data/sci.h5').clean_all().filter_vague_diagnoses().derive_readmission().omit_vbg()
-sci = (
-    SCIData.load("data/sci_processed_2.h5")
-    .fix_readmissionband()
-    .derive_critical_event(within=2)
-)
-
-
-X, y = (
-    sci.omit_redundant()
-    .drop(["ReadmissionBand", "AgeBand"], axis=1)
-    .omit_ae()
-    .raw_news()
-    .mandate_news()
-    .mandate_blood()
-    .augment_hsmr()
-    .encode_ccs_onehot()
-    .xy(
-        outcome="CriticalEvent", dropna=False, ordinal_encoding=True
-    )  # Use ordinal encoding because of bug in XGB that prevents use of SHAP when pd.categorical is involved
-)
-
-X_train, X_test, y_train, y_test = train_test_split(
-    X, y, stratify=y, test_size=0.33, random_state=42
-)
-
-
-model = ImbPipeline(
-    steps=[
-        ("undersampling", RandomUnderSampler(sampling_strategy=0.1)),
-        (
-            "XGB",
-            XGBClassifier(
-                tree_method="approx",
-                enable_categorical=True,
-                scale_pos_weight=(y_train.shape[0] / (2 * y_train.sum())),
-            ),
-        ),
-    ]
-)
-
-param_grid = {
-    "XGB__max_depth": [3, 5, 6, 10, 15, 20],
-    "XGB__learning_rate": [0.01, 0.1, 0.2, 0.3],
-    "XGB__subsample": np.arange(0.5, 1.0, 0.1),
-    "XGB__colsample_bytree": np.arange(0.4, 1.0, 0.1),
-    "XGB__colsample_bylevel": np.arange(0.4, 1.0, 0.1),
-    "XGB__n_estimators": [100, 500, 1000],
-}
-
-clf = RandomizedSearchCV(
-    model, param_distributions=param_grid, scoring=f2_score, n_iter=25, verbose=1
-)
-
-clf.fit(X_train, y_train)
-
[…]

### 2022-09-26 TeX/baseline/paper.tex
Cleanup

@@
   ]{biblatex}
-\addbibresource{references.bib}
+\addbibresource{../../literature/references.bib}
 
@@
 The NEWS was not effective in detecting which patients in our data would be readmitted to the same hospital in the future.
-Prior work has found that clinical data indicating the severity of the patient's condition, including constructed scores, is useful in predicting 30-day readmission \cite[pp.~5]{Mahmoudim20}.
+Prior work has found that clinical data indicating the severity of the patient's condition, including constructed scores, is useful in predicting 30-day readmission \cite[pp.~5]{Mahmoudi20}.
 Dedicated systems for predicting 30-day readmission, such as the HOSPITAL and LACE score, use parameters that the NEWS does not, such as blood test results, procedures the patient underwent, comorbidities and LOS \cite{Donze13}.

### 2022-09-27 utils/evaluation.py
Flipped confusion matrices

@@ -61,6 +61,38 @@ METRICS = {
 }
 
 
+def plot_alert_rate(y_pred_probas, y_test, n_days, intercept="NEWS", save=None):
+    fig, ax = plt.subplots(1, 1, figsize=(7, 7))
+    points = []
+    for model, y_pred_proba in y_pred_probas.items():
+        precision, recall, thresholds = precision_recall_curve(y_test, y_pred_proba)
+        n_pos = (
+            np.array(
+                [
+                    np.where(y_pred_proba > threshold, 1, 0).sum()
+                    for threshold in thresholds
+                ]
+            )
+            / n_days
+        )
+        r = [(*_, model) for _ in zip(recall, n_pos)]
+        if model != "NEWS":
+            r = r[::100]
+        points += r
+
+    data = pd.DataFrame(points, columns=["Sensitivity", "Alerts per day", "Model"])
+
+    sns.lineplot(
+        data=data, x="Sensitivity", y="Alerts per day", hue="Model", ax=ax, linewidth=2
+    )
+    if "NEWS" in y_pred_probas:
+        ax.lines[list(y_pred_probas.keys()).index("NEWS")].set_linestyle("--")
+
+    ax.set_title("Sensitivity vs. Alert Rate")
+    if save:
+        plt.savefig(save, bbox_inches="tight")
+
+
 def roc_auc_ci(y_true, y_score):
     """ Computes AUROC with 95% confidence intervals
     Uses the formula from 
@@ -330,9 +362,18 @@ def evaluate_from_pred(
     if (-1) in np.array(y_true):
         get = {1: True, -1: False}.get
         y_true, y_pred = (list(map(get, y_true)), list(map(get, y_pred)))
-    cm_fig = ConfusionMatrixDisplay.from_predictions(
-        y_true, y_pred, ax=ax[2], normalize="true", values_format=".2%"
-    )
+
+    cm_fig = ConfusionMatrixDisplay(
+        np.rot90(np.flipud(confusion_matrix(y_true, y_pred, normalize="true"))),
+        display_labels=[1, 0],
+    ).plot(values_format=".2%", ax=ax[2])
+
+    ax[2].set_xlabel("True Class")
+    ax[2].set_ylabel("Predicted Class")
+
+    # cm_fig = ConfusionMatrixDisplay.from_predictions(
+    #     y_true, y_pred, ax=ax[1], normalize="true", values_format=".2%"
+    # )
 
     plt.suptitle(plot_title)
 

### 2022-09-27 utils/evaluation.py
.

@@ -378,7 +378,7 @@ def evaluate_from_pred(
     plt.suptitle(plot_title)
 
     if save:
-        plt.savefig(save, bbox_inches="tight")
+        plt.savefig(save, bbox_inches="tight", dpi=100)
 
     return metric_df, roc_fig, pr_fig, cm_fig
 

### 2023-02-19 TeX/acute_deterioration_paper/main.tex
Further cleanup

\documentclass[fleqn,10pt]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{multirow}
\usepackage{bm}
\usepackage{svg}
\usepackage{longtable}
\usepackage{textgreek}
\usepackage{pdflscape}
\usepackage{amsmath, mathtools}
\usepackage{calrsfs}
\DeclareMathAlphabet{\pazocal}{OMS}{zplm}{m}{n}
\svgpath{{images/}}

\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
\DeclarePairedDelimiter\norm{\lVert}{\rVert}%
\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}

\title{Predicting Acute Clinical Deterioration with Interpretable Machine Learning to support Emergency Care Decision Making}

\author[1]{Stelios Boulitsakis Logothetis}
\author[2,3]{Darren Green}
\author[4]{Mark Holland}
\author[1,5]{*Noura Al Moubayed}
\affil[1]{Department of Computer Science, University of Durham, Durham, United Kingdom}
\affil[2]{Northern Care Alliance NHS Foundation Trust, Department of Renal Medicine, Manchester, United Kingdom}
\affil[3]{Division of Cardiovascular Sciences, University of Manchester, Manchester, United Kingdom}
\affil[4]{School of Clinical and Biomedical Sciences, University of Bolton, United Kingdom}
\affil[5]{Evergreen Life Ltd, Manchester, United Kingdom}

%\affil[+]{these authors contributed equally to this work}

%\keywords{Keyword1, Keyword2, Keyword3}

\begin{abstract}
The emergency department is a fast-paced environment responsible for large volumes of patients with varied acuity. Operational pressures on emergency departments are increasing, which creates the imperative to efficiently identify patients at risk of acute deterioration and minimise patients' risk of morbidity and mortality. We apply state-of-the-art machine learning methods to predict patient deterioration early, based on their first recorded vital signs, observations, laboratory results, and other predictors documented in Electronic Patient Records. We build on prior work by incorporating interpretable machine learning and fairness-aware modelling, and achieve improved classification performance over the current standard, the National Early Warning Score 2, measured by average precision and daily alert rate. We use a cross-sectional Electronic Patient Record dataset comprising $121,058$ unplanned admissions at Salford Royal Hospital, UK, to systematically compare model variations for predicting mortality and critical care utilisation within 24 hours of admission. Our models yield up to $0.375$ increase in average precision, up to $18.5\%$ reduction in daily alert rate, and a median $0.577$ reduction in differential bias amplification across the protected demographics of age and sex. We use Shapely Additive exPlanations to justify the models' outputs, verify that the captured data associations align with domain knowledge, and pair predictions with the causal context of each patient’s most influential characteristics. We encourage future research to follow a systematised approach to data-driven risk modelling and help obtain clinically applicable support tools.  
\end{abstract}
\begin{document}

\flushbottom
\maketitle

\thispagestyle{empty}

\section*{Introduction}
When patients deteriorate, care providers must be able to recognise their worsening condition immediately and intervene accordingly \cite{Hillman01b}. Delayed identification of deterioration is associated with preventable hospital deaths \cite{Hogan12}, while delaying the transfer of critically ill patients to intensive care puts them at higher risk of morbidity and mortality \cite{Chalfin07}. The importance of timely identification and appropriate response to clinical instability has motivated the development of 'track-and-trigger' systems. These systems tie clinical observations that are antecedent to patient deterioration with recommended interventions to be executed by care staff or dedicated response teams as part of a rapid response system \cite{Devita10}. In the United Kingdom, this system is recommended by both National Institute for Health and Care Excellence (NICE) and the Royal College of Physicians (RCP) to monitor all adult patients in acute hospital settings \cite{CG50, RCP17}.

In most cases, acute clinical instability and deterioration are preceded by abnormal vital signs \cite{Kause04}, therefore standard practice in acute secondary care settings is to monitor patients using basic homeostatic measures, which include heart rate, blood pressure, inspired oxygen, oxygen saturation, temperature, and level of consciousness \cite{Smith13}. To assist this process, weighted aggregate scores of these measures, known as Early Warning Scores (EWS), have been developed to characterise the patient’s acuity \cite{Morgan07}. These scores can act as the afferent component of a rapid response system, tying them to an escalation protocol or a set of recommended clinical interventions \cite{Devita10}.

Historically, data pertaining to an EWS were manually recorded and tallied on paper charts. As such, they often fell short of including the full breadth and variety of available predictive information \cite{Gerry20}. The gradual phasing-out of bedside paper charts has brought the transition to digital EWS solutions that draw patient data in real-time from Electronic Patient Records (EPR). Beyond digitising conventional EWS, EPR systems collate comprehensive patient data, which can be used to improve performance and clinical utility \cite{Goldstein17}. In particular, the large volume of available data makes it feasible to develop a purely or partly data-driven solution using machine learning. AI-based systems have already demonstrated suitability for assisting in medical imaging tasks, which makes AI-powered prognostic modelling a key research area of interest \cite{Obermeyer16}. Our study concentrates on analysing EPR data to model clinical risk, as we use machine learning methods to potentially identify acute clinical deterioration in patients presenting to the emergency department (ED).

Prior work has used machine learning to model inpatient admission, deterioration, critical care admission, cardiac arrest, and mortality, among other outcomes \cite{Grant20}. In a systematic review of studies published from 2009-2017, Goldstein et al. identified 107 applications of EPR data to training statistical and ML models \cite{Goldstein17}. Recently, Klug et al. \cite{Klug20} used gradient-boosted decision trees (GBDT) on a single-centre cohort of approximately $800,000$ ED episodes to predict short-term mortality risk and achieved improved performance over severity scores such as the Shock Index \cite{Torabi16}. Romero et al. \cite{Romero21} developed a gradient-boosting machine (GBM) model for use as an EWS and demonstrated superior performance compared to the National Early Warning Score 2 (NEWS2) \cite{RCP17}. Finally, Fernandes et al. \cite{Fernandes20} investigated the predictive value of ED patients’ presenting complaints compared to vital signs and other measurements. They used natural language processing (vectorisation with TF-IDF normalisation) to encode unstructured complaint information and trained models on a cohort of approximately $235,000$ patients to predict mortality or cardiac arrest. Their experimental findings showed improved predictive performance and calibration when including the chief complaint as a predictor.

This study applies state-of-the-art methods from contemporary machine learning practice to estimate risk of deterioration for acute medical patients in the ED. We bring together findings from prior studies to improve the differentiation of at-risk patients and address challenges that are prerequisites to clinical deployment for a proposed solution. The ED is a fast-paced environment that treats a large volume of patients with varied acuity and is responsible for their initial assessment and clinical management \cite{Mohr20}. Operational pressures in EDs are steadily increasing \cite{Ams21}, creating an imperative to differentiate the patients with the highest risk efficiently. In our study setting, 'obvious cases' of imminent critical deterioration usually bypass the acute medical team and are escalated immediately. By elimination, the remaining patients are 'less obvious' cases and thus have a greater need for decision support. Conventional, general-purpose EWS are not optimised for specific patient populations or contexts, while 'off-the-shelf' EWS, such as the NEWS2, have variable performance \cite{Alam14}. Recent work argues in favour of centre-specific, locally tailored scores and risk models \cite{Obrien20, Futoma20}; data-driven solutions deployable at scale can fulfil this role.

We systematically compare the performance of various learning algorithms based on logistic regression (LR), gradient-boosted decision trees (GBDT), and support vector machines (SVM) for predicting imminent clinical deterioration based on cross-sectional patient data extracted from EPR. Our outcome of interest is a composite of in-hospital mortality and admission to critical care to represent severe and time-sensitive medical conditions requiring intervention. We ensure the models’ outputted probabilities are well-calibrated and reliable to fit into existing frameworks for assessing clinical utility \cite{Calster18}. Rather than prescribe a specific threshold for classifying high-risk cases, we measure our models’ discriminative skill across sensitivities via precision-recall curves and through their daily alert rate, which expresses how they would operate when deployed. We compare our performance against NEWS2, the preferred EWS in the United Kingdom \cite{Williams22}.

An extant practical challenge we address is models not generalising to new application environments due to structural differences compared to the development environment \cite{Taylor16, Challen19}. Solutions with rigid data requirements unrealistically require providers to conform to a specific pattern of testing or treatment to produce all the requisite data correctly \cite{Taylor16}. To avoid making assumptions about data availability or its collection context (such as timing, reliability, or frequency), we conduct experiments using different sets of predictive features that providers might generate under their unique clinical workflow. Starting with vital signs, we gradually construct models with finer information, including manual observations, laboratory results, clinical notes, and service utilisation, to reveal the most influential features.

A further barrier is a requirement for models supporting the clinical workflow to be transparent, safe, fair, and traceable in their decision-making process \cite{Meyer18, Holzinger17}. Machine learning models have conventionally operated as 'black boxes' \cite{Velez17}, obscuring their internal reasoning and biases \cite{Holzinger17, Barocas16}. Advances in interpretable machine learning and fairness-aware modelling allow us to address this. We incorporate methods from the fair machine learning literature \cite{Mehrabi21, Barocas19} into our evaluation framework to ensure our constructed models do not exhibit unfair bias against individuals or protected demographic groups. Then, we utilise Shapely Additive exPlanations \cite{Lundberg18}, a recently popularised model-agnostic framework for interpreting predictive models, to produce justifications for our models’ risk predictions on the individual patient level. These justifications reveal the best-performing models' internal reasoning and allow us to examine and validate the relationships between the significant predictors and the outcome. In addition to predicting a patient’s risk, our interpretable models can justify their prediction to the user by isolating the relevant characteristics of the patient that led them to that result \cite{Shawi20, Lundberg18}.

\section*{Results}

Our selected data comprised $121,058$ presentations to the acute medical unit (AMU) at Salford Royal Hospital, Manchester, UK, corresponding to $62,162$ distinct patients over the study period of January 2015 to March 2022. We identified $8,341$ critical deterioration events, of which $2,893$ occurred within 24 hours after admission. The Methods Table \ref{tab:data_summary} summarises the dataset and presents the stratification of samples across the three data subsets we used in our analysis: we partitioned the samples chronologically 2:1 into a model development set and a validation set, and we additionally considered an 'unseen' subset of the validation set that excludes the $8,139$ patients ($13.09\%$, making up $42.24\%$ of the validation set's records) that had prior admission records in the training set. The rates of critical care admission, mortality, and composite critical deterioration were uniform across the chronological split.

\begin{figure}[htbp]
    \centering
    \includesvg[inkscapelatex=false, width=\linewidth]{images/ap_bar.svg}
    \includesvg[inkscapelatex=false, width=\linewidth]{images/auc_bar.svg}
    \caption{Average precision \textbf{(a)} and Area under receiver operating curve \textbf{(b)} achieved by the best predictive models per learning algorithm across tested sets of data features. Each group corresponds to independent models trained with the indicated feature set concatenated to all the previous feature sets to its right. The error bars represent $95\%$ bootstrapped confidence intervals. Obs: Supplemental observations \& phenotype, Labs: Laboratory results, Notes: Clinical notes, Services: Triage \& service utilisation. We detail the contents of the feature sets in Methods Table \ref{tab:feature_sets}.}
    \label{fig:score_barplots}
\end{figure}

We compared numerous modelling pipeline variations as described in the Methods section. From this comparison, we identified LightGBM, a variant of GBDT, as the best-performing learning algorithm overall and logistic regression with L2 penalty (LR-L2) as the best linear model. We summarise their performance in Table \ref{tab:summary_results}. Figure \ref{fig:score_barplots} compares the average precision (AP) and area under the receiver operating curve (AUROC) of the best predictive models across classifier types on the complete validation set against the measured performance of the reference model (NEWS2) on this patient cohort. The groups in each plot correspond to incrementally augmenting the training data - the leftmost groups of each section present models using only vital signs as predictors, and subsequent groups give the results when we concatenated the indicated feature set (as described in Methods Table \ref{tab:feature_sets}) to the previous training inputs. We test these sets of features in order of 'centre-specificity', so that the most clinically standardised predictors, such as vital signs, are considered first. We provide the actual measurements with bootstrapped confidence intervals and the performance on the 'unseen' validation set in Supplementary Tables 7 and 8.

\begin{figure}[htbp]
    \centering
    \includesvg[inkscapelatex=false, width=\linewidth]{images/alert_pr_all.svg}
    \includesvg[inkscapelatex=false, width=\linewidth]{images/alert_pr_lightgbm.svg}
    \caption{Alert Rate vs Sensitivity \textbf{(a,c)} and Precision-Recall curves \textbf{(b,d)}. Top Row \textbf{(a,b)}: All learning algorithms trained on the complete feature set (equiv. "\& Services"). Bottom Row \textbf{(c,d)}: GBDT (LightGBM) across feature sets (concatenated incrementally). In \textbf{(a,c)}, the Alert Rate curve plots the arithmetic mean of daily positive predictions (alerts) across the validation period for a given sensitivity value (y-axis) against that sensitivity value (x-axis). The point where two lines intersect corresponds to the maximum achievable sensitivity for which the model with the lower line maintains a lower daily alert rate than the model with the upper line. In \textbf{(b,d)}, the Precision-Recall (PR) curve presents the positive predictive value (PPV, or precision) on the y-axis against sensitivity on the x-axis. On the PR curve, an unskilled model giving random outputs would yield a horizontal line at $y=P/(P+N)$, where $P$ and $N$ are the numbers of positive and negative samples in the data, respectively, while a theoretical 'perfect' model would yield a single point $(1,1)$ in the upper-right corner of the plot. The curves are plotted from each model's outputted predictions for the complete validation set.}
    \label{fig:alert_pr_curves}
\end{figure}

\begin{table}[htbp]
    \renewcommand{\arraystretch}{1.2}
    \centering
    \begin{tabular}{lll|ccccc}
        \toprule
        \textbf{Metric}                 & \textbf{Estimator}                 & \textbf{Dataset}  & Vitals  & \& Obs  & \& Labs & \& Notes & \& Services  \\

        \midrule
        \multirow{4}{*}{\textbf{AP}}    & \multirow{2}{*}{\textbf{LR-L2}}    & \textbf{Complete} & $0.157$ & $0.258$ & $0.259$ & $0.264$  & $0.466$      \\
                                        &                                    & \textbf{Unseen}   & $0.172$ & $0.309$ & $0.305$ & $0.312$  & $0.478$      \\
        \cline{2-8}
                                        & \multirow{2}{*}{\textbf{LightGBM}} & \textbf{Complete} & $0.179$ & $0.314$ & $0.314$ & $0.323$  & $\bm{0.513}$ \\
                                        &                                    & \textbf{Unseen}   & $0.206$ & $0.370$ & $0.365$ & $0.374$  & $\bm{0.523}$ \\
        \cline{1-8}
        \multirow{4}{*}{\textbf{AUROC}} & \multirow{2}{*}{\textbf{LR-L2}}    & \textbf{Complete} & $0.812$ & $0.836$ & $0.838$ & $0.851$  & $0.894$      \\
                                        &                                    & \textbf{Unseen}   & $0.823$ & $0.843$ & $0.845$ & $0.854$  & $0.896$      \\
        \cline{2-8}
                                        & \multirow{2}{*}{\textbf{LightGBM}} & \textbf{Complete} & $0.835$ & $0.868$ & $0.888$ & $0.891$  & $\bm{0.922}$ \\
                                        &                                    & \textbf{Unseen}   & $0.842$ & $0.879$ & $0.896$ & $0.897$  & $\bm{0.923}$ \\
        \bottomrule
    \end{tabular}
    \caption{\label{tab:summary_results} Summary of model performance. Average precision (AP) and Area under receiver operating curve (AUROC) of GBDT (LightGBM) and logistic regression with L2 penalty (LR-L2) for predicting 24-hour critical deteriorations on the two validation sets: 'Complete', the full validation set, and 'Unseen', which includes only patients who had no admissions in the training dataset. Each column corresponds to independent models trained with the indicated feature set concatenated to all the previous feature sets to its right.}
\end{table}

Data-driven modelling matched or outperformed the reference model across all feature sets, with the complete feature set (rightmost group in each section of Figure \ref{fig:score_barplots}) giving the best performance. Both AP and AUROC trended upward as the number of predictors grew, though phenotype and supplemental observations ("\& Obs"), laboratory results ("\& Labs"), and clinical notes ("\& Notes") had a greater impact on the average precision while the AUROC remained more stable. Including triage and service utilisation ("\& Services") yielded the largest singular boost in AP (increase from $0.324 \rightarrow 0.513$ for GBDT). Figure \ref{fig:alert_pr_curves} illustrates the alert rate vs sensitivity and precision-recall curves for GBDT across different feature sets and for all classifier types trained on the complete feature set. GBDT produced fewer alerts per day on average compared to the reference model up to very high sensitivities ($0.966$), and all classifiers maintained an improved alert rate up to moderately high sensitivities ($>0.80$). The largest reduction of alert rate was at sensitivity $0.864$, where GBDT yielded $9.869$ daily alerts, $18.50\%$ less than NEWS2's $12.110$. The positive predictive value (PPV) of GBDT-Vitals behaves similarly to the reference model as we vary sensitivity. Performance was stable between the "Complete" and "Unseen" validation sets, with a median increase of $0.054$ for AP and $0.012$ for AUROC when the 'known' patients were removed, as shown in Supplementary Tables 7 and 8. All models had satisfactory calibration, though with a tendency to underestimate the probability of critical deterioration, as illustrated in Supplementary Figure 8.

\begin{figure}[tb]
    \centering
    \includegraphics[width=\linewidth]{images/shap_lgbm.png}
    \caption{Induced feature importances for GBDT (LightGBM) in decreasing order of mean absolute impact. In \textbf{(a)}, the bar lengths represent the mean absolute impact of each feature on the model's predictions for the validation set. In \textbf{(b)}, each point represents a value from one admission record. The points' colour corresponds to numerical value, and their position on the x-axis represents the magnitude of their contribution towards increasing the predicted risk (if $x>0$) or reducing it (if $x<0$).}
    \label{fig:shap_lightgbm}
\end{figure}

The feature interactions induced by SHAP for GBDT yielded further insight into the predictors' relationship with our clinical outcome. Figure \ref{fig:shap_lightgbm} (a) ranks all the included predictors by their mean absolute impact towards positive predictions (deterioration) and negative ones (no deterioration) across the validation set, (b) illustrates the patient-individual impact of each feature, and Supplementary Figure 10 breaks down the relative impact of the values taken by categorical data features. The presenting complaint ranked the highest and contributed similarly towards positives ("diabetes", "GI bleeding") and negatives ("back pain", "facial problems"). The model captured a non-linear relationship between risk and indicators of kidney function, such as creatinine and urea levels, which is consistent with clinical findings differentiating the mortality risk of acute kidney injury versus chronic disease \cite{James20}. Triage decisions were heavily influential, with same day emergency care (SDEC) invariably reducing the estimated risk, while certain clinical specialities, such as respiratory medicine, geriatric medicine, and general medicine (a catch-all for non-specialty cases), strongly contributed towards positives.

Similarly, we record the coefficients of the logistic regression models in Supplementary Tables 9 and 10 and find them to be consistent across the penalised models. SDEC, higher sodium levels, and specific presenting complaints (e.g., "facial problems", "ear problems") reduce the estimated risk. Conversely, elevated respiratory rate, potassium levels, requiring a bed, and certain clinical specialities and breathing devices yield increased risk estimates. It is interesting to notice that age is assigned a negative coefficient. Figure \ref{fig:shap_lightgbm} reveals that GBDT also identified age as a strong predictor, with advanced age driving the model towards negative predictions rather than positive ones. We explore this non-intuitive and potentially spurious association in Figure \ref{fig:age_temperature}(a,b) which compares the two models' patient-individual SHAP values for the age feature. We theorise this relationship is partly due to high-frailty patients (aged $\geq 80$ years), having the lowest proportion of 24-hour critical deterioration events out of all age groups (as shown in Supplementary Figure 7) despite being very frequent attendees at the ED.

\begin{figure}[htbp]
    \centering
    \includesvg[inkscapelatex=false, width=0.8\linewidth]{images/shap_scatter_age.svg}
    \includesvg[inkscapelatex=false, width=0.8\linewidth]{images/shap_scatter_temperature.svg}
    \caption{Feature-specific importances extracted by SHAP from GBDT \textbf{(a,c)} and LR-L2 \textbf{(b,d)}. The top section \textbf{(a,b)} presents the importances of patient age, while the lower section \textbf{(c,d)} presents body temperature. Each point represents a value from one validation set record. The points' position on x-axis represents the numerical feature value, while the y-axis indicates their contribution to the prediction for that patient, with values above $y=0$ (indicated in red) contributing towards making the prediction positive and values below $y=0$ (indicated in blue) contributing towards making the prediction negative.}
    \label{fig:age_temperature}
\end{figure}

\begin{figure}[htbp]
    \centering
    \includesvg[inkscapelatex=false, width=\linewidth]{images/raw_vs_scored_ap.svg}
    \caption{Average precision (AP) of \textbf{(a)} LR-L2 and \textbf{(b)} GBDT. Each pair of bars corresponds to incrementally including the indicated feature sets (from Methods Table \ref{tab:feature_sets}) as training data. For a given feature set, we measure the AP of two independently trained models, one using the direct measurements of vital signs (blue), and one with the vital signs encoded using the NEWS2 severity scales (red). The error bars represent $95\%$ bootstrapped confidence intervals.}
    \label{fig:raw_vs_scored}
\end{figure}

As an additional test, we trained logistic regression and GBDT models with vital signs encoded into integers $0-3$ per the NEWS2 severity scales \cite{RCP17}. We compared the results with the classifiers' performance when using the original vital sign values to investigate how each model type captures the non-linear relationship between vitals and clinical outcomes in Figure \ref{fig:raw_vs_scored}. We observe that the 'handcrafted' scales boosted the performance of logistic regression across feature sets, while GBDT's performance either dropped or remained stable. Figure \ref{fig:age_temperature}(c,d) presents an example of a diverging relationship learned by GBDT and LR from the same feature, temperature. Note that the presented results thus far assume 24 hours after admission as the cut-off point for identifying deterioration events. Supplementary Figure 9 illustrates how the AUROC of GBDT and LR-L2 varied when we increased the (cumulative) time threshold gradually from 24 hours to 30 days. Across all feature sets, the AUROC trended downwards as the cut-off widened and the on-admission measurements for each newly included sample became more distant from the outcome.

Finally, Figure \ref{fig:entropy} presents the generalised entropy index vs sensitivity for GBDT across the tested feature sets and all models trained on the complete feature set. Supplementary Figure 11 isolates the between-group fairness component of the generalised entropy index when we consider the population groups defined by the protected demographic characteristics of age group and sex (as specified in Supplementary Figure 7). All models except for GBDT-Vitals achieve an improved fairness score compared to the reference model across sensitivity thresholds. NEWS2 produces a better between-group fairness and, correspondingly, a more significant unfairness within the demographic groups, under the complete feature set above sensitivities of $\sim0.82$. To account for potential pre-existing inequalities in the cohort, we record the differential bias amplification of the models in Supplementary Table 11. These measurements corroborate the generalised entropy findings, with a positive bias amplification under the vital signs feature set when considering age groups. However, this diminishes when considering intersectional protected groups of both age and sex. Bias amplification values across all other feature sets are strongly negative - indicating removal of bias - or near zero. We theorise that this unusual amplification of inequality with respect to age groups is due to the vital signs feature set containing insufficient information to predict our tracked outcome correctly for patients of all ages.

\begin{figure}[htbp]
    \centering
    \includesvg[inkscapelatex=false, width=\textwidth,height=0.9\textheight,keepaspectratio]{images/entropy.svg}
    \caption{Generalised Entropy vs Sensitivity curves of \textbf{(a)}: GBDT across the tested feature sets, and \textbf{(b)}: All classifier types trained on the complete feature set. We plot each model's generalised entropy index for a given sensitivity value (y-axis) against that sensitivity value (x-axis). A lower value on the y-axis indicates a more fair distribution of 'benefit', i.e. of receiving a positive prediction. A theoretical 'perfect' model would yield a single point $(0,1)$ in the lower-right corner of the plot. }
    \label{fig:entropy}
\end{figure}

\section*{Discussion} In a large cohort of ED admissions, we developed and validated predictive models that can differentiate patients likely to deteriorate shortly after admission. GBDT methods received the most focus as they are state-of-the-art for sparse classification tasks (even compared to deep neural networks \cite{Borisov21}), they can capture non-linear interactions such as those present in clinical data, and they natively incorporate missing values, which are inevitable under typical clinical workflows. Using our trained models' coefficients and the extracted global justifications, we can identify which characteristics of our cohort were most predictive of the tracked clinical outcome both on the patient level and across the studied population. Features that encode the clinical context of the patient's condition, presentation, and comorbidities stood out as the most useful. These included presenting complaints, triage decisions such as the utilisation of SDEC, and the assigned clinical speciality, among others. Patient age stood out for being inversely correlated with our tracked outcome, against clinical intuition \cite{Metlay97}, which we theorise results from the low prevalence of the outcome within the highest age band. While it did not result in the model amplifying unfair bias, it presents a clear example of model interpretability revealing spurious associations that might require correcting prior to deployment.

This study differs further from prior literature in using as representative a data sample as possible and in prioritising practical concerns. Our cohort, with patients of varied acuity and conditions, reflects a typical real world ED acute medical workload. Frontline staff collected the patient data under everyday conditions, where operational pressures affect the timeliness and reliability of data entry. We excluded little data since, although comprehensive manual data curation is helpful for model development, it conflicts with scalable deployment and real-time use of data-driven systems \cite{Meyer18} and can lead us to discard valuable information for uncommon cases \cite{Rajkomar18}. We did not carry out a priori feature selection but instead used all available data and employed modelling methods that perform intrinsic feature selection and can differentiate useful features based on evidence. Healthcare digitalisation is an ongoing process \cite{NHS19}, so we made no assumptions about the level of EPR integration. Instead, through our experiments with different feature sets, we accommodate different levels of data availability. The lack of a standardised benchmark dataset makes direct comparisons between studies on this topic challenging, so we minimised centre-specific assumptions and standardised our modelling pipelines' structure to establish reproducibility.

We similarly designed our assessment methodology around the extant practical challenges and presented results with the context of their resource cost. We used a temporal split of the study data to assess performance but retained the records where the patient had presented to the same ED during the training period as frequent repeat attendees reflect the reality of clinical practice. To strengthen our results, however, we also examined removing these records and still demonstrated good performance. Calibration is often underappreciated \cite{Gerry20}, and alert frequency deserves attention as alert fatigue is a key critique aimed at existing solutions from frontline staff \cite{Obrien20}. We focused on measuring discriminative skill and avoided setting a threshold for positive or negative classifications, as setting it carries clinical, operational, and ethical complications. Directing care where it is needed promptly is vital and far outweighs the cost of false positives. However, excessive false alarms are detrimental to a model’s utility due to alert fatigue \cite{Ancker17, Kolic15, Bedoya19}. Balancing clinical risk against available capacity is a well-researched problem beyond the scope of our study \cite{Calster18, Azcarate20}; instead, we argue that early-stage researchers should aim to maximise the discriminative skill of their model, as might be measured by AUROC or the highest achievable sensitivity while preserving acceptable specificity.

Our observational dataset is limited to one acute secondary care centre, but many measured parameters and outcomes vary between providers. Even near-universal predictors such as vital signs may be measured differently. For example, manual measurement of respiratory rate is less precise than an electronic recording\cite{Kellett11}, provision of supplemental oxygen is subjective and depends on operational constraints, availability, guidelines, and expertise \cite{Kasereka20}, and the same oxygen saturation may represent different levels of clinical risk depending on whether it was measured before or after commencing oxygen \cite{Cuthbertson07}. Furthermore, we recorded symptoms, vital signs, and laboratory results from the point of admission. This information gives a cross-sectional view of the patient’s condition as seen by the admitting clinician but excludes longitudinal information, which prior work has collected via continuous vital sign monitoring and used to train highly effective models \cite{Kao21, Lundberg18}. Finally, we investigated unfair bias and group inequalities in the models to the best of our ability but limited our assessment to the available protected characteristics. While patients face divergent clinical risks depending on characteristics such as sex, age, or ethnic background \cite{Iezzoni13C3}, finer data such as economic stability, education, community context, and other social determinants of health are also strong predictors of clinical risk \cite{Mahmoudi20}. We recommend that researchers investigate fairness thoroughly, especially if the models they construct are intended to autonomously screen or prioritise patients’ access to care, to ensure healthcare inequalities are not perpetuated \cite{Yu19}.

There are key considerations researchers should take into account before adopting similar modelling methodologies. It is essential to consider the validity of jointly modelling outcomes and the reliability of any composite outcome as a surrogate for clinical deterioration. We considered critical care admission and mortality as a single outcome because we expect both to be preceded by deranged physiology, and the clinical response to both, in terms of urgency and skill, is similar \cite{Smith13}. The joint outcome served as a surrogate for any severe and time-sensitive medical condition encountered at the ED; this is a common modelling choice in the literature \cite{Gerry20, Levin18} and one we find reasonable, as our focus is on clinical escalation, which is the primary purpose of an EWS \cite{Morgan07}. However, critical care and mortality represent competing outcomes as the former intends to prevent the latter \cite{Wolbers09}. Future studies may prefer to avoid such assumptions and investigate multiclass modelling or compositing multiple binary classifiers, each trained to identify a single measurable outcome. Some features we utilised, such as triage outcomes, directly represent clinical decision-making. Their inclusion is in contrast with the 'one-size-fits-all' approach taken by the NEWS2 \cite{Inada18} or their explicit exclusion by some studies to avoid capturing and amplifying human-originated bias \cite{Meyer18}. If the purpose of a system is to 'sense-check' clinical decisions, its input data should ideally be as isolated as possible from those decisions. However, our findings show that these features efficiently stratify patient risk, making them valuable for producing reliable clinical risk estimates as long as the risks are made clear and considered.

In conclusion, we demonstrated the development of predictive models on a large, real-world sample of general ED patients. Considering the high and rising pressures EDs face and the potential for missed diagnoses, models built from continuing our work could be clinically valuable for decision support. We contend that this study demonstrates the power of machine learning for modelling or adapting to patient populations for this task. By incorporating modularised modelling pipelines from contemporary machine learning practice and leveraging the advances in interpretable modelling, we encourage future research to follow a systematised model-building approach and help obtain clinically useful prognostic tools.

\section*{Methods}
\subsection*{Data Collection and Preparation}

\begin{table}[htbp]
    \centering
    \renewcommand{\arraystretch}{1.2}

    \begin{tabular}{llrrrr}
        \toprule

        \textbf{Group}                                                     & \textbf{Variable}              & Total             & Training          & Test (Complete)  & Test (Unseen)    \\
        \midrule
        \multirow{3}{*}{\textbf{Episode}}                                  & \textbf{LOS (days)}            & 2.27 (0.65-7.13)  & 2.06 (0.63-6.68)  & 2.88 (0.71-8.49) & 2.20 (0.55-7.50) \\
                                                                           & \textbf{Patients}              & 62162             & 44752             & 25549            & 17410            \\
                                                                           & \textbf{Records}               & 121058            & 81108             & 39950            & 23075            \\
        \cline{1-6}
        \multirow{4}{*}{\textbf{Outcomes}}                                 & \textbf{30-day Mortality}      & 3967 (3.28\%)     & 2589 (3.19\%)     & 1378 (3.45\%)    & 692 (3\%)        \\
                                                                           & \textbf{Critical Care}         & 4012 (3.31\%)     & 2822 (3.48\%)     & 1190 (2.98\%)    & 716 (3.10\%)     \\
                                                                           & \textbf{Critical Events}       & 8341 (6.89\%)     & 5489 (6.77\%)     & 2852 (7.14\%)    & 1537 (6.66\%)    \\
                                                                           & \textbf{In-hospital Mortality} & 5124 (4.23\%)     & 3233 (3.99\%)     & 1891 (4.73\%)    & 965 (4.18\%)     \\
        \cline{1-6}
        \multirow{8}{*}{\textbf{Vitals}}                                   & \textbf{AVCPU-A}               & 119493 (98.71\%)  & 80013 (98.65\%)   & 39480 (98.82\%)  & 22827 (98.93\%)  \\
                                                                           & \textbf{Assisted Breathing}    & 12167 (10.05\%)   & 7835 (9.66\%)     & 4332 (10.84\%)   & 2245 (9.73\%)    \\
                                                                           & \textbf{NEWS}                  & 1 (0-2)           & 1 (0-2)           & 1 (0-2)          & 1 (0-2)          \\
                                                                           & \textbf{Pulse (beats/min)}     & 80 (70-90)        & 80 (70-90)        & 80 (70-90)       & 80 (70-90)       \\
                                                                           & \textbf{RR (breaths/min)}      & 17 (16-18)        & 17 (16-18)        & 18 (16-18)       & 17 (16-18)       \\
                                                                           & \textbf{SpO2 (\%)}             & 97 (96-98)        & 97 (96-98)        & 97 (96-98)       & 97 (96-98)       \\
                                                                           & \textbf{Systolic BP (mmHg)}    & 124 (113-139)     & 122 (112-138)     & 125 (114-140)    & 125 (114-140)    \\
                                                                           & \textbf{Temperature (oC)}      & 36.70 (36.40-37)  & 36.70 (36.40-37)  & 36.70 (36.40-37) & 36.70 (36.40-37) \\
        \cline{1-6}
        \multirow{8}{2cm}{\textbf{Supplemental Observations \& Phenotype}} & \textbf{Age (years)}           & 69 (50-82)        & 69 (50-82)        & 68 (50-81)       & 64 (44-79)       \\
                                                                           & \textbf{Diastolic BP (mmHg)}   & 70 (60-80)        & 70 (60-78)        & 70 (62-80)       & 70 (62-80)       \\
                                                                           & \textbf{Female}                & 63414 (52.38\%)   & 42768 (52.73\%)   & 20646 (51.68\%)  & 11571 (50.15\%)  \\
                                                                           & \textbf{FiO2 (\%)}             & 0 (0-0)           & 0 (0-0)           & 0 (0-0)          & 0 (0-0)          \\
                                                                           & \textbf{Lying Down}            & 56565 (46.73\%)   & 36197 (44.63\%)   & 20368 (50.98\%)  & 11368 (49.27\%)  \\
                                                                           & \textbf{Nausea}                & 1934 (1.60\%)     & 1407 (1.73\%)     & 527 (1.32\%)     & 278 (1.20\%)     \\
                                                                           & \textbf{Pain}                  & 18504.0 (15.29\%) & 13308.0 (16.41\%) & 5196.0 (13.01\%) & 3227.0 (13.98\%) \\
                                                                           & \textbf{Vomiting}              & 607 (0.50\%)      & 417 (0.51\%)      & 190 (0.48\%)     & 109 (0.47\%)     \\
        \cline{1-6}
        \multirow{5}{*}{\textbf{Labs}}                                     & \textbf{Creatinine (mmol/L)}   & 79 (63-105)       & 78 (63-103)       & 79 (64-106)      & 78 (63-100)      \\
                                                                           & \textbf{Haemoglobin (g/L)}     & 130 (114-143)     & 129 (114-142)     & 130 (114-143)    & 132 (117-145)    \\
                                                                           & \textbf{Potassium (mEg/L)}     & 4.20 (3.90-4.50)  & 4.20 (3.90-4.50)  & 4.20 (3.90-4.50) & 4.20 (3.90-4.50) \\
                                                                           & \textbf{Sodium (mmol/L)}       & 138 (135-140)     & 138 (135-140)     & 138 (135-140)    & 138 (135-140)    \\
                                                                           & \textbf{Urea (mmol/L)}         & 6.40 (4.60-9.60)  & 6.30 (4.60-9.40)  & 6.40 (4.60-9.70) & 6 (4.50-8.90)    \\
        \cline{1-6}
        \multirow{2}{2cm}{\textbf{Service Utilisation}}                    & \textbf{30-day Readmission}    & 15387 (12.71\%)   & 10732 (13.23\%)   & 4655 (11.65\%)   & 1919 (8.32\%)    \\
                                                                           & \textbf{SDEC}                  & 29096 (24.03\%)   & 21020 (25.92\%)   & 8076 (20.22\%)   & 5719 (24.78\%)   \\
        \bottomrule
    \end{tabular}

    \caption{\label{tab:data_summary} Summary statistics of the study sample. Numerical patient characteristics of acute medical unit admissions, chronologically partitioned into training and testing sets. "Test (Unseen)" corresponds to the chronologically split validation set but excluding patients who had any prior admissions in the training set. Binary variables are reported as "number of positives (\%)", while numerical variables are reported as quartiles.}
\end{table}

\paragraph{Study Setting.} Salford Royal Hospital is a digitally mature, 'paper light' NHS secondary care hospital with over $100,000$ emergency department attendances and $\sim 40,000$ unplanned admissions annually. The Hospital's EPR captures clinical episode data in real-time from arrival at the emergency department until discharge. Selected data are exported pseudonymously to an internal data warehouse to drive local quality improvement and service development projects. Our study considered all such records from 1st January 2015 to 31st March 2022. This starting date reflects the first calendar year after the introduction of electronic NEWS recording in the Hospital. We selected all patients aged $\geq 18$ years admitted to the Acute Medical Unit (at Salford, known as the Emergency Admissions Unit, EAU). Our data include patients who received ambulatory emergency care (AEC) and same-day emergency care (SDEC \cite{NHS18}) but exclude planned admissions and day case reviews. We present summary statistics of the dataset in Table \ref{tab:data_summary}, and further details of the collected categorical features in Supplementary Table 6.

\paragraph{Data Collection.} As a routine part of EAU admission, the responsible staff member (nurse or support worker) records the patient's vital signs within a target of 30 minutes of arrival. The vital signs that make up the NEWS2 \cite{RCP17} are measured in a standardised manner using Dinamap monitors, and manually transcribed into EPR. These data are body temperature ($^{\circ}C$), heart rate (beats/min), systolic (and diastolic) blood pressure (mmHg), and peripheral oxygen saturation ($\%$). Other parameters are measured using manual observation and direct questions. These are the patient's level of consciousness (AVCPU), presence of pain, nausea, or vomiting, whether the patient was receiving oxygen at the time of SpO2 measurement and, if applicable, the oxygen flow rate and mode of delivery.

Independent of this, blood test results are automatically recorded in the laboratory information management system (LIMS) and copied to EPR in real time. Whether a patient receives routine blood tests depends on operational pressures and considerations at the ED, not on the patient's presentation. Other information available upon arrival at the EAU includes identifier data such as the unique patient number; basic phenotypic information, such as their age and sex; admission pathway (e.g., ED, emergency GP referral); arrival time; and unstructured notes indicating their presenting complaint and the ED staff's primary diagnosis. For patients with prior hospital visits, significant comorbidities and previous admission events are available from the point of admission.

\paragraph{Retrospective Data Augmentation.} Following initial collection, our data are supplemented with downstream administrative and outcome information, including ICD-10, OPCS-4, and HRG codes, alongside service utilisation records, 30-day readmission, and community mortality events. The date and time of discharge, the total length of stay (LOS), the wards the patient was admitted to (in chronological order), and the LOS per ward are collated upon the patient’s discharge. Each record includes up to seven concurrent diagnoses, represented by 4-7-character alphanumeric codes per the ICD-10-CM standard. These diagnoses are compiled after discharge by a clinical coding team based on the existing codes and clinical notes recorded in EPR. Procedures and service utilisation are similarly recorded in detail in EPR and coded in retrospect using the OPCS-4 standard.

\paragraph{Feature Engineering.} Some of the collected data is not directly clinically relevant or may be unsuitable for modelling under a realistic use case. However, we can use it to engineer useful features or delineate subpopulations in the cohort for more detailed model evaluation. Other features are relevant but first require cleaning or modification. We derive the following features:
\begin{itemize}
    \item 30-day Readmission. We mark as readmissions those patient records that are preceded by a record bearing the same unique patient ID if the two records’ admission dates are $\leq$ 30 days apart.
    \item Unstructured Clinical (ED) Notes. The presenting complaint and ED diagnosis are unstructured text and thus could hold any string value. We cluster presenting complaints into a categorical variable representation since the 50 most frequent values account for nearly all records ($97.06\%$), and we assign the remainder a sentinel value. In contrast, the ED diagnosis varies greatly between records, so we compile a list of clinically relevant word stems and abbreviations based on expert opinion and construct a boolean Bag-of-Words vector for each record indicating which ones are present. We provide the prevalent presenting complaint values and diagnosis stems in Supplementary Table 6.
    \item Vital Signs. We investigate training models directly on vital sign readings or encoding them into integers $0-3$ per the NEWS2 severity scales \cite{RCP17}. The former approach forces models to form evidence-based weightings for values that correlate with adverse patient outcomes, while the latter allows us to incorporate the domain knowledge embedded in the NEWS2 into the models. Recorded vitals must be checked for spurious values as they are the only parameters transcribed into EPR manually under a typical workflow. We check each record against fixed ranges (e.g., 0-100$\%$ for SpO2) and soft thresholds based on the range of physiologically possible values determined by expert clinical opinion. We provide further details on filtering these values in Supplementary Table 4.
\end{itemize}

\paragraph{Data Labelling.} Our tracked outcome is a composite of in-hospital mortality or admission to critical care from the ward within a specified time threshold after presenting to the ED. The criteria to identify patient episodes that belong in the positive class are:
\begin{enumerate}
    \item The episode is not excluded by our filtering criteria, i.e. they have at least recorded vital signs, are $\geq 18$ years old, are not pregnant, and did not receive critical care interventions on-arrival.
    \item The discharge/end-of-episode record indicates the patient died in the hospital AND the record's timestamp is within 24 hours of the admission timestamp, OR their service utilisation indicates admission to critical care or provision of critical interventions on the ward AND this occurred within 24 hours of the admission timestamp.
\end{enumerate}
We identify critical care based on recorded admission into the hospital's critical care unit (CCU) or the high-dependency medical unit (H1). We use the length-of-stay per ward to determine how long after the patient's arrival they were admitted to critical care. A smaller subset of patients received critical care interventions without being moved to these wards, and we can detect most such cases through specific entries in their recorded procedures - OPCS-4 codes E85.1 (invasive ventilation), X50.3 (advanced cardiac pulmonary resuscitation), X50.4 (external ventricular defibrillation), or X56.* (intubation of the trachea).

\subsection*{Model Development.}
\paragraph{Modelling Pipeline.} We adopt a modularised model-building approach from contemporary machine learning practice. We consider \emph{pipelines} as sequences of distinct tasks in the model-building process, where each task's output becomes the subsequent task's input. Some tasks modify the data samples in preparation for modelling. At least one task in each pipeline is a learning/model-building algorithm. Then, subsequent post-processing tasks may alter the predictive model's output or aggregate multiple models. We implement the following tasks, executed in order:

\begin{enumerate}
    \item Data Pre-processing. Executes the data preparation tasks outlined previously to produce a vector representing each patient episode. We parameterise the processing component to include only the features we specify, so we may investigate selectively including features and the impact they have on performance. The sets of features we consider are listed in Table \ref{tab:feature_sets}.

    \item Data Splitting. Partitions the data into two subsets; we use one for model construction and reserve the second for validation. We prefer a temporal train-test split over standard random splitting \cite{Altman00}, and partition the dataset such that the first $2/3$ of records chronologically serve as the training set and the latter $1/3$ as the validation set. For some experiments we implement an additional filter that excludes any validation set records where the patient, as identified by their unique ID, had also appeared in the training set in a previous admission.

    \item Data Imputation. Supplements standard values into data samples with empty fields. We apply this only to those modelling algorithms that are incompatible with missing data in their inputs (logistic regression). We impute numerical features with the median over the training dataset and binary and categorical variables with appropriate constant values. The imputed values correspond to a patient in stable condition.

    \item Model Construction. A learning algorithm receives the data samples and produces a predictive model.
    \item Calibration. As a post-processing step, we map the numerical outputs of the trained predictive model into well-calibrated probabilities, substituting the model's original output $C(\bm{x_i})$ on input $\bm{x_i}$ for an estimate of $Pr(y_i=1|C(\bm{x_i}))$, the conditional probability of belonging to class $y_i$. We opt for isotonic calibration \cite{Niculescu05} and fit a meta-estimator that learns the isotonic (monotonically increasing) mapping $m$ that minimises a loss function $\pazocal{L} = \sum_i w_i(y_i - m(C(\bm{x_i}))^2$.
\end{enumerate}

\begin{table}[htbp]
    \centering
    \renewcommand{\arraystretch}{1.2}
    \begin{tabular}{p{0.2\linewidth} | p{0.75\linewidth}}
        \toprule
        \textbf{Feature Set}                    & \textbf{Features (Units)}                                                                                                                                                                                           \\ \midrule
        \textbf{Vital signs (NEWS2)}            & Body temperature ($^{\circ}C$), heart rate (beats/min), systolic blood pressure (mmHg), peripheral oxygen saturation ($\%$)                                                                                         \\
        \textbf{Supplemental Obs. \& Phenotype} & Sex (M/F), Age (years), Diastolic blood pressure (mmHg), breathing device (if applicable), prescribed oxygen (FiO2), presence of pain (Y/N), presence of nausea (Y/N), presence of vomiting (Y/N), lying down (Y/N) \\
        \textbf{Clinical Notes}                 & Presenting complaint (text), ED diagnosis (text)                                                                                                                                                                    \\
        \textbf{Laboratory Results}             & Haemoglobin (g/L), urea (serum, mmol/L), sodium (serum, mmol/L), potassium (serum, mEq/L), creatinine (mcmol/L)                                                                                                     \\
        \textbf{Service Utilisation}            & Triaged to SDEC (Y/N), readmission within 30 days (Y/N), admission speciality (category), admission pathway (category)                                                                                              \\
        \bottomrule
    \end{tabular}
    \caption{Dataset features and units categorised into feature sets. In the given units, "Y/N" indicates binary variables, "category" un-ordered categorical variables, "text" unstructured text data, and "M/F" indicates male or female.}
    \label{tab:feature_sets}
\end{table}

\paragraph{Model Training and Tuning.} We construct pipelines with each combination of available components. For each one, we execute a single-objective Bayesian optimisation process (Tree-Structured Parzen approach \cite{Bergstra11}) to sweep over the space of possible hyperparameter values and probabilistically settle on values that maximise our chosen performance metric, average precision. We construct the final models using the best-scoring hyperparameters after $1000$ tuning iterations. We report the resultant hyperparameters in Supplementary Table 5. We avoid training the calibration meta-estimator on the same data that trained the classifier and, instead, we combine calibration with k-fold cross-validation. We randomly separate the training dataset into $k$ equal-sized partitions (setting parameter $k=5$), train a model on four of the subsets and fit the calibrator using the remaining subset. We iteratively repeat this $k$ times to such that each partition serves as the calibration set once and produce $k$ independent models to serve as sub-estimators of a model ensemble. The final 'representative' probability prediction of the ensemble $C$ of sub-estimators $C_1, \cdots, C_k$ for input vector $\bm{x}$ is taken to be the arithmetic mean of the sub-estimators' predictions: $C(\bm{x}) = \frac{1}{k} \sum^k_{i=1} C_{i}(\bm{x})$.

\paragraph{Model Evaluation.} We assess the discriminative skill of the models by constructing the precision-recall curve and measuring the average precision, which is the mean of the PPV (or precision) over the interval of sensitivity (TPR/recall) values from 0-1. We approximate this with the weighted mean of the measured PPV across the observed sensitivity thresholds, where the weight of each element is the difference in sensitivity from the previous element \cite{Boyd13}.
\begin{equation*}
    AP = \int_0^1 p(r)dr \approx \sum^n_{k=1} P(k)\Delta r(k)
\end{equation*}
where $p(r)$ is the PPV as a function of sensitivity $r$, $P(k)$ is the precision at cut-off $k$ in the ranked sequence of data samples in the validation dataset, and $\Delta r(k)$ is the difference in recall $r_k - r_{k-1}$. We calculate the confidence intervals for our estimate of the AP by bootstrapping with $1000$ bootstrap samples over the validation set \cite{Kohavi95}. We construct the PR curve by plotting the PPV on the y-axis against sensitivity on the x-axis \cite{Davis06}. On the PR curve, an unskilled model giving random outputs would yield a horizontal line at $y=P/(P+N)$, where $P$ and $N$ are the numbers of positive and negative samples in the data, respectively, while a theoretical 'perfect' model would yield a single point $(1,1)$ in the upper-right corner of the plot.

We construct the receiver-operating characteristics (ROC) curve and compute the area under the receiver operating curve (AUROC). We plot the false-positive rate (1 minus the specificity) on the x-axis against the sensitivity on the y-axis. The minimum possible area under the curve is $0.5$, corresponding to a completely random relationship between the model’s output and the ground truth. Generally, $0.7-0.8$ indicates reasonable discrimination, and values over $0.8$ indicate good discrimination \cite{Smith13}. We compute confidence intervals for the AUROC as before.

The ROC and PR curves both provide a model-wide evaluation and, while the ROC curve is more common, we prefer the PR curve because it better indicates the skill of the model at predicting the minority (positive) class correctly and is less influenced by predicting the majority (negative) class correctly \cite{Saito15}. The PR curve further allows us to visually inspect how quickly PPV deteriorates as we increase model sensitivity \cite{Davis06}, which is helpful in a task where it may be appropriate to value sensitivity over specificity.

Finally, we investigate how a model’s daily alert rate varies with sensitivity \cite{Romero21}. We construct an alert rate curve by plotting the alert rate (the number of positive predictions divided by the number of days) on the y-axis over sensitivity on the x-axis. The point where two lines intersect corresponds to the maximum achievable sensitivity for which the model with the lower line maintains a lower daily alert rate than the model with the upper line.

%\paragraph{Model Interpretation} Todo.

\paragraph{Model Bias} We investigate two forms of undesirable bias: individual, representing how dissimilarly we treat individuals who deserve similar outcomes \cite{Dwork12}, and group-based, measuring the inequality of predictions between demographic groups defined by protected characteristics \cite{Binns20}. The generalised entropy index\cite{Speicher18} applies to both notions concurrently. Given a patient record $\bm{x}_i$ with ground-truth outcome $y_i$, we define the \emph{benefit} experienced by the patient due to model prediction $C(\bm{x}_i)$ as:
\begin{equation*}
    b_i = y_i - C(\bm{x}_i) +1
\end{equation*}
Under this representation, a false-positive patient experiences a large benefit ($b=2$), while a false-negative that the model missed has the heaviest penalty ($b=0$). Then, given the vector of benefit values over the validation set, $\bm{b} = (b_1, b_2, \cdots, b_n)$, and their arithmetic mean $\mu(\bm{b})$, we measure the generalised entropy index fairness score $\mathcal{E}^2_{\bm{b}}$, where:
\begin{equation*}
    \mathcal{E}^{\alpha}(\bm{b}) = \frac{1}{n \alpha (\alpha-1)} \sum_{i=1}^n \left(\left( \frac{b_i}{\mu(\bm{b})} \right)^{\alpha}-1 \right)
\end{equation*}
Furthermore, given protected groups $g \in G$, with each comprising $n_g$ patient records with benefit vectors $\bm{b}^g = (b^g_1, b^g_2, \cdots, b^g_{n_g})$, we decompose the generalised entropy into its between-group component $\mathcal{E}^2_{\beta}$ and its within-group component $\mathcal{E}^2_{\omega}$, representing group and individual fairness, respectively. We measure the between-group component $\mathcal{E}^2_{\beta}$, where:

\begin{equation*}
    \mathcal{E}^{\alpha}_{\beta}(\bm{b}) = \mathcal{E}^{\alpha}(\bm{b}) - \mathcal{E}^{\alpha}_{\omega}(\bm{b}) = \sum^{\abs{G}}_{g=1} \frac{n_g}{n \alpha (\alpha-1)} \left(\left( \frac{\mu(\bm{b}_g)}{\mu(\bm{b})} \right)^{\alpha}-1 \right)
\end{equation*}
We define demographic groups based on the available protected characteristics - age and biological sex.  We partition the continuous age variable into age groups, as illustrated in Supplementary Figure 7. For both scores, the ideal value is $0$ and higher values indicate unfair classification.

We additionally compute the differential fairness bias amplification exhibited by our models \cite{Foulds20}. The differential fairness metric is defined from the standpoint of intersectionality, i.e., equally protecting population sub-groups defined by multiple overlapping protected characteristics. Bias amplification measures a predictive model's unfairness compared to any pre-existing bias reflected in the dataset due to inequality in the real-life generative process of the data. Given a set of patient records $\bm{x}$ and protected groups $(g_i, g_j) \in G \times G$, the (smoothed) differential fairness $\varepsilon$ of a classifier $C$ is defined by the relation:
\begin{equation*}
    e^{-\varepsilon} \leq \frac{\sum_{\bm{x} \in g_i} C(\bm{x}) + \alpha}{\abs{g_i} + \abs{R_Y}\alpha} \frac{\abs{g_j} + \abs{R_Y}\alpha}{\sum_{\bm{x} \in g_j} C(\bm{x}) + \alpha} \leq e^{\varepsilon}
\end{equation*}
where $\abs{R_Y} \alpha$ is the Dirichlet smoothing concentration parameter (we set $\alpha = 1.0$, assuming no prior information). Then, the bias amplification metric is defined as the difference $\varepsilon_{C} - \varepsilon_{D}$ of the differential fairness value for the model $C$ minus the value for the dataset $D$'s ground truth.  A negative bias amplification indicates that the predictive model reduces differential unfairness, while a positive value means the estimator is more biased than the original data.

\section*{Data Availability}
The data that facilitated the experiments of this study are provided by the Northern Care Alliance NHS Trust, but restrictions apply to the availability of this data, which were used under a data sharing agreement with Durham University for the current study, and so are not publicly available.

\section*{Acknowledgements}
The authors would like to thank Durham University, Biophysical Sciences Institute for supporting this work via the Summer Research Bursary.

\section*{Author contributions statement}
All authors carried out method and experimentation design. D.G. collected the data. S.B.L. carried out the experiments, summarised the results and prepared figures. All authors analysed and interpreted the results. S.B.L. wrote the manuscript text and all authors reviewed the manuscript.

\bibliography{references}

\end{document}

\newpage
\section*{Supplement}
\subsection*{Data Preparation}
\label{appendix:preprocessing}
We filter all manually recorded data features using the ranges in Table \ref{tab:supp_news_valid_ranges}.
Where a value is missing or falls outside these ranges, we replace it with a value inferred from the recorded NEWS2 sub-score. Specifically, we set it to be the midpoint of the relevant NEWS2 range. Where the sub-score has not been recorded but the raw value is present (and not invalid), we use the value to compute the NEWS2 sub-score ourselves.

\begin{table}[htbp]
    \caption{\label{tab:supp_news_valid_ranges} Valid ranges for manually recorded data features.}
    \renewcommand{\arraystretch}{1.3}
    \centering
    \begin{tabular}{lll}
        \toprule
        Variable         & Range    & Unit        \\
        \midrule
        SpO2             & $40-100$ & $\%$        \\
        Systolic BP      & $40-300$ & mmHg        \\
        Diastolic BP     & $20-200$ & mmHg        \\
        Temperature      & $25-45$  & $^{\circ}$C \\
        Pulse            & $25-300$ & Beats/min   \\
        Respiration Rate & $5-80$   & Breaths/min \\
        \bottomrule
    \end{tabular}
\end{table}

Furthermore, we apply the following feature-specific filtering:
\begin{itemize}
    \item \textbf{O2 Saturation (SpO2).} The NEWS2 specification gives two scales for this parameter \cite{RCP17}:
          \begin{itemize}
              \item $SpO2_1$: By default.
              \item $SpO2_2$: For patients with a prescribed oxygen saturation requirement of $88-92\%$ (e.g., in patients with hypercapnic respiratory failure).
          \end{itemize}
          Since the choice of scale is determined by the responsible clinical staff on a case-by-case basis, we use the following criteria to infer which scale to use when re-computing the NEWS2 sub-score for SpO2:
          \begin{itemize}
              \item Patients receiving oxygen using NIV, as recorded directly or in their set of coded procedures (OPCS-4 E85.2).
              \item Patients with COPD (as determined by ICD-10 coded diagnosis J44.*) AND receiving oxygen via Venturi 24 or 28.
              \item Patients with COPD (as above) and SpO2 $<88\%$.
          \end{itemize}
          Finally, if the patient is receiving supplemental oxygen, there is ambiguity as to whether a high NEWS2 oxygen sub-score indicates very high or very low saturation. In that case, we mark the value as missing.

    \item \textbf{Respiration Rate.} In addition to the range given in Table \ref{tab:supp_news_valid_ranges}, we assume triple-digit values to be erroneous entries of two-digit values (e.g., $250 \rightarrow 25.0$).

    \item \textbf{Oxygen Flow Rate.} This supplemental parameter is recorded in mixed units (Litres/min or FiO2). We translate all values to FiO2 where possible:
          \begin{itemize}
              \item Values $1-15$ are inferred to be in Litres/min.
              \item Decimal values are inferred to be FiO2, with the exception of $0.5$.
              \item Values of $0.5$ and any remaining values are determined based on the device used to deliver the oxygen. Nasal cannula and simple mask correspond to Litres/min, while other devices correspond to FiO2.
          \end{itemize}
          We convert Litres/min to FiO2 using the formula $FiO2 = 0.2 + (Litres/min * 4)/100$.
\end{itemize}

\begin{figure}[htbp]
    \centering
    \includesvg[inkscapelatex=false, width=\linewidth]{images/age_sex_bars.svg}
    \caption{Outcome distribution across age groups and biological sex. Each bar length represents the proportion of positive labels (identified deterioration outcomes) of the corresponding dataset that belonged to the indicated demographic group.}
    \label{fig:supp_age_sex}
\end{figure}

\newpage

\subsection*{Implementation}

\begin{table}[htbp]
    \caption{\label{tab:supp_hyperparameters} Hyperparameters for each model across the examined feature sets. These values were chosen using bayesian optimisation after 1000 iterations. Any omitted parameters were assigned their default values. }
    \centering
    \renewcommand{\arraystretch}{1.2}
    \begin{tabular}{llrrrrr}
        \toprule
        \textbf{Model}                      & \textbf{Parameter}       & Vitals                   & \& Obs                   & \& Labs                  & \& Notes                 & \& Services              \\
        \midrule
        \textbf{LR}                         & \verb|class_weight|   & \verb|None|   & \verb|None|   & \verb|None|   & \verb|None|   & \verb|None|   \\
        \cline{1-7}
        \multirow{3}{*}{\textbf{LR-EN}}     & \verb|C|   & \verb|0.0592|   & \verb|8.4126|   & \verb|1.1313|  & \verb|8.2228|  & \verb|7.0032|  \\
                                            & \verb|class_weight|  & \verb|None|  & \verb|None|  & \verb|None|  & \verb|None|  & \verb|None|  \\
                                            & \verb|l1_ratio|  & \verb|0.4133|  & \verb|0.355|  & \verb|0.9356|  & \verb|0.5862|  & \verb|0.8086|  \\
        \cline{1-7}
        \multirow{2}{*}{\textbf{LR-L1}}     & \verb|C|  & \verb|4.1169|  & \verb|2.8061|  & \verb|7.6951|  & \verb|4.4909|  & \verb|0.2121|  \\
                                            & \verb|class_weight|  & \verb|None|  & \verb|None|  & \verb|balanced|  & \verb|balanced|  & \verb|None|  \\
        \cline{1-7}
        \multirow{2}{*}{\textbf{LR-L2}}     & \verb|C|  & \verb|5.7762|  & \verb|8.6765|  & \verb|9.9825|  & \verb|9.9151|  & \verb|3.5386|  \\
                                            & \verb|class_weight|  & \verb|None|  & \verb|None|  & \verb|None|  & \verb|None|  & \verb|None|  \\
        \cline{1-7}
        \multirow{9}{*}{\textbf{LightGBM}}  & \verb|colsample_bytree|  & \verb|0.6075|  & \verb|0.989|  & \verb|0.5526|  & \verb|0.4737|  & \verb|0.4804|  \\
                                            & \verb|is_unbalance|  & \verb|True|  & \verb|False|  & \verb|False|  & \verb|False|  & \verb|False|  \\
                                            & \verb|min_child_samples|  & \verb|120|  & \verb|147|  & \verb|73|  & \verb|23|  & \verb|147|  \\
                                            & \verb|num_leaves|  & \verb|29|  & \verb|63|  & \verb|16|  & \verb|202|  & \verb|13|  \\
                                            & \verb|reg_alpha|  & \verb|2.6785|  & \verb|0.0402|  & \verb|1.5138|  & \verb|8.8073|  & \verb|0.0001|  \\
                                            & \verb|reg_lambda|  & \verb|9.9052|  & \verb|0.0009|  & \verb|0.0941|  & \verb|0.0024|  & \verb|9.5277|  \\
                                            & \verb|scale_pos_weight|  & \verb|None|  & \verb|53|  & \verb|2|  & \verb|4|  & \verb|2|  \\
                                            & \verb|subsample|  & \verb|0.8714|  & \verb|0.958|  & \verb|0.5702|  & \verb|0.9042|  & \verb|0.4902|  \\
                                            & \verb|subsample_freq|  & \verb|4|  & \verb|4|  & \verb|1| & \verb|2| & \verb|1| \\
        \cline{1-7}
        \multirow{2}{*}{\textbf{LinearSVM}} & \verb|alpha| & \verb|0.0032| & \verb|0.0004| & \verb|0.0002| & \verb|0.0011| & \verb|0.0005| \\
                                            & \verb|class_weight| & \verb|balanced| & \verb|None| & \verb|None| & \verb|None| & \verb|None| \\
        \cline{1-7}
        \multirow{11}{*}{\textbf{XGBoost}}  & \verb|alpha| & \verb|0.0| & \verb|0.0| & \verb|0.0| & \verb|0.0| & \verb|0.0| \\
                                            & \verb|colsample_bytree| & \verb|0.8389| & \verb|0.6157| & \verb|0.6273| & \verb|0.4414| & \verb|0.597| \\
                                            & \verb|eta| & \verb|0.0196| & \verb|0.0454| & \verb|0.0231| & \verb|0.0| & \verb|0.2193| \\
                                            & \verb|gamma| & \verb|0.0| & \verb|0.0| & \verb|0.0005| & \verb|0.0| & \verb|0.0| \\
                                            & \verb|grow_policy| & \verb|lossguide| & \verb|lossguide| & \verb|lossguide| & \verb|depthwise| & \verb|lossguide| \\
                                            & \verb|lambda| & \verb|0.0372| & \verb|0.0019| & \verb|0.0624| & \verb|0.157| & \verb|0.3922| \\
                                            & \verb|max_depth| & \verb|9| & \verb|7| & \verb|9| & \verb|9| & \verb|3| \\
                                            & \verb|min_child_weight| & \verb|8| & \verb|9| & \verb|6| & \verb|5| & \verb|4| \\
                                            & \verb|scale_pos_weight| & \verb|3| & \verb|6| & \verb|6| & \verb|15| & \verb|81| \\
                                            & \verb|subsample| & \verb|0.2505| & \verb|0.7236| & \verb|0.4115| & \verb|0.3563| & \verb|0.9937| \\
                                            & \verb|tree_method| & \verb|approx| & \verb|hist| & \verb|approx| & \verb|approx| & \verb|approx| \\
        \bottomrule
    \end{tabular}
\end{table}

\newpage
\subsection*{Results}
\begin{figure}[htbp]
    \centering
    \includesvg[inkscapelatex=false, width=\linewidth]{images/calib_curves.svg}
    \caption{Calibration Curves for \textbf{(a)}: GBDT (LightGBM) across feature groups (concatenated incrementally) and \textbf{(b)}: All learning algorithms trained the complete feature set (equiv. "\& Services"). Curves are plotted from each model's prediction outputs for the complete testing set.}
    \label{fig:supp_calib_curves}
\end{figure}

\begin{table}[htbp]
    \centering
    \caption{\label{tab:supp_data_categorical_summary} Summary of categorical features in the study sample. Each table section (except for ED Diagnosis) presents the 5 most prevalent values for each feature, with the remaining values represented by "Other".}
    \renewcommand{\arraystretch}{1.2}

    \begin{tabular}{llrrrr}
        \toprule
        \textbf{Variable}                              & \textbf{Value}               & Total            & Training        & Test (Complete) & Test (Unseen)   \\
        \midrule
        \multirow{6}{*}{\textbf{Admission Pathway}}    & \textbf{Booked Adm.}         & 1807 (1.49\%)    & 899 (1.11\%)    & 908 (2.27\%)    & 582 (2.52\%)    \\
                                                       & \textbf{Emg. ED}             & 110627 (91.38\%) & 74530 (91.89\%) & 36097 (90.36\%) & 20461 (88.67\%) \\
                                                       & \textbf{Emg. GP Ref.}        & 5866 (4.85\%)    & 3733 (4.6\%)    & 2133 (5.34\%)   & 1496 (6.48\%)   \\
                                                       & \textbf{Emg. OPD}            & 898 (0.74\%)     & 631 (0.78\%)    & 267 (0.67\%)    & 149 (0.65\%)    \\
                                                       & \textbf{Non-Emg. Transf.}    & 1180 (0.97\%)    & 792 (0.98\%)    & 388 (0.97\%)    & 279 (1.21\%)    \\
                                                       & \textbf{Other}               & 680 (0.56\%)     & 523 (0.64\%)    & 157 (0.39\%)    & 108 (0.47\%)    \\
        \cline{1-6}
        \multirow{6}{*}{\textbf{Admission Specialty}}  & \textbf{ED}                  & 18319 (15.13\%)  & 12891 (15.89\%) & 5428 (13.59\%)  & 3761 (16.3\%)   \\
                                                       & \textbf{Acute Int. Med.}     & 91025 (75.19\%)  & 59212 (73.0\%)  & 31813 (79.63\%) & 17568 (76.13\%) \\
                                                       & \textbf{Gen. Surgery}        & 2612 (2.16\%)    & 2395 (2.95\%)   & 217 (0.54\%)    & 146 (0.63\%)    \\
                                                       & \textbf{Nephrology}          & 3454 (2.85\%)    & 2281 (2.81\%)   & 1173 (2.94\%)   & 709 (3.07\%)    \\
                                                       & \textbf{Other}               & 4415 (3.65\%)    & 3334 (4.11\%)   & 1081 (2.71\%)   & 745 (3.23\%)    \\
                                                       & \textbf{Trauma \& Ortho.}    & 1233 (1.02\%)    & 995 (1.23\%)    & 238 (0.6\%)     & 146 (0.63\%)    \\
        \cline{1-6}
        \multirow{6}{*}{\textbf{Breathing Device}}     & \textbf{A - Air}             & 108851 (89.92\%) & 73233 (90.29\%) & 35618 (89.16\%) & 20830 (90.27\%) \\
                                                       & \textbf{N - Nasal cannula}   & 8900 (7.35\%)    & 5547 (6.84\%)   & 3353 (8.39\%)   & 1654 (7.17\%)   \\
                                                       & \textbf{Other}               & 1797 (1.48\%)    & 1410 (1.74\%)   & 387 (0.97\%)    & 221 (0.96\%)    \\
                                                       & \textbf{RM - Reservoir mask} & 394 (0.33\%)     & 221 (0.27\%)    & 173 (0.43\%)    & 107 (0.46\%)    \\
                                                       & \textbf{SM - Simple mask}    & 685 (0.57\%)     & 308 (0.38\%)    & 377 (0.94\%)    & 249 (1.08\%)    \\
                                                       & \textbf{V28 - Venturi 28\%}  & 391 (0.32\%)     & 349 (0.43\%)    & 42 (0.11\%)     & 14 (0.06\%)     \\
        \cline{1-6}
        \multirow{6}{*}{\textbf{Presenting Complaint}} & \textbf{Other}               & 36819 (30.41\%)  & 26077 (32.15\%) & 10742 (26.89\%) & 6731 (29.17\%)  \\
                                                       & \textbf{abdominal pain}      & 6413 (5.3\%)     & 4514 (5.57\%)   & 1899 (4.75\%)   & 1033 (4.48\%)   \\
                                                       & \textbf{chest pain}          & 11585 (9.57\%)   & 7633 (9.41\%)   & 3952 (9.89\%)   & 2371 (10.28\%)  \\
                                                       & \textbf{falls}               & 5278 (4.36\%)    & 3421 (4.22\%)   & 1857 (4.65\%)   & 950 (4.12\%)    \\
                                                       & \textbf{other}               & 31485 (26.01\%)  & 21565 (26.59\%) & 9920 (24.83\%)  & 5432 (23.54\%)  \\
                                                       & \textbf{shortness of breath} & 14865 (12.28\%)  & 9337 (11.51\%)  & 5528 (13.84\%)  & 2659 (11.52\%)  \\
        \cline{1-6}
        \multirow{13}{*}{\textbf{ED Diagnosis}}        & \textbf{collaps-}            & 1245 (1.03\%)    & 1245 (1.53\%)   & 0 (0\%)         & 0 (0\%)         \\
                                                       & \textbf{confus}              & 356 (0.29\%)     & 356 (0.44\%)    & 0 (0\%)         & 0 (0\%)         \\
                                                       & \textbf{cope}                & 592 (0.49\%)     & 348 (0.43\%)    & 244 (0.61\%)    & 150 (0.65\%)    \\
                                                       & \textbf{dementia}            & 227 (0.19\%)     & 85 (0.1\%)      & 142 (0.36\%)    & 80 (0.35\%)     \\
                                                       & \textbf{diarrh}              & 326 (0.27\%)     & 325 (0.4\%)     & 1 (0.0\%)       & 0 (0\%)         \\
                                                       & \textbf{dizz-}               & 89 (0.07\%)      & 89 (0.11\%)     & 0 (0\%)         & 0 (0\%)         \\
                                                       & \textbf{fall}                & 1461 (1.21\%)    & 1461 (1.8\%)    & 0 (0\%)         & 0 (0\%)         \\
                                                       & \textbf{head}                & 2704 (2.23\%)    & 1992 (2.46\%)   & 712 (1.78\%)    & 434 (1.88\%)    \\
                                                       & \textbf{mobility}            & 102 (0.08\%)     & 102 (0.13\%)    & 0 (0\%)         & 0 (0\%)         \\
                                                       & \textbf{pain}                & 7162 (5.92\%)    & 7139 (8.8\%)    & 23 (0.06\%)     & 9 (0.04\%)      \\
                                                       & \textbf{sudden}              & 1 (0.0\%)        & 1 (0.0\%)       & 0 (0\%)         & 0 (0\%)         \\
                                                       & \textbf{tight}               & 5 (0.0\%)        & 5 (0.01\%)      & 0 (0\%)         & 0 (0\%)         \\
                                                       & \textbf{vomit}               & 296 (0.24\%)     & 264 (0.33\%)    & 32 (0.08\%)     & 13 (0.06\%)     \\
        \bottomrule
    \end{tabular}
\end{table}

\newpage
\begin{table}[htbp]
    \centering
    \caption{\label{tab:supp_auc} AUROC (95\% bootstrapped confidence interval) of each classifier type trained on each feature set. 'Test (Complete)' indicates the full testing dataset, 'Test (Unseen)' the testing dataset excluding all patients that had admission records in the training set, and 'Delta' gives the difference (95\% bootstrapped confidence interval) in measured performance between the two datasets. }
    \begin{tabular}{llrrrrrr}

        \toprule
        \textbf{Features}                     & \textbf{Estimator} & Test (Complete)           & Test (Unseen)             & Delta                          \\

        \midrule
        \textbf{Reference}                    & \textbf{NEWS2}     & 0.79837 (0.78178-0.81774) & 0.81211 (0.78853-0.83516) & -0.01374 (-0.02892 - -0.00001) \\
        \cline{1-5}
        \multirow{7}{*}{\textbf{Vitals}}      & \textbf{LR}        & 0.81202 (0.79589-0.82812) & 0.82302 (0.80074-0.84682) & -0.01375 (-0.02613-0.00152)    \\
                                              & \textbf{LR-EN}     & 0.81084 (0.79487-0.82711) & 0.82240 (0.79999-0.84633) & -0.01433 (-0.02630-0.00110)    \\
                                              & \textbf{LR-L1}     & 0.81202 (0.79587-0.82817) & 0.82300 (0.80067-0.84684) & -0.01371 (-0.02613-0.00152)    \\
                                              & \textbf{LR-L2}     & 0.81204 (0.79593-0.82814) & 0.82302 (0.80070-0.84683) & -0.01373 (-0.02611-0.00153)    \\
                                              & \textbf{LightGBM}  & 0.83451 (0.81795-0.84949) & 0.84203 (0.82264-0.86360) & -0.01110 (-0.02066-0.00401)    \\
                                              & \textbf{LinearSVM} & 0.80991 (0.79308-0.82680) & 0.81609 (0.79223-0.84143) & -0.01150 (-0.02470-0.00489)    \\
                                              & \textbf{XGBoost}   & 0.83810 (0.82243-0.85295) & 0.84792 (0.83019-0.86909) & -0.01432 (-0.02323 - -0.00065) \\
        \cline{1-5}
        \multirow{7}{*}{\textbf{\& Obs}}      & \textbf{LR}        & 0.83622 (0.82009-0.85170) & 0.84324 (0.82577-0.86259) & -0.01140 (-0.02059-0.00670)    \\
                                              & \textbf{LR-EN}     & 0.83645 (0.82043-0.85198) & 0.84346 (0.82612-0.86266) & -0.01140 (-0.02063-0.00673)    \\
                                              & \textbf{LR-L1}     & 0.83633 (0.82028-0.85191) & 0.84337 (0.82579-0.86271) & -0.01144 (-0.02058-0.00676)    \\
                                              & \textbf{LR-L2}     & 0.83621 (0.82002-0.85168) & 0.84324 (0.82580-0.86251) & -0.01139 (-0.02065-0.00666)    \\
                                              & \textbf{LightGBM}  & 0.86770 (0.85390-0.88193) & 0.87881 (0.86276-0.89762) & -0.01402 (-0.02116-0.00076)    \\
                                              & \textbf{LinearSVM} & 0.81734 (0.80009-0.83362) & 0.80908 (0.78420-0.83397) & -0.01264 (-0.02414-0.00700)    \\
                                              & \textbf{XGBoost}   & 0.86855 (0.85494-0.88231) & 0.87920 (0.86105-0.89896) & -0.01472 (-0.02253 - -0.00123) \\
        \cline{1-5}
        \multirow{7}{*}{\textbf{\& Labs}}     & \textbf{LR}        & 0.83829 (0.82257-0.85332) & 0.84489 (0.82329-0.86387) & -0.01094 (-0.02094-0.00506)    \\
                                              & \textbf{LR-EN}     & 0.83848 (0.82249-0.85357) & 0.84495 (0.82355-0.86405) & -0.01076 (-0.02057-0.00537)    \\
                                              & \textbf{LR-L1}     & 0.84601 (0.83053-0.86145) & 0.85271 (0.83039-0.87151) & -0.01042 (-0.01997-0.00472)    \\
                                              & \textbf{LR-L2}     & 0.83826 (0.82263-0.85332) & 0.84484 (0.82330-0.86376) & -0.01095 (-0.02098-0.00508)    \\
                                              & \textbf{LightGBM}  & 0.88789 (0.87566-0.90086) & 0.89579 (0.87922-0.91310) & -0.01215 (-0.01954-0.00080)    \\
                                              & \textbf{LinearSVM} & 0.81036 (0.79234-0.82745) & 0.80485 (0.78289-0.82659) & -0.01146 (-0.02057-0.00578)    \\
                                              & \textbf{XGBoost}   & 0.88099 (0.86808-0.89400) & 0.88979 (0.87112-0.90633) & -0.01266 (-0.02039-0.00064)    \\
        \cline{1-5}
        \multirow{7}{*}{\textbf{\& Notes}}    & \textbf{LR}        & 0.84874 (0.83381-0.86303) & 0.85299 (0.83578-0.87141) & -0.00879 (-0.01549-0.00595)    \\
                                              & \textbf{LR-EN}     & 0.85112 (0.83649-0.86548) & 0.85363 (0.83651-0.87210) & -0.00705 (-0.01354-0.00785)    \\
                                              & \textbf{LR-L1}     & 0.85621 (0.84200-0.87038) & 0.85986 (0.84446-0.87751) & -0.00734 (-0.01414-0.00693)    \\
                                              & \textbf{LR-L2}     & 0.85086 (0.83647-0.86516) & 0.85360 (0.83637-0.87172) & -0.00726 (-0.01399-0.00744)    \\
                                              & \textbf{LightGBM}  & 0.89065 (0.87932-0.90209) & 0.89731 (0.88408-0.91332) & -0.01079 (-0.01708-0.00224)    \\
                                              & \textbf{LinearSVM} & 0.81852 (0.80439-0.83384) & 0.83075 (0.81248-0.85447) & -0.01783 (-0.02225 - -0.00194) \\
                                              & \textbf{XGBoost}   & 0.88454 (0.87280-0.89649) & 0.89101 (0.87656-0.90610) & -0.01075 (-0.01679-0.00086)    \\
        \cline{1-5}
        \multirow{7}{*}{\textbf{\& Services}} & \textbf{LR}        & 0.89144 (0.87930-0.90370) & 0.89280 (0.87805-0.90846) & -0.00693 (-0.01128-0.00637)    \\
                                              & \textbf{LR-EN}     & 0.89407 (0.88197-0.90623) & 0.89555 (0.88032-0.91168) & -0.00696 (-0.01114-0.00591)    \\
                                              & \textbf{LR-L1}     & 0.89551 (0.88354-0.90677) & 0.89778 (0.88336-0.91315) & -0.00729 (-0.01167-0.00505)    \\
                                              & \textbf{LR-L2}     & 0.89436 (0.88216-0.90637) & 0.89591 (0.88078-0.91169) & -0.00701 (-0.01120-0.00562)    \\
                                              & \textbf{LightGBM}  & 0.92208 (0.91150-0.93140) & 0.92295 (0.91214-0.93775) & -0.00648 (-0.01088-0.00560)    \\
                                              & \textbf{LinearSVM} & 0.86435 (0.84973-0.87873) & 0.85779 (0.84106-0.87776) & -0.00368 (-0.00907-0.01095)    \\
                                              & \textbf{XGBoost}   & 0.91272 (0.90224-0.92283) & 0.91374 (0.90198-0.92932) & -0.00616 (-0.01027-0.00639)    \\
        \bottomrule
    \end{tabular}
\end{table}

\begin{table}[htbp]
    \centering
    \caption{\label{tab:supp_ap}Average Precision (95\% bootstrapped confidence interval) of each classifier type trained on each feature set. 'Test (Complete)' indicates the full testing dataset, 'Test (Unseen)' the testing dataset excluding all patients that had admission records in the training set, and 'Delta' gives the difference (95\% bootstrapped confidence interval) in measured performance between the two datasets. }
    \begin{tabular}{llrrrrrr}

        \toprule
        \textbf{Features}                     & \textbf{Estimator} & Test (Complete)           & Test (Unseen)             & Delta                          \\

        \midrule
        \textbf{Reference}                    & \textbf{NEWS2}     & 0.14620 (0.12840-0.16843) & 0.16527 (0.13716-0.19425) & -0.01906 (-0.03780-0.00128)    \\
        \cline{1-5}
        \multirow{7}{*}{\textbf{Vitals}}      & \textbf{LR}        & 0.15654 (0.13472-0.18050) & 0.17224 (0.14391-0.20818) & -0.01688 (-0.03439-0.00583)    \\
                                              & \textbf{LR-EN}     & 0.15472 (0.13326-0.17813) & 0.16963 (0.14095-0.20585) & -0.01602 (-0.03203-0.00650)    \\
                                              & \textbf{LR-L1}     & 0.15637 (0.13463-0.18024) & 0.17201 (0.14379-0.20773) & -0.01683 (-0.03424-0.00581)    \\
                                              & \textbf{LR-L2}     & 0.15654 (0.13472-0.18066) & 0.17208 (0.14379-0.20789) & -0.01675 (-0.03419-0.00589)    \\
                                              & \textbf{LightGBM}  & 0.17902 (0.15677-0.20542) & 0.20573 (0.17860-0.24097) & -0.02778 (-0.04823 - -0.00317) \\
                                              & \textbf{LinearSVM} & 0.15484 (0.13396-0.17888) & 0.17861 (0.14837-0.21558) & -0.01999 (-0.03769-0.00482)    \\
                                              & \textbf{XGBoost}   & 0.17117 (0.15002-0.19579) & 0.19383 (0.16532-0.22807) & -0.02574 (-0.04103 - -0.00242) \\
        \cline{1-5}
        \multirow{7}{*}{\textbf{\& Obs}}      & \textbf{LR}        & 0.25813 (0.22819-0.28952) & 0.30903 (0.27133-0.37279) & -0.05699 (-0.07260 - -0.02352) \\
                                              & \textbf{LR-EN}     & 0.25743 (0.22782-0.28908) & 0.30870 (0.27093-0.37123) & -0.05701 (-0.07294 - -0.02303) \\
                                              & \textbf{LR-L1}     & 0.25673 (0.22718-0.28836) & 0.30804 (0.26987-0.37190) & -0.05705 (-0.07247 - -0.02289) \\
                                              & \textbf{LR-L2}     & 0.25767 (0.22815-0.28926) & 0.30914 (0.27135-0.37176) & -0.05721 (-0.07297 - -0.02316) \\
                                              & \textbf{LightGBM}  & 0.31362 (0.28485-0.34723) & 0.37001 (0.32539-0.42758) & -0.06763 (-0.08137 - -0.02843) \\
                                              & \textbf{LinearSVM} & 0.25792 (0.22883-0.28772) & 0.28442 (0.24230-0.35246) & -0.06619 (-0.08441 - -0.03117) \\
                                              & \textbf{XGBoost}   & 0.30981 (0.28000-0.34358) & 0.36394 (0.32159-0.41987) & -0.06574 (-0.07730 - -0.02969) \\
        \cline{1-5}
        \multirow{7}{*}{\textbf{\& Labs}}     & \textbf{LR}        & 0.25956 (0.22983-0.29112) & 0.30524 (0.26482-0.37422) & -0.05144 (-0.06715 - -0.02000) \\
                                              & \textbf{LR-EN}     & 0.25752 (0.22833-0.28941) & 0.30210 (0.26191-0.37002) & -0.05019 (-0.06629 - -0.01838) \\
                                              & \textbf{LR-L1}     & 0.24392 (0.21573-0.27720) & 0.28401 (0.24618-0.35152) & -0.04440 (-0.06184 - -0.00980) \\
                                              & \textbf{LR-L2}     & 0.25893 (0.22947-0.29100) & 0.30484 (0.26505-0.37455) & -0.05161 (-0.06791 - -0.01961) \\
                                              & \textbf{LightGBM}  & 0.31363 (0.28367-0.34813) & 0.36479 (0.31386-0.42422) & -0.06396 (-0.06988 - -0.01894) \\
                                              & \textbf{LinearSVM} & 0.25743 (0.22934-0.28827) & 0.25249 (0.20524-0.32339) & -0.05987 (-0.08037 - -0.02942) \\
                                              & \textbf{XGBoost}   & 0.31982 (0.28903-0.35317) & 0.37818 (0.33927-0.43446) & -0.06876 (-0.07788 - -0.03054) \\
        \cline{1-5}
        \multirow{7}{*}{\textbf{\& Notes}}    & \textbf{LR}        & 0.26304 (0.23415-0.29418) & 0.31080 (0.26920-0.37828) & -0.05432 (-0.06790 - -0.02268) \\
                                              & \textbf{LR-EN}     & 0.26395 (0.23514-0.29546) & 0.31155 (0.27015-0.37778) & -0.05416 (-0.06802 - -0.02305) \\
                                              & \textbf{LR-L1}     & 0.24261 (0.21452-0.27399) & 0.28588 (0.24262-0.35254) & -0.04774 (-0.06149 - -0.01234) \\
                                              & \textbf{LR-L2}     & 0.26380 (0.23494-0.29479) & 0.31169 (0.27030-0.37863) & -0.05444 (-0.06805 - -0.02321) \\
                                              & \textbf{LightGBM}  & 0.32338 (0.29322-0.35592) & 0.37353 (0.32555-0.42557) & -0.06296 (-0.07043 - -0.02216) \\
                                              & \textbf{LinearSVM} & 0.25459 (0.22619-0.28545) & 0.25961 (0.21181-0.32756) & -0.06210 (-0.07710 - -0.02972) \\
                                              & \textbf{XGBoost}   & 0.29689 (0.26741-0.32912) & 0.34996 (0.31093-0.41418) & -0.06141 (-0.07232 - -0.02160) \\
        \cline{1-5}
        \multirow{7}{*}{\textbf{\& Services}} & \textbf{LR}        & 0.46334 (0.42731-0.49760) & 0.47583 (0.41855-0.52883) & -0.03587 (-0.04615-0.00279)    \\
                                              & \textbf{LR-EN}     & 0.46615 (0.43139-0.50070) & 0.47778 (0.41715-0.52816) & -0.03458 (-0.04480-0.00443)    \\
                                              & \textbf{LR-L1}     & 0.45835 (0.42212-0.49458) & 0.46817 (0.40793-0.51891) & -0.03170 (-0.04219-0.00539)    \\
                                              & \textbf{LR-L2}     & 0.46588 (0.43123-0.50015) & 0.47774 (0.41752-0.52838) & -0.03447 (-0.04454-0.00380)    \\
                                              & \textbf{LightGBM}  & 0.51268 (0.47813-0.54589) & 0.52347 (0.46740-0.57071) & -0.03662 (-0.05118-0.00287)    \\
                                              & \textbf{LinearSVM} & 0.44708 (0.41128-0.48127) & 0.43666 (0.39277-0.48465) & -0.03908 (-0.04888 - -0.00210) \\
                                              & \textbf{XGBoost}   & 0.49718 (0.46321-0.53057) & 0.50369 (0.45329-0.55213) & -0.03409 (-0.04570-0.00897)    \\
        \bottomrule
    \end{tabular}
\end{table}

\begin{figure}[htbp]
    \centering
    \includesvg[inkscapelatex=false, width=\textwidth,height=0.85\textheight,keepaspectratio]{images/timeseries_auc.svg}
    \caption{AUROC of logistic regression (L2 penalty) and GBDT (LightGBM) across thresholds for labelling critical deterioration events. Each section \textbf{(a-e)} corresponds to incrementally including the indicated feature sets (from Methods Table \ref{tab:feature_sets}) as training data. For each one, and for each x-axis value, we train independent models to identify critical deterioration up to the corresponding number of days after admission and measure their AUROC on the validation set (y-axis). The lines in colour represent the performance for the indicated feature set, and the lines in gray represent the lines from the other sections for easier visual comparison. }
    \label{fig:supp_auc_over_time}
\end{figure}

\newpage

\begin{figure}[htbp]
    \centering
    \includesvg[inkscapelatex=false, width=\linewidth]{images/shap_categorical.svg}
    \caption{SHAP feature importances for categorical feature values under the GBDT (LightGBM) model. Each bar shows the mean SHAP value for the indicated value of the corresponding categorical feature. Values to the right of $x=0$ (red) contribute, on average, towards predictions of deterioration, while values to the left (blue) contribute towards negative predictions (no deterioration). We provide the highest-ranked values per feature by mean absolute SHAP value. }
    \label{fig:supp_shap_lightgbm_categorical}
\end{figure}

\begin{table}[htbp]
    \centering
    \renewcommand{\arraystretch}{1.2}
    \caption{\label{tab:supp_lr_coefficients} Coefficients of numerical features in logistic regression models. LR: Logistic regression; LR-L1: LR with L1 penalty regularisation; LR-L2: LR with L2 penalty regularisation; LR-EN: LR with Elastic Net regularisation.}
    \begin{tabular}{lrrrr}
        \toprule
        {}                          & LR        & LR-L1     & LR-L2     & LR-EN     \\
        \midrule
        \textbf{Sent To SDEC      } & $-1.0367$ & $-1.0309$ & $-1.0391$ & $-1.0384$ \\
        \textbf{Sodium            } & $-0.8081$ & $-0.7383$ & $-0.8016$ & $-0.8073$ \\
        \textbf{Age               } & $-0.546$  & $-0.4507$ & $-0.5434$ & $-0.5433$ \\
        \textbf{FiO2              } & $-0.4453$ & $-0.4357$ & $-0.4423$ & $-0.4440$ \\
        \textbf{SpO2              } & $-0.1996$ & $-0.1972$ & $-0.1994$ & $-0.1993$ \\
        \textbf{Creatinine        } & $-0.1968$ & $-0.1889$ & $-0.1953$ & $-0.1950$ \\
        \textbf{Assisted Breathing} & $-0.1301$ & $-0.1242$ & $-0.1292$ & $-0.1288$ \\
        \textbf{30-day Readmission} & $-0.1131$ & $-0.0779$ & $-0.1144$ & $-0.1143$ \\
        \textbf{Temperature       } & $-0.0525$ & $-0.0449$ & $-0.0517$ & $-0.0520$ \\
        \textbf{Alert (AVCPU)     } & $-0.0456$ & $-0.0183$ & $-0.0408$ & $-0.0413$ \\
        \textbf{Systolic BP       } & $-0.0205$ & $-0.0216$ & $-0.0212$ & $-0.0215$ \\
        \textbf{Urea              } & $0.0075$  & $-0.0013$ & $0.0059$  & $0.0060$  \\
        \textbf{Pain              } & $0.0766$  & $0.0748$  & $0.0790$  & $0.0785$  \\
        \textbf{Nausea            } & $0.0796$  & $0.0407$  & $0.0798$  & $0.0791$  \\
        \textbf{Diastolic BP      } & $0.1157$  & $0.0803$  & $0.1114$  & $0.1130$  \\
        \textbf{Pulse             } & $0.1193$  & $0.1191$  & $0.1192$  & $0.1191$  \\
        \textbf{Haemoglobin       } & $0.1403$  & $0.0989$  & $0.1384$  & $0.1384$  \\
        \textbf{Vomiting          } & $0.1891$  & $0.1940$  & $0.1902$  & $0.1899$  \\
        \textbf{Respiration Rate  } & $0.1976$  & $0.1984$  & $0.1971$  & $0.1975$  \\
        \textbf{Female            } & $0.2662$  & $0.2556$  & $0.2664$  & $0.2670$  \\
        \textbf{Lying Down        } & $0.3488$  & $0.3320$  & $0.3484$  & $0.3488$  \\
        \textbf{Potassium         } & $0.9721$  & $0.4844$  & $1.4895$  & $0.6155$  \\
        \bottomrule
    \end{tabular}
\end{table}

\begin{table}[htbp]
    \centering
    \caption{\label{tab:supp_lr_categorical_coefficients} Coefficients of categorical feature values in logistic regression models. We provide the 5 highest ranked values per feature by mean absolute coefficient over the presented models. LR: Logistic regression; LR-L1: LR with L1 penalty regularisation; LR-L2: LR with L2 penalty regularisation; LR-EN: LR with Elastic Net regularisation.}
    \renewcommand{\arraystretch}{1.2}

    \begin{tabular}{llrrrr}
        \toprule
        \textbf{Feature}                               & \textbf{Value}                 & LR      & LR-EN   & LR-L1   & LR-L2   \\
        \midrule
        \multirow{5}{*}{\textbf{ED Diagnosis}}         & \textbf{dementia}              & -5.7765 & -1.1784 & 0.0000  & -0.9183 \\
                                                       & \textbf{dizz-}                 & -4.5919 & -0.5101 & 0.0000  & -0.5284 \\
                                                       & \textbf{Unknown}               & 1.1955  & 0.5185  & 0.4662  & 0.6791  \\
                                                       & \textbf{diarrh}                & 0.9823  & 0.2567  & 0.0000  & 0.4210  \\
                                                       & \textbf{collaps-}              & 0.8968  & 0.2167  & 0.0000  & 0.3797  \\
        \cline{1-6}
        \multirow{5}{*}{\textbf{Admission Pathway}}    & \textbf{Emg. Tran. Oth. Prov.} & 1.1180  & 1.5083  & 0.8563  & 1.6927  \\
                                                       & \textbf{Elective Booked}       & -2.8855 & -1.2208 & 0.0000  & -0.8683 \\
                                                       & \textbf{Emg. OPD}              & -1.7830 & -1.2437 & 0.0000  & -0.9072 \\
                                                       & \textbf{Booked Adm.}           & -1.5348 & -1.0707 & 0.0000  & -0.7858 \\
                                                       & \textbf{Emg. ED}               & 0.4351  & 0.8734  & 0.9080  & 1.1078  \\
        \cline{1-6}
        \multirow{5}{*}{\textbf{Admission Specialty}}  & \textbf{General Med.}          & 6.0295  & 4.4717  & 4.5527  & 4.2806  \\
                                                       & \textbf{Respiratory Med.}      & 5.7150  & 4.1316  & 3.7710  & 3.8547  \\
                                                       & \textbf{Geriatric Med.}        & 5.1367  & 3.5265  & 2.4572  & 3.1878  \\
                                                       & \textbf{Bariatric Surg.}       & 8.3252  & 2.2719  & 0.0000  & 1.7883  \\
                                                       & \textbf{Spinal Surg. Svc.}     & -6.1272 & -2.0428 & 0.0000  & -1.5927 \\
        \cline{1-6}
        \multirow{5}{*}{\textbf{Breathing Device}}     & \textbf{NHF}                   & 3.1081  & 3.1803  & 3.0982  & 3.0330  \\
                                                       & \textbf{Unknown}               & -5.8139 & -1.2859 & 0.0000  & -0.6555 \\
                                                       & \textbf{A - Air}               & -1.4371 & -1.6811 & -1.3501 & -0.8618 \\
                                                       & \textbf{NIV - NIV}             & 1.1222  & 1.2162  & 1.2235  & 1.1433  \\
                                                       & \textbf{Other}                 & 0.8460  & 0.9446  & 1.1644  & 0.8756  \\
        \cline{1-6}
        \multirow{5}{*}{\textbf{Presenting Complaint}} & \textbf{ear problems}          & -6.0519 & -1.5286 & 0.0000  & -1.2409 \\
                                                       & \textbf{facial problems}       & -5.5940 & -1.4631 & 0.0000  & -1.1675 \\
                                                       & \textbf{gi bleeding}           & 2.8015  & 1.7446  & 1.3005  & 1.7253  \\
                                                       & \textbf{diabetes}              & 2.7086  & 1.6517  & 1.1506  & 1.6313  \\
                                                       & \textbf{testicular pain}       & -4.8856 & -0.9064 & 0.0000  & -0.8156 \\
        \bottomrule
    \end{tabular}
\end{table}

\begin{table}[htbp]
    \centering
    \caption{\label{tab:supp_bias_amp} Differential Fairness Bias Amplification (95\% bootstrapped confidence interval) of each classifier type trained on each feature set. The columns "Sex", "Age", and "Sex \& Age" indicate the protected characteristic for each measurement - biological sex,  age group (per Figure \ref{fig:supp_age_sex}), or both.  }
    \begin{tabular}{llrrrrrr}

        \toprule
        \textbf{Features}                     & \textbf{Estimator} & Sex                      & Age                      & Sex \& Age               \\
        \midrule
        \multirow{7}{*}{\textbf{Vitals}}      & \textbf{LR}        & -0.240 (-0.374 - -0.115) & 0.107 (0.060 - 0.267)    & -0.213 (-0.343 - -0.010) \\
                                              & \textbf{LR-EN}     & -0.240 (-0.373 - -0.114) & 0.110 (0.065 - 0.269)    & -0.211 (-0.337 - -0.007) \\
                                              & \textbf{LR-L1}     & -0.240 (-0.374 - -0.115) & 0.106 (0.060 - 0.267)    & -0.214 (-0.344 - -0.011) \\
                                              & \textbf{LR-L2}     & -0.240 (-0.374 - -0.115) & 0.107 (0.060 - 0.267)    & -0.213 (-0.343 - -0.011) \\
                                              & \textbf{LightGBM}  & -0.245 (-0.396 - -0.123) & 0.030 (-0.019 - 0.194)   & -0.312 (-0.428 - -0.089) \\
                                              & \textbf{LinearSVM} & -0.253 (-0.385 - -0.125) & 0.200 (0.142 - 0.359)    & -0.106 (-0.229 - 0.077)  \\
                                              & \textbf{XGBoost}   & -0.242 (-0.392 - -0.125) & 0.089 (0.045 - 0.252)    & -0.250 (-0.348 - -0.036) \\
        \cline{1-5}
        \multirow{7}{*}{\textbf{\& Obs}}      & \textbf{LR}        & 0.050 (-0.114 - 0.114)   & -0.483 (-0.554 - -0.287) & -0.617 (-0.773 - -0.441) \\
                                              & \textbf{LR-EN}     & 0.050 (-0.115 - 0.113)   & -0.484 (-0.555 - -0.288) & -0.619 (-0.775 - -0.442) \\
                                              & \textbf{LR-L1}     & 0.049 (-0.116 - 0.112)   & -0.484 (-0.555 - -0.288) & -0.619 (-0.774 - -0.442) \\
                                              & \textbf{LR-L2}     & 0.051 (-0.115 - 0.114)   & -0.484 (-0.555 - -0.288) & -0.618 (-0.776 - -0.442) \\
                                              & \textbf{LightGBM}  & 0.006 (-0.140 - 0.081)   & -0.320 (-0.407 - -0.138) & -0.465 (-0.622 - -0.302) \\
                                              & \textbf{LinearSVM} & -0.147 (-0.271 - -0.055) & -0.687 (-0.750 - -0.554) & -1.016 (-1.111 - -0.826) \\
                                              & \textbf{XGBoost}   & -0.022 (-0.174 - 0.058)  & -0.298 (-0.349 - -0.101) & -0.494 (-0.636 - -0.298) \\
        \cline{1-5}
        \multirow{7}{*}{\textbf{\& Labs}}     & \textbf{LR}        & 0.074 (-0.098 - 0.143)   & -0.500 (-0.564 - -0.312) & -0.604 (-0.769 - -0.431) \\
                                              & \textbf{LR-EN}     & 0.074 (-0.099 - 0.141)   & -0.499 (-0.562 - -0.311) & -0.602 (-0.771 - -0.431) \\
                                              & \textbf{LR-L1}     & 0.095 (-0.079 - 0.168)   & -0.504 (-0.552 - -0.314) & -0.589 (-0.685 - -0.351) \\
                                              & \textbf{LR-L2}     & 0.074 (-0.098 - 0.143)   & -0.499 (-0.564 - -0.312) & -0.602 (-0.768 - -0.430) \\
                                              & \textbf{LightGBM}  & 0.034 (-0.123 - 0.111)   & -0.317 (-0.388 - -0.144) & -0.416 (-0.583 - -0.265) \\
                                              & \textbf{LinearSVM} & -0.047 (-0.172 - 0.040)  & -0.592 (-0.671 - -0.406) & -0.660 (-0.805 - -0.464) \\
                                              & \textbf{XGBoost}   & -0.015 (-0.165 - 0.067)  & -0.329 (-0.404 - -0.140) & -0.514 (-0.648 - -0.324) \\
        \cline{1-5}
        \multirow{7}{*}{\textbf{\& Notes}}    & \textbf{LR}        & 0.065 (-0.110 - 0.139)   & -0.475 (-0.539 - -0.285) & -0.577 (-0.760 - -0.404) \\
                                              & \textbf{LR-EN}     & 0.069 (-0.109 - 0.143)   & -0.476 (-0.543 - -0.286) & -0.573 (-0.761 - -0.402) \\
                                              & \textbf{LR-L1}     & 0.076 (-0.104 - 0.155)   & -0.499 (-0.556 - -0.310) & -0.596 (-0.766 - -0.420) \\
                                              & \textbf{LR-L2}     & 0.069 (-0.109 - 0.143)   & -0.476 (-0.542 - -0.286) & -0.573 (-0.761 - -0.401) \\
                                              & \textbf{LightGBM}  & 0.013 (-0.132 - 0.084)   & -0.313 (-0.371 - -0.131) & -0.453 (-0.606 - -0.292) \\
                                              & \textbf{LinearSVM} & -0.047 (-0.187 - 0.055)  & -0.657 (-0.733 - -0.514) & -0.885 (-0.996 - -0.694) \\
                                              & \textbf{XGBoost}   & -0.061 (-0.215 - 0.017)  & -0.187 (-0.242 - -0.032) & -0.472 (-0.568 - -0.245) \\
        \cline{1-5}
        \multirow{7}{*}{\textbf{\& Services}} & \textbf{LR}        & 0.041 (-0.088 - 0.114)   & -0.454 (-0.532 - -0.241) & -0.588 (-0.734 - -0.372) \\
                                              & \textbf{LR-EN}     & 0.047 (-0.085 - 0.120)   & -0.459 (-0.538 - -0.244) & -0.587 (-0.728 - -0.372) \\
                                              & \textbf{LR-L1}     & 0.047 (-0.086 - 0.117)   & -0.456 (-0.543 - -0.239) & -0.584 (-0.721 - -0.374) \\
                                              & \textbf{LR-L2}     & 0.050 (-0.083 - 0.123)   & -0.459 (-0.539 - -0.244) & -0.584 (-0.727 - -0.370) \\
                                              & \textbf{LightGBM}  & 0.004 (-0.118 - 0.086)   & -0.263 (-0.312 - -0.048) & -0.396 (-0.501 - -0.212) \\
                                              & \textbf{LinearSVM} & -0.137 (-0.229 - -0.043) & -0.579 (-0.647 - -0.399) & -0.878 (-0.959 - -0.627) \\
                                              & \textbf{XGBoost}   & -0.005 (-0.122 - 0.078)  & -0.260 (-0.322 - -0.049) & -0.420 (-0.527 - -0.219) \\
        \bottomrule
    \end{tabular}

\end{table}

\begin{figure}[htbp]
    \centering
    \includesvg[inkscapelatex=false, width=\textwidth,height=0.9\textheight,keepaspectratio]{images/entropy_between.svg}
    \caption{Between-Group Generalised Entropy vs Sensitivity curvesof \textbf{(a)}: GBDT across the tested feature sets, and \textbf{(b)}: All classifier types trained on the complete feature set. We plot the between-group component of the generalised entropy index, representing unfairness between demographic groups defined by the protected characteristics of age group and sex, per Figure \ref{fig:supp_age_sex}. The remainder of the generalised entropy, as presented in Figure \ref{fig:entropy} is the \emph{within-group} component, representing all other potential biases. A lower value on the y-axis indicates a more fair distribution of 'benefit', i.e. of receiving a positive prediction, between the demographic groups we consider. A theoretical 'perfect' model would yield a single point $(0,1)$ in the lower-right corner of the plot. }
    \label{fig:supp_entropy_between}
\end{figure}

### 2023-02-19 dataset.py

@@ -371,7 +371,7 @@ class SCIData(pd.DataFrame):
         if return_series:
             return col
 
-        r[col_name]
+        r[col_name] = col
 
         return SCIData(r)
 
